### 2024-11-10 paper/abstract/icid.Rmd
Pre poster

---
title: "Gastro-intestinal transmission and colonisation of pathogenic bacteria: calibration of an agent-based model"
author: "Alessandro Gerada, Conor Rosato, Alex Howard, Peter L Green, William Hope"
date: "2024-03-18"
output:
  word_document: default
  pdf_document: default
---

__Introduction__
Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection. Examples include _Clostridioides difficile_ infection (CDI) and Vancomycin-resistant _Enterococcus_. In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents. To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by Salgado et al [American Journal of Infection Control 37, 458–464 (2009)].

__Methods__
An agent-based model was implemented using NetLogo -- a virtual hospital was built with a configurable structure (number of wards, beds per ward, number of toilets, etc.). Input parameters included patient factors (e.g., community colonisation rate), population behaviour (e.g., frequency of toilet use), and infection control measures (e.g., toilet cleaning frequency). Risk of infection changed as patients shared use of contaminated toilets, or were prescribed antimicrobial therapy. Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data. Using cases per 1000 patient days as a summary statistic, simulated data were compared to the actual reported rate of nosocomial CDI. Priors were determined from domain expertise or extracted from existing literature. Firstly, the model was calibrated to pre-outbreak data to identify parameters that simulated the baseline rate. Next, we simulated outbreaks and suppression by incrementally raising colonisation pressure over time (via increased antimicrobial consumption, toilet use, etc.) and enhancing infection control measures, respectively. Finally, the estimated parameter means were used as an input in 56 outbreak random simulations.

__Results__
After calibration, key parameters included: community colonisation rate of 3.8%, rising to 5.7% during the outbreak period; 31.9% patients were administered an antimicrobial during their admission, rising to 45.8% during the outbreak, and reducing to 27.9% after enhancing infection control measures. The rate of patient toilet frequenting rose from a mean of 1.6 to 2.8 times during the outbreak. Simulated data had a pre-outbreak baseline median rate of 0.75 cases per 1000 patient days [interquartile (IQR) range 0--1.66], compared with the real rate of 1.25 cases per 1000 [IQR 0.94--1.56]. During the outbreak period, simulations had a median rate of 4.65 [IQR 2.34--7.65] cases, compared with the real median rate of 3.74 [IQR 3.32--4.41].

__Discussion__
Repeated simulations using the model generated nosocomial CDI rates that showed a clear outbreak peak, followed by suppression of colonisation/infection over time, resembling a real hospital CDI outbreak. The interactive nature of the model can be used for outbreak preparedness by testing out different infection control strategies.

__Conclusion__
We report a pathogen-agnostic model that can be used to simulate any infection with features such as faecal-oral transmission, colonisation, and antimicrobial selective pressure effects.

### 2024-11-10 calibrate.R

@@ -575,6 +575,12 @@ results_outbreak_control_summary <- results_outbreak_control %>% distinct(siminp
 
 params_names_outbreak_control <- names(nl@experiment@variables)
 
+# keep simulated parameters that fit with outbreak generation
+results_outbreak_control_summary <- results_outbreak_control_summary %>% 
+  dplyr::filter(`o-toilet-contamination-effect` >= `toilet-contamination-effect` &
+                  `o-toilet-frequenting-rate` >= `toilet-frequenting-rate` &
+                  `o-community-colonisation-rate` >= `community-colonisation-rate`)
+
 outbreak_control_sims <- results_outbreak_control_summary %>% 
   dplyr::select(all_of(params_names_outbreak_control), siminputrow, month, year, rate) %>% 
   mutate(date_sim = ymd(paste(year, month, "01", sep="-"))) %>%
@@ -582,13 +588,6 @@ outbreak_control_sims <- results_outbreak_control_summary %>%
   dplyr::select(!date_sim) %>% 
   pivot_wider(names_from = c(year, month), values_from = rate,
               names_prefix = "rate_")
-
-# try to make sure only parameters compatible with outbreak are kept
-# outbreak_control_sims <- outbreak_control_sims %>% 
-#   ungroup %>% 
-#   dplyr::filter(`o-toilet-contamination-effect` >= `toilet-contamination-effect` & 
-#                   `o-toilet-frequenting-rate` >= `toilet-frequenting-rate` &
-#                   `o-community-colonisation-rate` >= `community-colonisation-rate`)
 
 sumstats <- outbreak_control_sims %>% 
   ungroup %>% 
@@ -623,6 +622,17 @@ sim_mean <- long_retained_ss %>%
   summarise(m = mean(rates),
             sd = sd(rates))
 
+# sim_mean <- long_retained_ss %>% 
+#   group_by(x) %>% 
+#   summarise(m = mean(rates),
+#             sd = sd(rates))
+
+sim_mean <- long_retained_ss %>% 
+  group_by(x) %>% 
+  summarise(m = median(rates),
+            q_low = quantile(rates, 0.25),
+            q_high = quantile(rates, 0.75))
+
 abc_params_outbreak_control$unadj.values %>% 
   as_tibble %>% 
[…]

### 2024-11-11 calibrate.R
parameters fixed for day run

@@ -4,7 +4,7 @@ library(tidyverse)
 library(lubridate)
 library(abc)
 netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- "hospital.nlogo"
+model_path <- "salgado_v0.3_day.nlogo"
 out_path <- "out/"
 wards_total <- 16
 bedspaces_per_ward <- 14
@@ -12,6 +12,7 @@ bedspaces_per_ward <- 14
 ##============== pre-outbreak =============##
 
 sim_days <- (365 * 3) + 30
+sim_hours <- sim_days * 24
 
 nl <- nl(nlversion = "6.2.2",
          nlpath = netlogo_path,
@@ -24,7 +25,7 @@ nl@experiment <- experiment(expname = "baseline_rate",
                             tickmetrics = "true",
                             idsetup = "setup",
                             idgo = "go",
-                            runtime = sim_days,
+                            runtime = sim_hours,
                             metrics = c("total-colonised",
                                         "total-patients-admitted",
                                         "total-hospital-infections",
@@ -40,6 +41,8 @@ nl@experiment <- experiment(expname = "baseline_rate",
                                              "toilet-cleaning-rate" = list(min=1, max=3, qfun='qunif'),
                                              "random-colonisation-thresh" = list(min=0.01, max=0.99, qfun='qunif')),
                             constants = list("wards-total" = wards_total,
+                                             "outbreak?" = FALSE,
+                                             "infection-control?" = FALSE,
                                              "bedspaces-per-ward" = bedspaces_per_ward,
                                              "antibiotic-prescription-rate" = 0.319, # Vesporten 2018
                                              "admission-days" = 8.3,
@@ -140,7 +143,7 @@ ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
 
 sim_days <- (365 * 7) + 30
 
-outbreak_control_model <- "salgado.nlogo"
+outbreak_control_model <- "salgado_v0.3.nlogo"
 
 nl <- nl(nlversion = "6.2.2",
          nlpath = netlogo_path,

### 2024-11-29 calibrate.R
calibrate update

@@ -5,11 +5,16 @@ library(lubridate)
 library(abc)
 library(progressr)
 handlers("cli")
-netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- "salgado_v0.42_day_time.nlogo"
-out_path <- "out/"
 wards_total <- 16
 bedspaces_per_ward <- 14
+set.seed(42)
+
+netlogo_path <- "/Applications/NetLogo 6.2.2/"
+model_path <- "salgado_v0.42_day_time.nlogo"
+out_path <- "../out/"
+data_path <- "../data/"
+
+run_sims <- FALSE
 
 ##============== pre-outbreak =============##
 
@@ -59,9 +64,13 @@ nl@simdesign <- simdesign_lhs(nl,
 
 plan(list(sequential, multisession))
 
-results_baseline <- progressr::with_progress(
-  run_nl_all(nl))
-write_rds(results_baseline, file.path(out_path, "results_baseline.rds"))
+if (run_sims) {
+  results_baseline <- progressr::with_progress(
+    run_nl_all(nl))
+  write_rds(results_baseline, file.path(out_path, "results_baseline.rds"))
+}
+
+results_baseline <- read_rds(file.path(out_path, "results_baseline.rds"))
 
 results_baseline_bak <- results_baseline
 results_baseline <- results_baseline %>% 
@@ -84,7 +93,7 @@ results_baseline_summary <- results_baseline %>% distinct(siminputrow, .keep_all
 
 params_names_pre_outbreak <- names(nl@experiment@variables)
 
-salgado <- read_csv("data/salgado_et_al.csv") %>%
+salgado <- read_csv(file.path(data_path, "salgado_et_al.csv")) %>%
   mutate(x = seq(ymd('2002-01-01'), by = '1 month', length.out=nrow(.)))
 
 pre_outbreak_data <- salgado %>% 
@@ -239,9 +248,13 @@ nl@simdesign <- simdesign_lhs(nl,
 
 plan(list(sequential, multisession))
 
-results_outbreak_control <- progressr::with_progress(
-  run_nl_all(nl))
-write_rds(results_outbreak_control, file.path(out_path, "results_outbreak_control.rds"))
+if (run_sims) {
+  results_outbreak_control <- progressr::with_progress(
+    run_nl_all(nl))
+  write_rds(results_outbreak_control, file.path(out_path, "results_outbreak_control.rds"))
+}
+
[…]

### 2024-11-29 paper/index.qmd
paper poster abstract update

---
title: Paper
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
---

# Background

```{r, include=FALSE}

# I could not get R code to run before, until I did this:
# https://github.com/brentthorne/posterdown/issues/42

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
source(file.path("..", "calibrate.R"))
# load(file.path("..", "out", "calibrate.RData"))
```

Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection. Examples include *Clostridioides difficile* infection (CDI) and Vancomycin-resistant *Enterococcus*. In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents. To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by Salgado et al @salgado2009.

# Methods

## Agent-based model

An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.1999]. Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities. Next, the hospital was populated with patients, each having an infected/colonised or non-infected/colonised state, defined as:

$$
x_p(t) \epsilon \{0, 1\}
$$

where:

-   $x_p(t) = 0$ if patient $p$ is not infected/colonised at time $t$,
-   $x_p(t) = 1$ if patient $p$ is infected/colonised at time $t$. It was assumed that infection/colonisation was not reversible during the simulation, i.e., that infected/colonised patients remained so until discharge. Therefore, in contrast to a traditional susceptible-infected-recovered model [@anderson1991], this model was defined as:

$$
\frac{dH_S}{dt} = A_S - QH_S(t)H_I(t) - DH_S(t)
$$

$$
\frac{dH_I}{dt} = A_I + QH_S(t)H_I(t) - DH_I(t)
$$

where:

-   $H_S$ is the number of susceptible inpatients,
-   $H_I$ is the number of infected inpatients,
-   $A_S$ is the rate of new susceptible admissions,
-   $A_I$ is the rate of new infected admissions,
-   $Q$ is the rate of transmission,
-   $D$ is the rate of patients that are discharged.

The hospital capacity was constantly kept at 100% occupancy, therefore:

$$
D = A_S + A_I
$$

Users can run the model interactively using the interface shown in @fig-screenshot. Input parameters included patient factors (e.g., community colonisation rate), population behaviour (e.g., frequency of toilet use), and infection control measures (e.g., toilet cleaning frequency). Risk of infection changed as patients shared use of contaminated toilets, or were prescribed antimicrobial therapy.

## Data

Data from a *Clostridioides difficile* outbreak reported by Salgado et al, was used to calibrate the model.

Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data. The parameters in @tbl-consts were kept constant during the calibration. Using cases per 1000 patient days as a summary statistic, simulated data were compared to the actual reported rate of nosocomial CDI. Priors were determined from domain expertise or extracted from existing literature. Firstly, the model was calibrated to pre-outbreak data to identify parameters that simulated the baseline rate. Next, we simulated the emergence of an outbreak and its subsequent control by incrementally raising colonisation pressure over time (via increased antimicrobial consumption, toilet use, etc.) and enhancing infection control measures, respectively. Finally, the posterior parameter distributions were used to simulate 20 outbreaks.

| Name                                                       | Value    | Description                                                             | Reference                |
|------------------|------------------|--------------------|------------------|
| Baseline (pre-outbreak) antimicrobial prescription rate    | 0.319    | Proportion of patients administered antimicrobials during admission     | @versporten2018          |
| Admission duration                                         | 8.3 days | Mean of a Poisson distribution                                          | @thehealthfoundation2023 |
| Proportion of toilets that are shared between \>1 patients | 0.6      | The remaining proportion (0.4) of toilets were dedicated to one patient | N/A                      |

: Parameters that were assumed to be constant during calibration. {#tbl-consts}

![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="30%"}

# Results

```{r, echo=FALSE}
posteriors <- abc_params_outbreak_control$unadj.values
comm_colonisation <- round(mean(posteriors[, "community-colonisation-rate"] * 100), 2)
outbreak_colonisation <- round(mean(posteriors[, "o-community-colonisation-rate"] * 100), 2)
outbreak_antibiotic_rate <- round(mean(posteriors[, "o-antibiotic-prescription-rate"] * 100), 2)
control_antibiotic_rate <- round(mean(posteriors[, "c-antibiotic-prescription-rate"] * 100), 2)
```

After calibration, key parameters included: community colonisation rate of `r comm_colonisation`%, rising to `r outbreak_colonisation`% during the outbreak period. Furthermore, the proportion of patients prescribed an antimicrobial during their admission increased from `r consts[["antibiotic-prescription-rate"]] * 100`% to `r outbreak_antibiotic_rate`% during the outbreak; then reduced to `r control_antibiotic_rate`% during the period of enhanced infection control measures. The posterior parameter distributions are shown in @fig-posterior-densities. Simulated data had a pre-outbreak baseline median rate of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\]. During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).

```{r}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| warning: false
#| label: fig-posterior-densities
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
#| fig-subcap:
#|   - "General parameters"
#|   - "Toilet-related parameters"
#| layout-ncol: 1
abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))

abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~name, scales = "free_x") +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))
```

```{r}
#| echo: false
#| label: fig-simulated-rates
#| fig.cap: "Observed and simulated rates of _C. difficile_ infection over the simulated period. Approximate Bayesian computation was used to determine the unobserved parameters of the agent-based model. The red line shows the true observed infection rates. The black solid line shows the mean simulated rates; the dashed lines show the standard deviation."
#| fig.width: 18
#| fig.height: 8
ggplot(long_observed_ss, aes(x = x, y = rates)) + geom_line(color = 'red') +
  geom_line(data = sim_mean, aes(x = x, y = m)) +
  geom_line(data = sim_mean, aes(x = x, y = q_high), linetype='dashed') +
  geom_line(data = sim_mean, aes(x = x, y = q_low), linetype='dashed') + 
  theme(text=element_text(size=24)) +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

```{r}
#| echo: false
#| fig.width: 18
#| fig.height: 8
#| include: false
#| message: false
consts_sim_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  #slice_sample(n=10) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(siminputrow, date_sim) %>%
  summarise(rate = mean(rate)) %>%
  ggplot(aes(x = date_sim,
             y = rate,
             color = factor(`siminputrow`))) +
  geom_line() +
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
  guides(color="none")
```

# Discussion

Repeated simulations using the model generated nosocomial CDI rates that showed a clear outbreak peak, followed by control of colonisation/infection through a change in infection control practice, resembling a real hospital CDI outbreak. The interactive nature of the model can be used for outbreak preparedness by testing out different infection control strategies. In summary, we report a pathogen-agnostic model that can be used to simulate any infection with features such as faecal-oral transmission, colonisation, and antimicrobial selective pressure effects.

### 2024-12-12 paper/index.qmd
manuscript

@@
     corresponding: true
-bibliography: references.bib
 ---
@@
 It was assumed that infection/colonisation was not reversible during the simulation, i.e., that infected/colonised patients remained so until discharge.
-Therefore, in contrast to a traditional susceptible-infected-recovered model [@anderson1991], this model was defined as:
+Hence, in contrast to a traditional susceptible-infected-recovered model [@anderson1991], the differential equations of this model were:
 
@@
 
+-   \$dH_s/dt\$ is the rate of
 -   $H_S$ is the number of susceptible inpatients,
@@
 -   $A_I$ is the rate of new infected admissions,
--   $Q$ is the rate of transmission,
+-   $Q$ is the transmission coefficient,
 -   $D$ is the rate of patients that are discharged.
 
-The hospital capacity was constantly kept at 100% occupancy, therefore:
+The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore:
 
@@
 $$
+
+At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:
+
+$$ 
+P_\text{de novo}(t) = 1 - \exp {( - r_0 T(t) )}
+$$
+
+and 2) the usage of contaminated toilets:
+
+$$
+P_\text{toilet}(t) = 1 - \exp ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
+$$
+
+where:
+
+Administration of antimicrobial therapy modified the risk of infection/colonisation.
+
+$$
+R_i(t) = r_0 \cdot T(t) + \sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]
+$$
+
+where:
+
+-   $R_i(t)$ is the risk of patient $i$ acquiring infection/colonisation at time $t$,
+
+-   $r_0$ is the baseline/*de novo* risk of infection/colonisation,
+
+-   $T(t)$ is a binary indicator for antimicrobial therapy:
+
+$$
+\mathrm{T}(t) = \begin{cases}
+    m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
+    1 & \text{otherwise.}
+\end{cases}
+$$
+
+-   $m$ is the increased (or decreased) relative risk of infection/colonisation due to antimicrobial therapy,
 
@@
 | Name                                                       | Value    | Description                                                             | Reference                |
-|------------------|------------------|--------------------|------------------|
+|------------------|------------------|-------------------|------------------|
 | Baseline (pre-outbreak) antimicrobial prescription rate    | 0.319    | Proportion of patients administered antimicrobials during admission     | @versporten2018          |
@@
 ```{r}
+#| eval: true
 #| echo: false

### 2024-12-13 paper/index.qmd
methods

@@
     corresponding: true
+bibliography: references.bib
 ---
@@
 
--   \$dH_s/dt\$ is the rate of
+-   $dH_s/dt$ is the rate of
 -   $H_S$ is the number of susceptible inpatients,
@@
 
-Administration of antimicrobial therapy modified the risk of infection/colonisation.
-
-$$
-R_i(t) = r_0 \cdot T(t) + \sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]
-$$
-
-where:
-
 -   $R_i(t)$ is the risk of patient $i$ acquiring infection/colonisation at time $t$,
@@
 
-Users can run the model interactively using the interface shown in @fig-screenshot.
-Input parameters included patient factors (e.g., community colonisation rate), population behaviour (e.g., frequency of toilet use), and infection control measures (e.g., toilet cleaning frequency).
-Risk of infection changed as patients shared use of contaminated toilets, or were prescribed antimicrobial therapy.
+-   $U_i(t)$ is the number of times patient $i$ frequents the toilet within the time interval $t$,
 
-## Data
+-   $C(t_j)$ is the contamination level of the toilet at time $t_j$.
 
-Data from a *Clostridioides difficile* outbreak reported by Salgado et al, was used to calibrate the model.
+The agent-based model conducted Bernoulli trials to determine if a patient acquired infection/colonisation at each time step $t$:
 
-Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data.
-The parameters in @tbl-consts were kept constant during the calibration.
-Using cases per 1000 patient days as a summary statistic, simulated data were compared to the actual reported rate of nosocomial CDI.
-Priors were determined from domain expertise or extracted from existing literature.
+$$
+X_\text{de novo} \sim \text{Bernoulli}(P_\text{de novo}(t)),
+$$
+
+where $X_\text{de novo}$ = 1 if patient acquires infection/colonisation *de novo* at time $t$, and 0 otherwise.
+
+$$
+X_\text{toilet} \sim \text{Bernoulli}(P_\text{toilet}(t)),
+$$
+
+where $X_\text{toilet}$ = 1 if patient acquires infection/colonisation from a contaminated toilet at time $t$, and 0 otherwise.
+
+The results of the trials were combined to determine the outcome for patient $i$ at time $t$:
+
+$$
+X_i(t) = \max(X_\text{de novo}, X_\text{toilet}),
+$$
+
+where $X_i(t)$ = 1 if patient $i$ acquires infection/colonisation at time $t$, and 0 otherwise.
+
+## Model inputs and outputs
+
+Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community colonisation rate, antimicrobial prescription rate); and 3) colonisation parameters (e.g., relative risk increase of antimicrobial therapy on colonisation/infection).
+
+The outputs of the model included: 1) number of patients admitted; 2) number of patients with community-acquired colonisation/infection; 3) number of patients with hospital-acquired (nosocomial) colonisation/infection.
+The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.
+
+## Data sources
+
+Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009].
+Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchell2020].
+
+## Model calibration
+
+Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data [@sunnåker2013].
+Priors were determined from domain knowledge or extracted from existing literature.
+The parameters in @tbl-consts were kept constant during the calibration process.
+
+Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$.
+Next, each sample parameter point $\theta_i$ was used as an input to the agent-based model, generating a total of $n$ simulated datasets.
+The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
+Results from the first 30 simulated days were discarded to ensure that the model had reached a steady state.
+
+Each simulated summary statistic $S_i$ was compared to the summary statistic generated from the observed data, $S_\text{obs}$, and $S_i$ was accepted if:
+
+$$
+d(S_i, S_\text{obs}) \leq \epsilon,
+$$
+
+where:
+
+-   $d$ is the Euclidean distance between $S_i$ and $S_\text{obs}$, and
+
+-   $\epsilon$ is the tolerance.
+
+Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to generate a posterior distribution.
+
 Firstly, the model was calibrated to pre-outbreak data to identify parameters that simulated the baseline rate.
@@
 ![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="30%"}
+
+Users can run the model interactively using the interface shown in @fig-screenshot.
+
+## Outbreak generation
+
+To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.
+Changes in parameters such as increased antimicrobial prescription rates and decreased toilet cleaning frequency were used to encourage increased disease transmission, leading to an outbreak.
+On the other hand, these parameters were reversed to control the outbreak.
+
+## Model availability
 

### 2024-12-13 paper/index.qmd
deterministic version

@@
 $$
-\frac{dH_S}{dt} = A_S - QH_S(t)H_I(t) - DH_S(t)
+\frac{dH_S}{dt} = A_S - Q\frac{H_S(t)}{H_I(t)} - DH_S(t)
 $$
@@
 $$
-\frac{dH_I}{dt} = A_I + QH_S(t)H_I(t) - DH_I(t)
+\frac{dH_I}{dt} = A_I + Q\frac{H_S(t)}{H_I(t)} - DH_I(t)
 $$
@@
 
+```{r}
+single_outbreak_sim_mod <- single_outbreak_sim %>% 
+  mutate(prop_infected = `current-colonised` / `current-inpatients`,
+         current_not_infected = `current-inpatients` - `current-colonised`)
+single_outbreak_sim_mod %>% 
+  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
+  geom_line() +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment")
+
+# single_outbreak_sum_stat <- single_outbreak_sim_mod %>% 
+#   group_by(`[step]`) %>%
+#   summarise(m = mean(prop_infected))
+```
+
+```{r}
+# Parameters and time-dependent functions
+parameters <- list(
+  p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
+  A = 50        # Total admission/discharge rate
+)
+# 
+# baselinerate <- seq(from=0.01, to=0.99, length.out = 5)
+# outbreakrate <- seq(0.01, 0.99, length.out = 5)
+# rate_grid <- expand.grid(baselinerate, outbreakrate)
+
+# Time-varying beta function (example: sinusoidal variation over time)
+beta_function <- function(t, baseline_rate, outbreak_rate) {
+  if (t > unique(single_outbreak_sim$`outbreak-start`) &
+      t < unique(single_outbreak_sim$`outbreak-end`)) {
+    return (outbreak_rate)
+  }
+  return (baseline_rate)
+}
+
+# Initial conditions
+initial_state <- c(N = 600, I = 71)  # Initial populations of non-infected and infected patients
+time <- seq(0, run_to, by = 1)         # Time steps
+dt <- 1                             # Time step size
+# 
+# plan(multisession)
+# furrr::future_map2(rate_grid$Var1, rate_grid$Var2,
+#         \(x, y) {
+#           print("1")
+#           N <- numeric(length(time))
+#           I <- numeric(length(time))
+#           N[1] <- initial_state["N"]
+#           I[1] <- initial_state["I"]
+#           
+#           # Euler method for numerical integration with time-varying parameters
+#           for (t in 1:(length(time) - 1)) {
+#             # Current beta value
+#             beta_t <- beta_function(time[t], x, y)
+#             
+#             # Total population at current time
+#             total <- N[t] + I[t]
+#             
+#             # Proportions of non-infected and infected
+#             prop_N <- ifelse(total > 0, N[t] / total, 0)
+#             prop_I <- ifelse(total > 0, I[t] / total, 0)
+#             
+#             # Calculate derivatives
+#             dN <- parameters$p * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
+#             dI <- (1 - parameters$p) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+#             
+#             # Update populations using Euler's method
+#             N[t + 1] <- N[t] + dN * dt
+#             I[t + 1] <- I[t] + dI * dt
+#           }
+#           I[31:length(I)] / ( N[31:length(N)] + I[31:length(I)] )
+#         })
+
+baselinerate <- 0.02
+outbreakrate <- 0.07
+
+# Initialize vectors to store results
+N <- numeric(length(time))
+I <- numeric(length(time))
+N[1] <- initial_state["N"]
+I[1] <- initial_state["I"]
+
+# Euler method for numerical integration with time-varying parameters
+for (t in 1:(length(time) - 1)) {
+  # Current beta value
+  beta_t <- beta_function(time[t], baselinerate, outbreakrate)
+
+  # Total population at current time
+  total <- N[t] + I[t]
+
+  # Proportions of non-infected and infected
+  prop_N <- ifelse(total > 0, N[t] / total, 0)
+  prop_I <- ifelse(total > 0, I[t] / total, 0)
+
+  # Calculate derivatives
+  dN <- parameters$p * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
+  dI <- (1 - parameters$p) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+
+  # Update populations using Euler's method
+  N[t + 1] <- N[t] + dN * dt
+  I[t + 1] <- I[t] + dI * dt
+}
+
+# Combine results into a data frame for plotting
+output_df <- data.frame(
+  time = time,
+  N = N,
+  I = I
+)
+
+output_df <- filter(output_df, time > 31)
+
+# Plot the results
+library(ggplot2)
+output_long <- reshape2::melt(output_df, id = "time")
+ggplot(output_long, aes(x = time, y = value, color = variable)) +
+  geom_line(size = 1) +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment") +
+  scale_color_manual(values = c("blue", "red"),
+                     labels = c("Non-Infected (N)", "Infected (I)")) +
+  theme_minimal()
+
+single_outbreak_sim_mod %>% 
+  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
+  geom_line() +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment") +
+  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
+  geom_line(data = output_df, aes(x = time, y = I), color = "red")
+```
+
 # Discussion

### 2024-12-13 calibrate.R

@@ -835,3 +835,290 @@ outbreak_sims_rates <- consts_sim_rates %>%
   summarise(across(rate, list(median = median, q_low = \(x) quantile(x, 0.25), q_high = \(x) quantile(x, 0.75), min = min, max = max)))
 
 # save.image(file.path(out_path, "calibrate.RData"))
+
+##================= sensitivity analysis ==================##
+
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = 0.319, # Vesporten 2018
+               "admission-days" = 8.3,
+               "bay-proportion" = 0.6)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-antibiotic-prescription-rate"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-antibiotic-prescription-rate" = list(min=0.1,
+                                                                                  max=0.7,
+                                                                                  qfun='qunif')
[…]

### 2024-12-14 paper/index.qmd
odes work

--- 2024-12-13 paper/index.qmd
@@
 
+```{r}
+single_outbreak_sim_mod <- single_outbreak_sim %>% 
+  mutate(prop_infected = `current-colonised` / `current-inpatients`,
+         current_not_infected = `current-inpatients` - `current-colonised`)
+single_outbreak_sim_mod %>% 
+  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  ggplot(aes(x = `[step]`, y = value, color = compartment)) +
+  geom_line() +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment")
+
+# single_outbreak_sum_stat <- single_outbreak_sim_mod %>% 
+#   group_by(`[step]`) %>%
+#   summarise(m = mean(prop_infected))
+```
+
+```{r}
+# Parameters and time-dependent functions
+parameters <- list(
+  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
+  A = 50        # Total admission/discharge rate
+)
+#
+
+single_seeds <- single_outbreak_sim$`random-seed`
+single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
+  filter(`random-seed` == single_seeds[1])
+
+target_baseline_rate <- single_outbreak_one_seed %>%
+  filter(`[step]` > 30 & `[step]` < 200) %>%
+  summarise(prop_infected = mean(prop_infected)) %>% 
+  pull(prop_infected)
+target_outbreak_rate <- single_outbreak_one_seed %>%
+  filter(`[step]` > 1100 & `[step]` < 1200) %>%
+  summarise(prop_infected = mean(prop_infected)) %>% 
+  pull(prop_infected)
+
+baselinerate <- seq(from=0.005, to=0.1, length.out = 100)
+outbreakrate <- seq(0.005, 0.1, length.out = 100)
+baselinecolonisation <- seq(0.9, 0.99, length.out = 100)
+outbreakcolonisation <- seq(0.9, 0.99, length.out = 100)
+rate_grid <- expand.grid(b_base = baselinerate,
+                         b_out = outbreakrate,
+                         c_base = baselinecolonisation,
+                         c_out = outbreakcolonisation)
+rate_grid <- rate_grid %>% 
+  filter(b_base < b_out & c_base > c_out) %>% 
+  slice_sample(n = 100)
+
+# Time-varying beta function (example: sinusoidal variation over time)
+beta_function <- function(t, baseline_rate, outbreak_rate) {
+  if (t > unique(single_outbreak_sim$`outbreak-start`) &
+      t < unique(single_outbreak_sim$`outbreak-end`)) {
+    return (outbreak_rate)
+  }
+  return (baseline_rate)
+}
+
+# Initial conditions
+initial_state <- c(N = 600, I = 17)  # Initial populations of non-infected and infected patients
+time <- seq(0, run_to, by = 1)         # Time steps
+dt <- 1                             # Time step size
+
+plan(multisession)
+ode_sims <- furrr::future_pmap_dfr(rate_grid,
+        \(b_base, b_out, c_base, c_out) {
+          N <- numeric(length(time))
+          I <- numeric(length(time))
+          N[1] <- initial_state["N"]
+          I[1] <- initial_state["I"]
+
+          # Euler method for numerical integration with time-varying parameters
+          for (t in 1:(length(time) - 1)) {
+            # Current beta value
+            beta_t <- beta_function(time[t], b_base, b_out)
+            p_t <- beta_function(time[t], c_base, c_out)
+
+            # Total population at current time
+            total <- N[t] + I[t]
+
+            # Proportions of non-infected and infected
+            prop_N <- ifelse(total > 0, N[t] / total, 0)
+            prop_I <- ifelse(total > 0, I[t] / total, 0)
+
+            # Calculate derivatives
+            dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
+            dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+
+            # Update populations using Euler's method
+            N[t + 1] <- N[t] + dN * dt
+            I[t + 1] <- I[t] + dI * dt
+          }
+          data.frame(baseline = mean(I[31:200] / ( N[31:200] + I[31:200] )),
+                     outbreak = mean(I[1100:1200] / ( N[1100:1200] + I[1100:1200] )))
+        })
+
+abc(c(target_baseline_rate, target_outbreak_rate), rate_grid, ode_sims[c("baseline", "outbreak")], tol=0.5, method='rejection') %>% 
+  summary()
+
+baselinerate <- 0.0236
+outbreakrate <- 0.0482
+
+# baselinecolonisation <- 1 - unique(single_outbreak_sim$`community-colonisation-rate`)
+# outbreakcolonisation <- 1 - unique(single_outbreak_sim$`o-community-colonisation-rate`)
+baselinecolonisation <- 0.9721
+outbreakcolonisation <- 0.9403
+
+# Initialize vectors to store results
+N <- numeric(length(time))
+I <- numeric(length(time))
+N[1] <- initial_state["N"]
+I[1] <- initial_state["I"]
+
+# Euler method for numerical integration with time-varying parameters
+for (t in 1:(length(time) - 1)) {
+  # Current beta value
+  beta_t <- beta_function(time[t], baselinerate, outbreakrate)
+  p_t <- beta_function(time[t], baselinecolonisation, outbreakcolonisation)
+
+  # Total population at current time
+  total <- N[t] + I[t]
+
+  # Proportions of non-infected and infected
+  prop_N <- ifelse(total > 0, N[t] / total, 0)
+  prop_I <- ifelse(total > 0, I[t] / total, 0)
+
+  # Calculate derivatives
+  dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
+  dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+
+  # Update populations using Euler's method
+  N[t + 1] <- N[t] + dN * dt
+  I[t + 1] <- I[t] + dI * dt
+}
+
+# Combine results into a data frame for plotting
+output_df <- data.frame(
+  time = time,
+  N = N,
+  I = I
+)
+
+output_df <- filter(output_df, time > 31)
+
+# Plot the results
+library(ggplot2)
+output_long <- reshape2::melt(output_df, id = "time")
+ggplot(output_long, aes(x = time, y = value, color = variable)) +
+  geom_line(size = 1) +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment") +
+  scale_color_manual(values = c("blue", "red"),
+                     labels = c("Non-Infected (N)", "Infected (I)")) +
+  theme_minimal()
+
+single_outbreak_one_seed %>% 
+  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  group_by(compartment, `[step]`) %>% 
+  summarise(m = mean(value)) %>% 
+  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
+  geom_line() +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment") +
+  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
+  geom_line(data = output_df, aes(x = time, y = I), color = "red")
+```
+
 # Discussion

### 2024-12-14 paper/index.qmd
simpler ODE

@@
 library(abc)
+library(lhs)
 source(file.path("..", "calibrate.R"))
@@
 -   $x_p(t) = 1$ if patient $p$ is infected/colonised at time $t$.
+
 It was assumed that infection/colonisation was not reversible during the simulation, i.e., that infected/colonised patients remained so until discharge.
-Hence, in contrast to a traditional susceptible-infected-recovered model [@anderson1991], the differential equations of this model were:
-
-$$
-\frac{dH_S}{dt} = A_S - Q\frac{H_S(t)}{H_I(t)} - DH_S(t)
-$$
-
-$$
-\frac{dH_I}{dt} = A_I + Q\frac{H_S(t)}{H_I(t)} - DH_I(t)
-$$
-
-where:
-
--   $dH_s/dt$ is the rate of
--   $H_S$ is the number of susceptible inpatients,
--   $H_I$ is the number of infected inpatients,
--   $A_S$ is the rate of new susceptible admissions,
--   $A_I$ is the rate of new infected admissions,
--   $Q$ is the transmission coefficient,
--   $D$ is the rate of patients that are discharged.
-
-The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore:
-
-$$
-D = A_S + A_I
-$$
-
 At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:
@@
 $$ 
-P_\text{de novo}(t) = 1 - \exp {( - r_0 T(t) )}
+P_\text{de novo}(t) = 1 - \exp {( - r_0 T(t) )},
 $$
@@
 where $X_i(t)$ = 1 if patient $i$ acquires infection/colonisation at time $t$, and 0 otherwise.
+
+A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991;@keeling2011]:
+
+$$
+\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S- rS
+$$
+
+$$
+\frac{dI}{dt} = (1 - p)A - D \cdot \frac{I}{S + I} + \beta S + rS
+$$
+
+where:
+
+-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
+-   $S$ is the number of susceptible inpatients,
+-   $I$ is the number of infected/colonised inpatients,
+-   $A$ is the rate of new admissions,
+-   $p$ is the proportion of new admissions that are susceptible/non-infected, while $(1 - p)$ is community colonisation/infection rate,
+-   $\beta$ is the transmission coefficient,
+-   $r$ is the rate of random (_de novo_) colonisation/infection,
+-   $D$ is the rate of patient discharges.
+
+The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $D = A$.
 
@@
 
-baselinerate <- seq(from=0.005, to=0.1, length.out = 100)
-outbreakrate <- seq(0.005, 0.1, length.out = 100)
-baselinecolonisation <- seq(0.9, 0.99, length.out = 100)
-outbreakcolonisation <- seq(0.9, 0.99, length.out = 100)
-rate_grid <- expand.grid(b_base = baselinerate,
+randomLHS <- randomLHS(10000, 5)
+
+baselinerate <- qunif(randomLHS[, 1], 0.0005, 0.01)
+outbreakrate <- qunif(randomLHS[, 2], 0.0005, 0.01)
+baselinecolonisation <- qunif(randomLHS[, 3], 0.9, 0.99)
+outbreakcolonisation <- qunif(randomLHS[, 4], 0.9, 0.99)
+r <- qunif(randomLHS[, 5], 0.0001, 0.001)
+# rate_grid <- expand.grid(b_base = baselinerate,
+#                          b_out = outbreakrate,
+#                          c_base = baselinecolonisation,
+#                          c_out = outbreakcolonisation,
+#                          r = r)
+rate_grid <- data.frame(b_base = baselinerate,
                         b_out = outbreakrate,
                         c_base = baselinecolonisation,
-                         c_out = outbreakcolonisation)
+                        c_out = outbreakcolonisation,
+                        r = r)
 rate_grid <- rate_grid %>% 
   filter(b_base < b_out & c_base > c_out) %>% 
-  slice_sample(n = 100)
+  slice_sample(n = 300)
 
@@
 
+run_to <- max(single_outbreak_sim$`[step]`)
+
 # Initial conditions
@@
 ode_sims <- furrr::future_pmap_dfr(rate_grid,
-        \(b_base, b_out, c_base, c_out) {
+        \(b_base, b_out, c_base, c_out, r) {
           N <- numeric(length(time))
@@
             # Calculate derivatives
-            dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
-            dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+            dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t]  - r * N[t]
+            dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t]  + r * N[t]
 
@@
 
-abc(c(target_baseline_rate, target_outbreak_rate), rate_grid, ode_sims[c("baseline", "outbreak")], tol=0.5, method='rejection') %>% 
-  summary()
+abc_single_outbreak <- abc(c(target_baseline_rate, target_outbreak_rate), rate_grid, ode_sims[c("baseline", "outbreak")], tol=0.025, method='rejection')
 
-baselinerate <- 0.0236
-outbreakrate <- 0.0482
+baselinerate <- mean(abc_single_outbreak$unadj.values[, "b_base"])
+outbreakrate <- mean(abc_single_outbreak$unadj.values[, "b_out"])
 
@@
 # outbreakcolonisation <- 1 - unique(single_outbreak_sim$`o-community-colonisation-rate`)
-baselinecolonisation <- 0.9721
-outbreakcolonisation <- 0.9403
+baselinecolonisation <- mean(abc_single_outbreak$unadj.values[, "c_base"])
+outbreakcolonisation <- mean(abc_single_outbreak$unadj.values[, "c_out"])
+r <- mean(abc_single_outbreak$unadj.values[, "r"])
 
@@
   # Calculate derivatives
-  dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t] * prop_I
-  dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t] * prop_I
+  dN <- p_t * parameters$A - parameters$A * prop_N - beta_t * N[t]  - r * N[t]
+  dI <- (1 - p_t) * parameters$A - parameters$A * prop_I + beta_t * N[t]  + r * N[t]
 
@@
   pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  filter(`[step]` > 31) %>%
   group_by(compartment, `[step]`) %>% 

### 2024-12-14 calibrate.R
single outbreak sim

@@ -836,6 +836,105 @@ outbreak_sims_rates <- consts_sim_rates %>%
 
 # save.image(file.path(out_path, "calibrate.RData"))
 
+###========== single outbreak only simulation =========###
+
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = 0.319, # Vesporten 2018
+               "admission-days" = 8.3,
+               "bay-proportion" = 0.6)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_sd <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, sd)
+
+estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
+  list(min = x - y,
+       max = x + y,
+       qfun = 'qunif')
+})
+
+estimated_consts_fixed <- c(estimated_consts_mean, consts)
+
+nl@experiment <- experiment(expname = "single_outbreak_sim",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = consts, 
+                            variables = estimated_consts)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 1,
+                              nseeds = 10,
+                              precision = 3)
+
[…]

### 2024-12-17 calibrate.R
parameters tweaked

@@ -5,16 +5,22 @@ library(lubridate)
 library(abc)
 library(progressr)
 handlers("cli")
-wards_total <- 16
-bedspaces_per_ward <- 14
 set.seed(42)
 
 netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- "salgado_v0.42_day_time.nlogo"
+model_path <- "../salgado_v0.42_day_time.nlogo"
 out_path <- "../out/"
 data_path <- "../data/"
 
-run_sims <- FALSE
+# run_sims <- FALSE
+
+calibration_samples <- 100
+calibration_seeds <- 3
+abc_tol <- 0.05
+
+antibiotic_prescription_rate = 0.499 # Vesporten 2018
+admission_days <- 4.6
+bay_proportion <- 0.6
 
 ##============== pre-outbreak =============##
 
@@ -44,22 +50,22 @@ nl@experiment <- experiment(expname = "baseline_rate",
                                              "toilet-contamination-effect" = list(min=0.01, max=0.99, qfun='qunif'),
                                              "toilet-cleaning-effect" = list(min=0.01, max=0.99, qfun='qunif'),
                                              "toilet-frequenting-rate" = list(min=0.5, max=5, qfun='qunif'),
-                                             "antibiotic-effect" = list(min=0.1, max=12, qfun='qunif'),
-                                             "toilet-cleaning-rate" = list(min=0.5, max=5, qfun='qunif'),
-                                             "random-colonisation" = list(min=1, max=20, qfun='qunif'),
+                                             "antibiotic-effect" = list(min=1.31, max=1.87, qfun='qunif'),
+                                             "toilet-cleaning-rate" = list(min=0.5, max=3, qfun='qunif'),
+                                             "random-colonisation" = list(min=2.8, max=12.1, qfun='qunif'),
                                              "proportion-redistributed" = list(min=0.01, max=0.99, qfun='qunif')
                                              ),
                             constants = list("wards-total" = wards_total,
                                              "outbreak?" = FALSE,
                                              "infection-control?" = FALSE,
                                              "bedspaces-per-ward" = bedspaces_per_ward,
-                                             "antibiotic-prescription-rate" = 0.319, # Vesporten 2018
-                                             "admission-days" = 8.3,
-                                             "bay-proportion" = 0.6))
+                                             "antibiotic-prescription-rate" = antibiotic_prescription_rate,
[…]

### 2024-12-17 calibrate.R
more sens analysis

@@ -1259,3 +1259,304 @@ list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
   ggplot(aes(x = value, y = mean_cases)) + 
   geom_line() + 
   facet_wrap(~ param)
+
+#####============= sens toilet cleaning rate ==================##
+
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-toilet-cleaning-rate"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-toilet-cleaning-rate" = list(min=0.01,
+                                                                                 max=0.99,
+                                                                                 qfun='qunif')
[…]

### 2024-12-17 paper/index.qmd
paper update

@@
 
+run_sims <- FALSE
+
 # I could not get R code to run before, until I did this:
@@
 library(lhs)
+library(flextable)
 source(file.path("..", "calibrate.R"))
@@
 $$ 
-P_\text{de novo}(t) = 1 - \exp {( - r_0 T(t) )},
+P_\text{de novo}(t) = 1 - \text{Exp} {( - r_0 T(t) )},
 $$
@@
 $$
-P_\text{toilet}(t) = 1 - \exp ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
+P_\text{toilet}(t) = 1 - \text{Exp} ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
 $$
@@
 $$
-\mathrm{T}(t) = \begin{cases}
+T(t) = \begin{cases}
     m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
@@
 $$
-X_i(t) = \max(X_\text{de novo}, X_\text{toilet}),
+X_i(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
 $$
@@
 
-A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991;@keeling2011]:
+A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991; @keeling2011]:
 
 $$
-\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S- rS
+\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S - rS
 $$
@@
 -   $\beta$ is the transmission coefficient,
--   $r$ is the rate of random (_de novo_) colonisation/infection,
+-   $r$ is the rate of random (*de novo*) colonisation/infection,
 -   $D$ is the rate of patient discharges.
@@
 
-Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data [@sunnåker2013].
+Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnåker2013].
 Priors were determined from domain knowledge or extracted from existing literature.
-The parameters in @tbl-consts were kept constant during the calibration process.
+A summary of the priors used is available in @tbl-priors.
 
 Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$.
-Next, each sample parameter point $\theta_i$ was used as an input to the agent-based model, generating a total of $n$ simulated datasets.
+Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
 The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
-Results from the first 30 simulated days were discarded to ensure that the model had reached a steady state.
+Results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.
 
@@
 
-Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to generate a posterior distribution.
+Otherwise, $S_i$ was rejected.
+The parameter samples that generated the accepted simulations were used to produce a posterior distribution.
+$\epsilon$ was set so that the acceptance rate was approximately 2.5%.
 
-Firstly, the model was calibrated to pre-outbreak data to identify parameters that simulated the baseline rate.
-Next, we simulated the emergence of an outbreak and its subsequent control by incrementally raising colonisation pressure over time (via increased antimicrobial consumption, toilet use, etc.) and enhancing infection control measures, respectively.
-Finally, the posterior parameter distributions were used to simulate 20 outbreaks.
+The approach to calibration was sequential.
+Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate.
+Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgado2009].
+To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset.
+The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
+At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
+The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.
 
-| Name                                                       | Value    | Description                                                             | Reference                |
-|------------------|------------------|-------------------|------------------|
-| Baseline (pre-outbreak) antimicrobial prescription rate    | 0.319    | Proportion of patients administered antimicrobials during admission     | @versporten2018          |
-| Admission duration                                         | 8.3 days | Mean of a Poisson distribution                                          | @thehealthfoundation2023 |
-| Proportion of toilets that are shared between \>1 patients | 0.6      | The remaining proportion (0.4) of toilets were dedicated to one patient | N/A                      |
+The posterior parameter distributions were used to simulate 20 outbreak scenarios. Parameter samples were drawn from a uniform distribution:
 
-: Parameters that were assumed to be constant during calibration. {#tbl-consts}
+$$
+U_[a,b] = U(\mu - \sigma, \mu + \sigma),
+$$
+
+where: $\mu$ and \$\sigma\$ are the mean and standard deviation of the posterior distribution, respectively.
+The simulated rates of nosocomial CDI were compared to the observed rates to assess the model's performance.
+One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.
+Since the agent-based model was stochastic, all simulations were aggregated
+
+| Name                                                       | Distribution | Parameters                                                                                                   | Description                                                                     | Reference                                          |
+|---------------|---------------|---------------|---------------|---------------|
+| Baseline (pre-outbreak) antimicrobial prescription rate    | Constant     | 49.9%                                                                                                        | Proportion of patients administered antimicrobials during admission             | @versporten2018                                    |
+| Admission duration                                         | Poisson      | $\lambda = 4.6  \text{ days}$                                                                                |                                                                                 | @levitkryankelixhauserastrangesekassedccoffeyr2007 |
+| Proportion of toilets that are shared between \>1 patients | Constant     | 60%                                                                                                          | The remaining proportion (40%) of toilets were dedicated to one patient         | N/A                                                |
+| Community colonisation rate                                | Uniform      | $a = 0.01, b = 0.1$                                                                                          |                                                                                 | @crobach2018                                       |
+| Antimicrobial effect                                       | Uniform      | $ a = 1.31, b = 1.87 $                                                                                     | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimings2014                                      |
+| CDI incidence                                              | Uniform      | $a=2.8, b=12.1$                               | Per 10,000 patient-days                                                         | @marra2020                                         |
+| Mean toilet frequenting rate                               | Uniform      | $a=0.5,b=5$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
+| Mean toilet cleaning rate                                  | Uniform      | $a=0.5, b=3$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
+| Toilet contamination effect                                | Uniform      | $a=0.01,b=0.99$                               | Contamination impact by infected/colonised patients through toilet use          |                                                    |
+| Toilet cleaning effect                                     | Uniform      | $a=0.01, b=0.99$                               | Impact of toilet cleaning on contamination                                      |                                                    |
+
+: Parameters that were assumed to be constant during calibration. {#tbl-priors}
 
@@
 To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.
-Changes in parameters such as increased antimicrobial prescription rates and decreased toilet cleaning frequency were used to encourage increased disease transmission, leading to an outbreak.
+During calibration, the upper or lower bounds for parameter priors during the outbreak period were extended by 50% of baseline.
+For example, the upper bound of prior for the parameter determining the rate of antimicrobial prescriptions was 50% higher than the baseline rate.
 On the other hand, these parameters were reversed to control the outbreak.
 
-## Model availability
+## Software implementation
+
+The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.1999].
+The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@salecker2019; @rcoreteam2022].
 
@@
 control_antibiotic_rate <- round(mean(posteriors[, "c-antibiotic-prescription-rate"] * 100), 2)
+```
+
+```{r, include=T}
+params_table <- abc_params_outbreak_control$unadj.values %>% 
+  as_tibble() %>% 
+  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
+  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
+  summarise(across(everything(), ~ paste0(
+    round(mean(.x), 2),
+    " (",
+    round(sd(.x), 2),
+    ")"
+    )))
+
+param_units <- c("toilet-frequenting-rate" = "times per day",
+                 "toilet-cleaning-rate" = "times per day",
+                 "toilet-contamination-effect" = "",
+                 "toilet-cleaning-effect" = "",
+                 "community-colonisation-rate" = "proportion of patients",
+                 "antibiotic-prescription-rate" = "proportion of admissions",
+                 "antibiotic-effect" = "relative risk",
+                 "random-colonisation" = "cases per 10,000 bed days")
+
+param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
+                 "toilet-cleaning-rate" = "Toilet cleaning rate",
+                 "toilet-contamination-effect" = "Toilet contamination effect",
+                 "toilet-cleaning-effect" = "Toilet cleaning effect",
+                 "community-colonisation-rate" = "Community colonisation rate",
+                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
+                 "antibiotic-effect" = "Antibiotic effect",
+                 "random-colonisation" = "Random colonisation")
+
+params_table %>% 
+  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
+               names_pattern = "(b|o|c)-(.*)") %>% 
+  pivot_wider(names_from = period, values_from = value) %>% 
+  # reorder cols
+  dplyr::select(param, b, o, c) %>% 
+  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
+  mutate(units = param) %>% 
+  relocate(units, .after = param) %>%
+  flextable() %>% 
+  labelizor(j = "units", part = "body",
+            labels = param_units) %>% 
+  labelizor(j = "param", part = "body",
+            labels = param_names)
+
+params_table %>% 
+  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
+               names_pattern = "(b|o|c)-(.*)") %>% 
+  pivot_wider(names_from = period, values_from = value) %>% 
+  # reorder cols
+  dplyr::select(param, b, o, c) %>% 
+  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
+  mutate(units = param) %>% 
+  relocate(units, .after = param) %>%
+  knitr::kable()
 ```
@@
 
-```{r}
+```{r, cache=TRUE}
 # Parameters and time-dependent functions
@@
 
-single_seeds <- single_outbreak_sim$`random-seed`
+single_seeds <- unique(single_outbreak_sim$`random-seed`)
 single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
-  filter(`random-seed` == single_seeds[1])
+  filter(`random-seed` == single_seeds[3])
 
@@
 # Initial conditions
-initial_state <- c(N = 600, I = 17)  # Initial populations of non-infected and infected patients
+total_patients <- unique(single_outbreak_one_seed$`current-inpatients`)
+init_N <- unique(single_outbreak_one_seed$`community-colonisation-rate`) * total_patients
+init_I <- total_patients - init_N
+initial_state <- c(N = init_N, I = init_I)  # Initial populations of non-infected and infected patients
 time <- seq(0, run_to, by = 1)         # Time steps
@@
   geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
-  geom_line(data = output_df, aes(x = time, y = I), color = "red")
+  geom_line(data = output_df, aes(x = time, y = I), color = "red") +
+  xlim(500, 2000)
 ```

### 2025-02-26 archive.R
started cleanup

@@ -1,100 +1,0 @@
-
-nl <- nl(nlversion = "6.2.2",
-         nlpath = netlogo_path,
-         modelpath = model_path,
-         jvmmem = 1024)
-
-
-nl@experiment <- experiment(expname = "bay_proportions",
-                            outpath = out_path,
-                            repetition = 1,
-                            tickmetrics = "true",
-                            idsetup = "setup",
-                            idgo = "go",
-                            runtime = sim_days,
-                            metrics = c("total-colonised",
-                                        "total-patients-admitted",
-                                        "current-community-infections",
-                                        "current-hospital-infections",
-                                        "current-inpatients",
-                                        "current-colonised"),
-                            variables = list("bay-proportion" = list(values = seq(0.1, 0.9, 0.1))),
-                            constants = list("wards-total" = 4,
-                                             "bedspaces-per-ward" = 12,
-                                             "toilet-frequenting-rate" = 2.0,
-                                             "toilet-contamination-effect" = 0.2,
-                                             "toilet-cleaning-effect" = 0.9,
-                                             "toilet-cleaning-rate" = 2,
-                                             "community-colonisation-rate" = 0.03,
-                                             "discharge-rate" = 0.20))
-
-nl@simdesign <- simdesign_distinct(nl, 1)
-
-plan(list(sequential, multisession))
-results <- run_nl_all(nl)
-
-results <- results %>% 
-  filter(`[step]` > 0)
-
-dates <- seq(from = date("2024-01-01"), to = date("2024-12-31"), by = 1)
-dates <- dates[1:sim_days]
-results <- results %>% 
-  group_by(siminputrow) %>% 
-  mutate(date = dates) %>% 
-  mutate(month = month(date))
-
-results %>% 
-  group_by(`bay-proportion`, month) %>% 
-  summarise(hosp_rate = sum(`current-hospital-infections`) / sum(`current-inpatients`)) %>% 
-  ggplot(aes(x = month, y = hosp_rate)) + geom_line() + facet_wrap(~ factor(`bay-proportion`))
-
-
-# assuming bay proportion 0.6, try to find community rate
-
-nl <- nl(nlversion = "6.2.2",
-         nlpath = netlogo_path,
-         modelpath = model_path,
-         jvmmem = 1024)
-
-nl@experiment <- experiment(expname = "bay_proportions",
-                            outpath = out_path,
-                            repetition = 1,
-                            tickmetrics = "true",
-                            idsetup = "setup",
[…]

### 2025-02-26 calibrate.R

@@ -5,10 +5,12 @@ library(lubridate)
 library(abc)
 library(progressr)
 handlers("cli")
+wards_total <- 9
+bedspaces_per_ward <- 9
 set.seed(42)
 
 netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- "../salgado_v0.42_day_time.nlogo"
+model_path <- "../models/salgado_v0.43.nlogo"
 out_path <- "../out/"
 data_path <- "../data/"
 

### 2025-02-27 paper/index.qmd
reorganise scripts

@@
 
-```{r, include=FALSE}
-
+```{r setup, include=FALSE}
+# here::i_am("paper/index.qmd")
 run_sims <- FALSE
@@
 library(flextable)
-source(file.path("..", "calibrate.R"))
+```
+
+```{r}
+# print(here::here())
+# file.exists(here::here("scripts", "calibrate.R"))
+```
+
+```{r load-data, include=FALSE}
+# print(file.exists(here::here("scripts", "calibrate.R")))
+# source(here::here("scripts", "calibrate.R"))
+source(file.path("..", "scripts", "calibrate.R"), chdir = TRUE)
 # load(file.path("..", "out", "calibrate.RData"))
@@
 In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents.
-To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by Salgado et al @salgado2009.
+To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by @salgadoAnalysisOutbreakClostridium2009.
 
@@
 
-A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991; @keeling2011]:
+A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
 
@@
 Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009].
-Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchell2020].
+Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].
 
@@
 
-Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnåker2013].
+Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnakerApproximateBayesianComputation2013].
 Priors were determined from domain knowledge or extracted from existing literature.
@@
 |---------------|---------------|---------------|---------------|---------------|
-| Baseline (pre-outbreak) antimicrobial prescription rate    | Constant     | 49.9%                                                                                                        | Proportion of patients administered antimicrobials during admission             | @versporten2018                                    |
-| Admission duration                                         | Poisson      | $\lambda = 4.6  \text{ days}$                                                                                |                                                                                 | @levitkryankelixhauserastrangesekassedccoffeyr2007 |
+| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
+| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
 | Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
-| Community colonisation rate                                | Uniform      | $a = 0.01, b = 0.1$                                                                                          |                                                                                 | @crobach2018                                       |
-| Antimicrobial effect                                       | Uniform      | $ a = 1.31, b = 1.87 $                                                                                     | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimings2014                                      |
-| CDI incidence                                              | Uniform      | $a=2.8, b=12.1$                               | Per 10,000 patient-days                                                         | @marra2020                                         |
+| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
+| Antimicrobial effect | Uniform | \$ a = 1.31, b = 1.87 \$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
+| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
 | Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
@@
 
-The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.1999].
-The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@salecker2019; @rcoreteam2022].
+The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.NetLogo1999].
+The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@rcoreteamLanguageEnvironmentStatistical2022; @saleckerNlrxPackageNextgeneration2019].
 
@@
 ```{r, include=T}
+#| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
+#| label: tbl-posteriors
+#| echo: false
 params_table <- abc_params_outbreak_control$unadj.values %>% 
@@
   labelizor(j = "param", part = "body",
-            labels = param_names)
-
-params_table %>% 
-  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
-               names_pattern = "(b|o|c)-(.*)") %>% 
-  pivot_wider(names_from = period, values_from = value) %>% 
-  # reorder cols
-  dplyr::select(param, b, o, c) %>% 
-  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
-  mutate(units = param) %>% 
-  relocate(units, .after = param) %>%
-  knitr::kable()
+            labels = param_names) %>% 
+  width(width = 1) %>% 
+  font(fontname="Times New Roman",
+       part = "all") %>% 
+  fontsize(size = 10, part = "all")
 ```
@@
 output_long <- reshape2::melt(output_df, id = "time")
-ggplot(output_long, aes(x = time, y = value, color = variable)) +
-  geom_line(size = 1) +
-  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
-       x = "Time",
-       y = "Number of Patients",
-       color = "Compartment") +
-  scale_color_manual(values = c("blue", "red"),
-                     labels = c("Non-Infected (N)", "Infected (I)")) +
-  theme_minimal()
 

### 2025-02-27 calibrate.R

@@ -1,1564 +1,0 @@
-library(nlrx)
-library(future)
-library(tidyverse)
-library(lubridate)
-library(abc)
-library(progressr)
-handlers("cli")
-wards_total <- 9
-bedspaces_per_ward <- 9
-set.seed(42)
-
-netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- "../models/salgado_v0.43.nlogo"
-out_path <- "../out/"
-data_path <- "../data/"
-
-# run_sims <- FALSE
-
-calibration_samples <- 100
-calibration_seeds <- 3
-abc_tol <- 0.05
-
-antibiotic_prescription_rate = 0.499 # Vesporten 2018
-admission_days <- 4.6
-bay_proportion <- 0.6
-
-##============== pre-outbreak =============##
-
-sim_days <- (365 * 3) + 30
-# sim_hours <- sim_days * 24
-
-nl <- nl(nlversion = "6.2.2",
-         nlpath = netlogo_path,
-         modelpath = model_path,
-         jvmmem = 1024)
-
-nl@experiment <- experiment(expname = "baseline_rate",
-                            outpath = out_path,
-                            repetition = 1,
-                            tickmetrics = "true",
-                            idsetup = "setup",
-                            idgo = "go",
-                            runtime = sim_days,
-                            metrics = c("total-colonised",
-                                        "total-patients-admitted",
-                                        "total-hospital-infections",
-                                        "current-community-infections",
-                                        "current-hospital-infections",
-                                        "current-inpatients",
-                                        "current-colonised"),
-                            variables = list("community-colonisation-rate" = list(min=0.01, max=0.1, qfun='qunif'),
-                                             "toilet-contamination-effect" = list(min=0.01, max=0.99, qfun='qunif'),
-                                             "toilet-cleaning-effect" = list(min=0.01, max=0.99, qfun='qunif'),
-                                             "toilet-frequenting-rate" = list(min=0.5, max=5, qfun='qunif'),
-                                             "antibiotic-effect" = list(min=1.31, max=1.87, qfun='qunif'),
-                                             "toilet-cleaning-rate" = list(min=0.5, max=3, qfun='qunif'),
-                                             "random-colonisation" = list(min=2.8, max=12.1, qfun='qunif'),
-                                             "proportion-redistributed" = list(min=0.01, max=0.99, qfun='qunif')
[…]

### 2025-02-27 scripts/calibrate.R

@@ -1,0 +1,85 @@
+library(nlrx)
+library(future)
+library(tidyverse)
+library(lubridate)
+library(abc)
+library(progressr)
+library(here)
+handlers("cli")
+wards_total <- 9
+bedspaces_per_ward <- 9
+set.seed(42)
+
+netlogo_path <- "/Applications/NetLogo 6.2.2/"
+model_path <- here::here("models", "salgado_v0.43.nlogo")
+out_path <- here::here("out")
+data_path <- here::here("data")
+
+# run_sims <- FALSE
+
+calibration_samples <- 100
+calibration_seeds <- 3
+abc_tol <- 0.05
+
+antibiotic_prescription_rate = 0.499 # Vesporten 2018
+admission_days <- 4.6
+bay_proportion <- 0.6
+
+##============== pre-outbreak =============##
+
+source(here::here("scripts", "tune", "pre_outbreak.R"))
+
+##============== outbreak and control ===============##
+
+source(here::here("scripts", "tune", "outbreak_and_control.R"))
+
+##=========== fine tune pre-outbreak ===================================##
+
+source(here::here("scripts", "tune", "fine_tune_pre_outbreak.R"))
+
+##=========== use variables for pre-outbreak params as well ============##
+
+source(here::here("scripts", "tune", "variable_pre_outbreak.R"))
+
+##================= simulate using consts only ==================##
+
+source(here::here("scripts", "tune", "simulate_using_consts.R"))
+
+
+###========== single outbreak only simulation =========###
+
+source(here::here("scripts", "tune", "single_outbreak.R"))
+
+# save.image(file.path(out_path, "calibrate.RData"))
+
+##================= sensitivity analysis ==================##
+
+source(here::here("scripts", "sensitivity", "abx_prescription_rates.R"))
+
+#####============= sens proportion redistributed
+
+source(here::here("scripts", "sensitivity", "proportion_redistributed.R"))
+
+#####============= sens toilet cleaning rate ==================##
+
+source(here::here("scripts", "sensitivity", "toilet_cleaning_rate.R"))
+
+####================ sensitivity toilet cleaning effect ==============####
+
+source(here::here("scripts", "sensitivity", "toilet_cleaning_effect.R"))
+
+list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
+     proportion_redistributed = sens_prop_redist_outbreak_cases,
[…]

### 2025-02-27 scripts/sensitivity/abx_prescription_rates.R

@@ -1,0 +1,134 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-antibiotic-prescription-rate"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-antibiotic-prescription-rate" = list(min=0.1,
+                                                                                     max=0.7,
+                                                                                     qfun='qunif')
+                            )
+)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 20,
+                              nseeds = 10,
+                              precision = 3)
+
+plan(list(sequential, multisession))
+
+if (run_sims) {
[…]

### 2025-02-27 scripts/sensitivity/proportion_redistributed.R

@@ -1,0 +1,147 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-proportion-redistributed"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-proportion-redistributed" = list(min=0.01,
+                                                                                 max=0.99,
+                                                                                 qfun='qunif')
+                            )
+)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 20,
+                              nseeds = 10,
+                              precision = 3)
+
+plan(list(sequential, multisession))
+
[…]

### 2025-02-27 scripts/sensitivity/toilet_cleaning_effect.R

@@ -1,0 +1,134 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-toilet-cleaning-effect"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-toilet-cleaning-effect" = list(min=0.01,
+                                                                               max=0.99,
+                                                                               qfun='qunif')
+                            )
+)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 20,
+                              nseeds = 10,
+                              precision = 3)
+
[…]

### 2025-02-27 scripts/sensitivity/toilet_cleaning_rate.R

@@ -1,0 +1,147 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-toilet-cleaning-rate"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("c-toilet-cleaning-rate" = list(min=0.25,
+                                                                             max=4,
+                                                                             qfun='qunif')
+                            )
+)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 20,
+                              nseeds = 10,
+                              precision = 3)
[…]

### 2025-02-27 scripts/tune/fine_tune_pre_outbreak.R

@@ -1,0 +1,146 @@
+sim_days <- (365 * 3) + 30
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+pre_outbreak_priors <- abc_params$unadj.values
+pre_outbreak_priors_means <- apply(pre_outbreak_priors, 2, mean)
+pre_outbreak_priors_sd <- apply(pre_outbreak_priors, 2, sd)
+pre_outbreak_priors_lower_sd <- pre_outbreak_priors_means - pre_outbreak_priors_sd
+pre_outbreak_priors_upper_sd <- pre_outbreak_priors_means + pre_outbreak_priors_sd
+
+pre_outbreak_variables <- apply(pre_outbreak_priors, 2, \(x) {
+  list(min=unname(quantile(x, 0.25)),
+       max=unname(quantile(x, 0.75)),
+       qfun='qunif')
+})
+
+nl@experiment <- experiment(expname = "fine_tune",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            variables = pre_outbreak_variables,
+                            constants = list("wards-total" = wards_total,
+                                             "bedspaces-per-ward" = bedspaces_per_ward,
+                                             "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+                                             "outbreak?" = FALSE,
+                                             "infection-control?" = FALSE,
+                                             "admission-days" = admission_days,
+                                             "bay-proportion" = bay_proportion))
+
[…]

### 2025-02-27 scripts/tune/outbreak_and_control.R

@@ -1,0 +1,189 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado_v0.3.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = 0.319, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# use abc mean parameter estimates as constants for baseline
+consts <- c(consts, apply(abc_params$unadj.values, MARGIN = 2, mean))
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+# Set control measures start, which in Salgado was third week of November.
+# Calculate how many ticks into simulation this would be
+eicm_start <- as.numeric(difftime(ymd('2004-11-15'), ymd('2002-01-01'), units = 'days'))
+consts <- c(consts, "control-start" = eicm_start)
+
+# in Salgado, the outbreak starts in October 2004, calculate how many ticks/days
+# into simulation this is, then offset by 90 days in the search field
+outbreak_start <- as.numeric(difftime(ymd('2004-10-01'), ymd('2002-01-01'), units = 'days'))
+outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
+                                                   max=outbreak_start + 90,
+                                                   qfun='qunif'),
+                           "o-toilet-frequenting-rate" = list(min=consts[["toilet-frequenting-rate"]],
+                                                              max=8,
+                                                              qfun='qunif'),
+                           "o-toilet-contamination-effect" = list(min=consts[["toilet-contamination-effect"]],
+                                                                  max=0.99,
+                                                                  qfun='qunif'),
+                           "o-toilet-cleaning-effect" = list(min=0.01,
+                                                             max=consts[["toilet-cleaning-effect"]],
+                                                             qfun='qunif'),
+                           "o-toilet-cleaning-rate" = list(min=0.5,
+                                                           max=consts[["toilet-cleaning-rate"]],
+                                                           qfun='qunif'),
[…]

### 2025-02-27 scripts/tune/pre_outbreak.R

@@ -1,0 +1,143 @@
+sim_days <- (365 * 3) + 30
+# sim_hours <- sim_days * 24
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+nl@experiment <- experiment(expname = "baseline_rate",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            variables = list("community-colonisation-rate" = list(min=0.01, max=0.1, qfun='qunif'),
+                                             "toilet-contamination-effect" = list(min=0.01, max=0.99, qfun='qunif'),
+                                             "toilet-cleaning-effect" = list(min=0.01, max=0.99, qfun='qunif'),
+                                             "toilet-frequenting-rate" = list(min=0.5, max=5, qfun='qunif'),
+                                             "antibiotic-effect" = list(min=1.31, max=1.87, qfun='qunif'),
+                                             "toilet-cleaning-rate" = list(min=0.5, max=3, qfun='qunif'),
+                                             "random-colonisation" = list(min=2.8, max=12.1, qfun='qunif'),
+                                             "proportion-redistributed" = list(min=0.01, max=0.99, qfun='qunif')
+                            ),
+                            constants = list("wards-total" = wards_total,
+                                             "outbreak?" = FALSE,
+                                             "infection-control?" = FALSE,
+                                             "bedspaces-per-ward" = bedspaces_per_ward,
[…]

### 2025-02-27 scripts/tune/simulate_using_consts.R

@@ -1,0 +1,117 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_sd <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, sd)
+
+estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
+  list(min = x - y,
+       max = x + y,
+       qfun = 'qunif')
+})
+
+estimated_consts_fixed <- c(estimated_consts_mean, consts)
+
+nl@experiment <- experiment(expname = "consts_sim",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = consts, 
+                            variables = estimated_consts)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 20,
+                              nseeds = 10,
+                              precision = 3)
+
+plan(list(sequential, multisession))
+
[…]

### 2025-02-27 scripts/tune/single_outbreak.R

@@ -1,0 +1,96 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_sd <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, sd)
+
+estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
+  list(min = x - y,
+       max = x + y,
+       qfun = 'qunif')
+})
+
+estimated_consts_fixed <- c(estimated_consts_mean, consts)
+
+nl@experiment <- experiment(expname = "single_outbreak_sim",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = consts, 
+                            variables = estimated_consts)
+
+nl@simdesign <- simdesign_lhs(nl,
+                              samples = 1,
+                              nseeds = 10,
+                              precision = 3)
+
+plan(multisession)
+
+if (run_sims) {
+  single_outbreak_sim <- progressr::with_progress(
+    run_nl_all(nl))
[…]

### 2025-02-27 scripts/tune/variable_pre_outbreak.R

@@ -1,0 +1,239 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+# Set control measures start, which in Salgado was third week of November.
+# Calculate how many ticks into simulation this would be
+eicm_start <- as.numeric(difftime(ymd('2004-11-15'), ymd('2002-01-01'), units = 'days'))
+consts <- c(consts, "control-start" = eicm_start)
+
+# in Salgado, the outbreak starts in October 2004, calculate how many ticks/days
+# into simulation this is, then offset by 90 days in the search field
+outbreak_start <- as.numeric(difftime(ymd('2004-10-01'), ymd('2002-01-01'), units = 'days'))
+
+# also need to calculate the end date of the outbreak
+outbreak_end <- as.numeric(difftime(ymd('2005-05-31'), ymd('2002-01-01'), units = 'days'))
+
+# set an end to enhanced control measures, assume these were stepped down 
+# three months after end of outbreak
+control_end <- outbreak_end + 90
+
+pre_outbreak_priors <- abc_params$unadj.values
+pre_outbreak_priors_means <- apply(pre_outbreak_priors, 2, mean)
+pre_outbreak_priors_sd <- apply(pre_outbreak_priors, 2, sd)
+pre_outbreak_priors_lower_sd <- pre_outbreak_priors_means - pre_outbreak_priors_sd
+pre_outbreak_priors_upper_sd <- pre_outbreak_priors_means + pre_outbreak_priors_sd
+
+pre_outbreak_variables <- apply(pre_outbreak_priors, 2, \(x) {
+  list(min=unname(quantile(x, 0.25)),
+       max=unname(quantile(x, 0.75)),
+       qfun='qunif')
+})
+
+outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
+                                                   max=outbreak_start + 90,
+                                                   qfun='qunif'),
[…]

### 2025-03-01 scripts/calibrate.R
calibrate update

@@ -11,15 +11,11 @@ bedspaces_per_ward <- 9
 set.seed(42)
 
 netlogo_path <- "/Applications/NetLogo 6.2.2/"
-model_path <- here::here("models", "salgado_v0.43.nlogo")
+model_path <- here::here("models", "hospital_transmission.nlogo")
 out_path <- here::here("out")
 data_path <- here::here("data")
 
-# run_sims <- FALSE
-
-calibration_samples <- 100
-calibration_seeds <- 3
-abc_tol <- 0.05
+# run_sims <- TRUE
 
 antibiotic_prescription_rate = 0.499 # Vesporten 2018
 admission_days <- 4.6
@@ -27,18 +23,30 @@ bay_proportion <- 0.6
 
 ##============== pre-outbreak =============##
 
+calibration_samples <- 100
+calibration_seeds <- 3
+abc_tol <- 0.05
 source(here::here("scripts", "tune", "pre_outbreak.R"))
 
 ##============== outbreak and control ===============##
 
+calibration_samples <- 1000
+calibration_seeds <- 3
+abc_tol <- 0.025
 source(here::here("scripts", "tune", "outbreak_and_control.R"))
 
 ##=========== fine tune pre-outbreak ===================================##
 
+calibration_samples <- 1000
+calibration_seeds <- 3
+abc_tol <- 0.025
 source(here::here("scripts", "tune", "fine_tune_pre_outbreak.R"))
 
 ##=========== use variables for pre-outbreak params as well ============##
 
+calibration_samples <- 1000
+calibration_seeds <- 3
+abc_tol <- 0.025
 source(here::here("scripts", "tune", "variable_pre_outbreak.R"))
 
 ##================= simulate using consts only ==================##
@@ -48,6 +56,9 @@ source(here::here("scripts", "tune", "simulate_using_consts.R"))
 
 ###========== single outbreak only simulation =========###
 
+calibration_samples <- 100
+calibration_seeds <- 3
+abc_tol <- 0.05
 source(here::here("scripts", "tune", "single_outbreak.R"))
 
 # save.image(file.path(out_path, "calibrate.RData"))
@@ -67,6 +78,10 @@ source(here::here("scripts", "sensitivity", "toilet_cleaning_rate.R"))
 ####================ sensitivity toilet cleaning effect ==============####
 
 source(here::here("scripts", "sensitivity", "toilet_cleaning_effect.R"))
+
[…]

### 2025-03-01 paper/paper.qmd
paper update

--- 2025-02-27 paper/index.qmd
@@
 
-An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.1999].
+An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.NetLogo1999].
 Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities.
@@
 
-Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgado2009].
+Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009].
 Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].
 
-## Model calibration
+## Model calibration {#sec-model-calibration}
 
@@
 Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate.
-Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgado2009].
+Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
 To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset.
@@
 
-The posterior parameter distributions were used to simulate 20 outbreak scenarios. Parameter samples were drawn from a uniform distribution:
+The posterior parameter distributions were used to simulate `r length(unique(consts_sim$siminputrow))` outbreak scenarios. Parameter samples were drawn from a uniform distribution:
 
@@
 
-: Parameters that were assumed to be constant during calibration. {#tbl-priors}
+: Parameter prior distributions or constants. {#tbl-priors}
 
@@
 
-## Outbreak generation
+## Outbreak generation {#sec-outbreak-generation}
 
@@
 On the other hand, these parameters were reversed to control the outbreak.
+
+## Sensitivity analysis {#sec-sens-analysis}
+
+```{r, include=FALSE}
+sens_n_seeds <- length(unique(consts_sim$siminputrow))
+```
+
+To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
+Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after colonisation/infection; 3) toilet cleaning rate; and 4) toilet cleaning effect.
+To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection ABC.
+Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value.
+The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period.
+For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.
 
@@
 
-```{r, echo=FALSE}
+```{r, include=FALSE}
 posteriors <- abc_params_outbreak_control$unadj.values
@@
 control_antibiotic_rate <- round(mean(posteriors[, "c-antibiotic-prescription-rate"] * 100), 2)
+b_toilet_clean_rate <- round(mean(posteriors[, "toilet-cleaning-rate"]), 2)
+o_toilet_clean_rate <- round(mean(posteriors[, "o-toilet-cleaning-rate"]), 2)
+c_toilet_clean_rate <- round(mean(posteriors[, "c-toilet-cleaning-rate"]), 2)
 ```
 
-```{r, include=T}
+An agent-based model that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced.
+After calibrating the agent-based model using approximate Bayesian computation (ABC), three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control.
+A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection ABC) is shown in @tbl-posteriors.
+The densities of the posterior parameter distributions are shown in @fig-posterior-densities.
+In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).
+
+```{r tbl-posteriors, include=TRUE}
 #| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
-#| label: tbl-posteriors
 #| echo: false
@@
   as_tibble() %>% 
+  mutate("community-colonisation-rate" = `community-colonisation-rate` * 100,
+         "o-community-colonisation-rate" = `o-community-colonisation-rate` * 100,
+         "o-antibiotic-prescription-rate" = `o-antibiotic-prescription-rate` * 100,
+         "c-antibiotic-prescription-rate" = `c-antibiotic-prescription-rate` * 100) %>% 
   dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
@@
                  "toilet-cleaning-effect" = "",
-                 "community-colonisation-rate" = "proportion of patients",
-                 "antibiotic-prescription-rate" = "proportion of admissions",
+                 "community-colonisation-rate" = "percentage of patients",
+                 "antibiotic-prescription-rate" = "percentage of admissions",
                  "antibiotic-effect" = "relative risk",
-                 "random-colonisation" = "cases per 10,000 bed days")
+                 "random-colonisation" = "cases per 10,000 bed days",
+                 "proportion-redistributed" = "proportion of patients")
 
@@
                  "antibiotic-effect" = "Antibiotic effect",
-                 "random-colonisation" = "Random colonisation")
+                 "random-colonisation" = "Random colonisation",
+                 "proportion-redistributed" = "Proportion redistributed")
 
@@
 
-After calibration, key parameters included: community colonisation rate of `r comm_colonisation`%, rising to `r outbreak_colonisation`% during the outbreak period.
-Furthermore, the proportion of patients prescribed an antimicrobial during their admission increased from `r consts[["antibiotic-prescription-rate"]] * 100`% to `r outbreak_antibiotic_rate`% during the outbreak; then reduced to `r control_antibiotic_rate`% during the period of enhanced infection control measures.
-The posterior parameter distributions are shown in @fig-posterior-densities.
-Simulated data had a pre-outbreak baseline median rate of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
-During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
-
-```{r}
+```{r fig-posterior-densities, include=TRUE}
 #| eval: true
@@
 
-```{r}
+Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-simulated-rates).
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
+
+```{r fig-simulated-rates, include=TRUE}
 #| echo: false
@@
 
-```{r}
+```{r, include=FALSE}
 #| echo: false
@@
 #| fig.height: 8
-#| include: false
 #| message: false
@@
 
-```{r}
+```{r, include=FALSE}
 single_outbreak_sim_mod <- single_outbreak_sim %>% 
@@
 
-```{r, cache=TRUE}
+The sensitivity analysis explored the impact of different infection control practices on the number of cases observed during the outbreak period -- the results are shown in @fig-sensitivity-analysis.
+The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateaeu beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits.
+The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate.
+
+```{r fig-sensitivity-analysis, include=TRUE}
+#| echo: false
+list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
+     proportion_redistributed = sens_prop_redist_outbreak_cases,
+     toilet_cleaning_rate = sens_toilet_cleaning_rate_outbreak_cases,
+     toilet_cleaning_effect = sens_toilet_cleaning_effect_outbreak_cases) %>%
+  bind_rows() %>% 
+  pivot_longer(cols = starts_with("c-"),
+               names_to = "param",
+               values_drop_na = TRUE) %>% 
+  group_by(param, value) %>% 
+  summarise(mean_cases = mean(`outbreak-cases`),
+            sd_cases = sd(`outbreak-cases`),
+            .groups = "keep") %>% 
+  ggplot(aes(x = value, y = mean_cases)) + 
+  geom_line() + 
+  # shade sd
+  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
+  facet_wrap(~ param, scales = "free")
+```
+
+```{r}
+sens_prop_redist_perfect_clean_outbreak_cases %>% pivot_longer(cols = `c-proportion-redistributed`,
+               names_to = "param",
+               values_drop_na = TRUE) %>% 
+  group_by(param, value) %>% 
+  summarise(mean_cases = mean(`outbreak-cases`),
+            sd_cases = sd(`outbreak-cases`),
+            .groups = "keep") %>% 
+  ggplot(aes(x = value, y = mean_cases)) + 
+  geom_line() + 
+  # shade sd
+  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
+  facet_wrap(~ param, scales = "free")
+```
+
+```{r run-deterministic-model, include=FALSE, cache=TRUE}
 # Parameters and time-dependent functions
@@
 output_df <- filter(output_df, time > 31)
+```
 
-# Plot the results
-library(ggplot2)
+```{r fig-single-outbreak, include=TRUE}
+#| echo: false
 output_long <- reshape2::melt(output_df, id = "time")

### 2025-03-01 scripts/sensitivity/proportion_redistributed_perfect_clean.R
perfect clean

@@ -1,0 +1,160 @@
+# The initial sensitivity analysis suggested that isolating no patients was better
+# than isolating a few patients (in terms of reducing the number of cases). This
+# is counterintuitive but probably because if no patients are isolated then it
+# avoids contaminating side room toilets.
+
+# To test this, this runs the same sensitivity analysis but with the toilet cleaning
+# effect 100% effective.
+
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-proportion-redistributed"]
+
+# fix to perfect clean
+estimated_consts_mean["c-toilet-cleaning-effect"] <- 1
+estimated_consts_mean["toilet-cleaning-effect"] <- 1
+estimated_consts_mean["o-toilet-cleaning-effect"] <- 1
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
[…]

### 2025-03-02 scripts/calibrate.R
messages on scripts, changed params

@@ -15,7 +15,7 @@ model_path <- here::here("models", "hospital_transmission.nlogo")
 out_path <- here::here("out")
 data_path <- here::here("data")
 
-# run_sims <- TRUE
+run_sims <- FALSE
 
 antibiotic_prescription_rate = 0.499 # Vesporten 2018
 admission_days <- 4.6
@@ -23,48 +23,70 @@ bay_proportion <- 0.6
 
 ##============== pre-outbreak =============##
 
-calibration_samples <- 100
+calibration_samples <- availableCores() * 8
 calibration_seeds <- 3
 abc_tol <- 0.05
+
+message("Running pre-outbreak calibration")
+message(glue::glue("Calibration samples: {calibration_samples},
+                   Calibration seeds: {calibration_seeds},
+                   ABC tolerance: {abc_tol}"))
 source(here::here("scripts", "tune", "pre_outbreak.R"))
 
 ##============== outbreak and control ===============##
 
-calibration_samples <- 1000
-calibration_seeds <- 3
-abc_tol <- 0.025
-source(here::here("scripts", "tune", "outbreak_and_control.R"))
+# calibration_samples <- 1000
+# calibration_seeds <- 3
+# abc_tol <- 0.025
+# message("Running outbreak and control calibration")
+# message(glue::glue("Calibration samples: {calibration_samples},
+#                    Calibration seeds: {calibration_seeds},
+#                    ABC tolerance: {abc_tol}"))
+# source(here::here("scripts", "tune", "outbreak_and_control.R"))
 
 ##=========== fine tune pre-outbreak ===================================##
 
-calibration_samples <- 1000
-calibration_seeds <- 3
-abc_tol <- 0.025
-source(here::here("scripts", "tune", "fine_tune_pre_outbreak.R"))
+# calibration_samples <- 1000
+# calibration_seeds <- 3
+# abc_tol <- 0.025
+# message("Running fine tune pre-outbreak calibration")
+# message(glue::glue("Calibration samples: {calibration_samples},
+#                    Calibration seeds: {calibration_seeds},
+#                    ABC tolerance: {abc_tol}"))
+# source(here::here("scripts", "tune", "fine_tune_pre_outbreak.R"))
 
 ##=========== use variables for pre-outbreak params as well ============##
 
-calibration_samples <- 1000
+calibration_samples <- availableCores() * 10
 calibration_seeds <- 3
 abc_tol <- 0.025
[…]

### 2025-03-02 scripts/sensitivity/proportion_redistributed_perfect_clean.R

@@ -56,8 +56,8 @@ nl@experiment <- experiment(expname = "sens_analysis",
 )
 
 nl@simdesign <- simdesign_lhs(nl,
-                              samples = 20,
-                              nseeds = 10,
+                              samples = sens_samples,
+                              nseeds = sens_seeds,
                               precision = 3)
 
 plan(list(sequential, multisession))

### 2025-03-02 scripts/sensitivity/toilet_cleaning_rate.R

@@ -43,8 +43,8 @@ nl@experiment <- experiment(expname = "sens_analysis",
 )
 
 nl@simdesign <- simdesign_lhs(nl,
-                              samples = 20,
-                              nseeds = 10,
+                              samples = sens_samples,
+                              nseeds = sens_seeds,
                               precision = 3)
 
 plan(list(sequential, multisession))

### 2025-03-02 scripts/tune/outbreak_and_control.R

@@ -31,7 +31,7 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                                                    max=outbreak_start + 90,
                                                    qfun='qunif'),
                            "o-toilet-frequenting-rate" = list(min=consts[["toilet-frequenting-rate"]],
-                                                              max=8,
+                                                              max=4,
                                                               qfun='qunif'),
                            "o-toilet-contamination-effect" = list(min=consts[["toilet-contamination-effect"]],
                                                                   max=0.99,

### 2025-03-02 scripts/tune/simulate_using_consts.R

@@ -48,8 +48,8 @@ nl@experiment <- experiment(expname = "consts_sim",
                             variables = estimated_consts)
 
 nl@simdesign <- simdesign_lhs(nl,
-                              samples = 20,
-                              nseeds = 10,
+                              samples = calibration_samples,
+                              nseeds = calibration_seeds,
                               precision = 3)
 
 plan(list(sequential, multisession))

### 2025-03-02 scripts/tune/variable_pre_outbreak.R

@@ -54,7 +54,7 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                                                 max=control_end + 30,
                                                 qfun='qunif'),
                            "o-toilet-frequenting-rate" = list(min=pre_outbreak_variables[["toilet-frequenting-rate"]]$min,
-                                                              max=8,
+                                                              max=4,
                                                               qfun='qunif'),
                            "o-toilet-contamination-effect" = list(min=pre_outbreak_variables[["toilet-contamination-effect"]]$min,
                                                                   max=0.99,

### 2025-03-03 scripts/sensitivity/abx_prescription_rates.R
wider param sweep

@@ -36,8 +36,8 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-antibiotic-prescription-rate" = list(min=0.1,
-                                                                                     max=0.7,
+                            variables = list("c-antibiotic-prescription-rate" = list(min=0.01,
+                                                                                     max=0.99,
                                                                                      qfun='qunif')
                             )
 )

### 2025-03-03 scripts/sensitivity/proportion_redistributed_perfect_clean.R
plot added to sens

@@ -158,3 +158,16 @@ list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
   ggplot(aes(x = value, y = mean_cases)) + 
   geom_line() + 
   facet_wrap(~ param)
+
+sens_prop_redist_perfect_clean_outbreak_cases %>% pivot_longer(cols = `c-proportion-redistributed`,
+                                                               names_to = "param",
+                                                               values_drop_na = TRUE) %>% 
+  group_by(param, value) %>% 
+  summarise(mean_cases = mean(`outbreak-cases`),
+            sd_cases = sd(`outbreak-cases`),
+            .groups = "keep") %>% 
+  ggplot(aes(x = value, y = mean_cases)) + 
+  geom_line() + 
+  # shade sd
+  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
+  facet_wrap(~ param, scales = "free")

### 2025-03-03 paper/paper.qmd
paper update

@@
 
-```{r fig-simulated-rates, include=TRUE}
+```{r fig-abc-rates-retained-posterios, include=TRUE}
 #| echo: false
-#| label: fig-simulated-rates
+#| label: fig-abc-rates-retained-posteriors
 #| fig.cap: "Observed and simulated rates of _C.
@@
 
-```{r, include=FALSE}
+```{r fig-consts-sims-coloured, include=FALSE}
 #| echo: false
@@
 
-```{r, include=FALSE}
+```{r fig-consts-sims-mean-sd, include=TRUE}
+# plot but with mean and sd instead rather than individual
+consts_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim) %>%
+  summarise(m = mean(rate),
+            q_high = quantile(rate, 0.75),
+            q_low = quantile(rate, 0.25),
+            .groups = "keep") %>%
+  ggplot() +
+  geom_line(aes(x = date_sim, y = m)) +
+  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
+  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none")
+```
+
+```{r fig-single-outbreak, include=FALSE}
 single_outbreak_sim_mod <- single_outbreak_sim %>% 
@@
 The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateaeu beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits.
-The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate.
+The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate during the outbreak period.
+The cohorting of patients into single-occupancy rooms after colonisation/infection had an almost linear relationship with the number of cases.
 
-```{r fig-sensitivity-analysis, include=TRUE}
+```{r fig-sensitivity-analysis, include=TRUE, eval=TRUE}
 #| echo: false
@@
 
-```{r}
-sens_prop_redist_perfect_clean_outbreak_cases %>% pivot_longer(cols = `c-proportion-redistributed`,
-               names_to = "param",
-               values_drop_na = TRUE) %>% 
-  group_by(param, value) %>% 
+```{r fig-sensitvity-pre-outbreak-abx, include=TRUE, eval=TRUE}
+sens_abx_prescription_pre_outbreak_outbreak_cases %>% 
+  group_by(`antibiotic-prescription-rate`) %>% 
   summarise(mean_cases = mean(`outbreak-cases`),
-            sd_cases = sd(`outbreak-cases`),
-            .groups = "keep") %>% 
-  ggplot(aes(x = value, y = mean_cases)) + 
+            sd_cases = sd(`outbreak-cases`)) %>% 
+  ggplot(aes(x = `antibiotic-prescription-rate`, y = mean_cases)) +
   geom_line() + 
-  # shade sd
-  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
-  facet_wrap(~ param, scales = "free")
+  # add sd
+  geom_line(data = . %>% 
+              mutate(upper = mean_cases + sd_cases,
+                     lower = mean_cases - sd_cases),
+            aes(y = upper), linetype = "dashed") +
+  geom_line(data = . %>% 
+              mutate(upper = mean_cases + sd_cases,
+                     lower = mean_cases - sd_cases),
+            aes(y = lower), linetype = "dashed")
 ```
@@
 output_df <- filter(output_df, time > 31)
+output_long <- reshape2::melt(output_df, id = "time")
 ```
 
-```{r fig-single-outbreak, include=TRUE}
+```{r fig-single-outbreak-deterministic, include=TRUE}
 #| echo: false
-output_long <- reshape2::melt(output_df, id = "time")
-
 single_outbreak_one_seed %>% 
@@
 
+Infection prevention and control (IPC) is the practice of preventing the spread of infectious diseases in healthcare settings.
+A central responsibility is the operational management of an outbreak, defined as an increase in the number of infected cases above the expected rate.
+This research describes the design and calibration of an agent-based model that simulates the spread of infection in a hospital setting.
+The primary motivation of this work is to assess the impact of different infection prevention interventions (such as change in practice) on rates of infection.
+The model was designed to simulate a paradigm of infections that are faecal-oral-transmitted, with the risk increasing in the presence of the selective pressure introduced by antimicrobial therapy.
+Since some of the pathogens that fit this definition may be associated with long-term colonisation of the gastro-intestinal tract (without the presence of clinical illness), infection and colonisation are used inter-changeably within this research.
+The model can be used to simulate any infection with these characteristics, such as: _Clostridioides difficile_, vancomycin-resistant _Enterococcus_, extended-spectrum beta-lactamase-producing Enterobacterales, and carbapenem-resistant _Enterobacterales_.
+
+The model was designed to simulate the spread of infection/colonisation at a baseline (or background rate), followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate.
+The model was calibrated to a report of a _C.
+difficile_ outbreak reported by @salgadoAnalysisOutbreakClostridium2009.
+Approximate Bayesian computation was used to calibrate the model to the reported rates of infection.
+When model inputs were drawn from the posterior distributions, the model was able to simulate plausible outbreak scenarios that were similar to the observed rates.
+A sensitivity analysis was performed over parameters that represent infection control interventions (such as the frequency of toilet cleaning), demonstrating the possible applications of this model for informing operational infection control practices.
+Interventions that controlled the spread of infection, such as the cohorting of patients to single-occupancy beds, and the increased frequency of toilet cleaning, led to a reduced number of cases during the outbreak period.
+
 Repeated simulations using the model generated nosocomial CDI rates that showed a clear outbreak peak, followed by control of colonisation/infection through a change in infection control practice, resembling a real hospital CDI outbreak.

### 2025-03-03 scripts/calibrate.R
pre outbreak sens

@@ -120,3 +120,6 @@ source(here::here("scripts", "sensitivity", "toilet_cleaning_effect.R"))
 #   geom_line() + 
 #   facet_wrap(~ param, scales = "free")
 
+####================= sensitivity analysis abx prescription rates preoutbreak ==============####
+
+source(here::here("scripts", "sensitivity", "abx_prescription_rates_pre_outbreak.R"))

### 2025-03-03 scripts/sensitivity/abx_prescription_rates_pre_outbreak.R

@@ -1,0 +1,132 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               # "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+
+estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+  apply(., 2, mean)
+
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "antibiotic-prescription-rate"]
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "o-antibiotic-prescription-rate"]
+
+nl@experiment <- experiment(expname = "sens_analysis",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = c(consts, estimated_consts_mean), 
+                            variables = list("antibiotic-prescription-rate" = list(values = seq(from=0.01, to=0.99, length.out = sens_samples)),
+                                             "o-antibiotic-prescription-rate" = list(values = seq(from=0.01, to=0.99, length.out = sens_samples))
+                            )
+)
+
+nl@simdesign <- simdesign_distinct(nl,
[…]

### 2025-03-03 scripts/tune/variable_pre_outbreak.R
tighter priors

@@ -54,10 +54,10 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                                                 max=control_end + 30,
                                                 qfun='qunif'),
                            "o-toilet-frequenting-rate" = list(min=pre_outbreak_variables[["toilet-frequenting-rate"]]$min,
-                                                              max=4,
+                                                              max=pre_outbreak_variables[["toilet-frequenting-rate"]]$max + 0.5 * pre_outbreak_variables[["toilet-frequenting-rate"]]$max,
                                                               qfun='qunif'),
                            "o-toilet-contamination-effect" = list(min=pre_outbreak_variables[["toilet-contamination-effect"]]$min,
-                                                                  max=0.99,
+                                                                  max=min(0.99, pre_outbreak_variables[["toilet-contamination-effect"]]$max + 0.5 * pre_outbreak_variables[["toilet-contamination-effect"]]$max),
                                                                   qfun='qunif'),
                            "o-toilet-cleaning-effect" = list(min=0.01,
                                                              max=pre_outbreak_variables[["toilet-cleaning-effect"]]$max,

### 2025-03-03 scripts/tune/variable_pre_outbreak.R
using prior means rather than max bounds for 50pcent outbreak change

@@ -54,22 +54,22 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                                                 max=control_end + 30,
                                                 qfun='qunif'),
                            "o-toilet-frequenting-rate" = list(min=pre_outbreak_variables[["toilet-frequenting-rate"]]$min,
-                                                              max=pre_outbreak_variables[["toilet-frequenting-rate"]]$max + 0.5 * pre_outbreak_variables[["toilet-frequenting-rate"]]$max,
+                                                              max=pre_outbreak_priors_means[["toilet-frequenting-rate"]] + 0.5 * pre_outbreak_priors_means[["toilet-frequenting-rate"]],
                                                               qfun='qunif'),
                            "o-toilet-contamination-effect" = list(min=pre_outbreak_variables[["toilet-contamination-effect"]]$min,
-                                                                  max=min(0.99, pre_outbreak_variables[["toilet-contamination-effect"]]$max + 0.5 * pre_outbreak_variables[["toilet-contamination-effect"]]$max),
+                                                                  max=min(0.99, pre_outbreak_priors_means[["toilet-contamination-effect"]] + 0.5 * pre_outbreak_priors_means[["toilet-contamination-effect"]]),
                                                                   qfun='qunif'),
-                           "o-toilet-cleaning-effect" = list(min=0.01,
+                           "o-toilet-cleaning-effect" = list(min=max(0.01, pre_outbreak_priors_means[["toilet-cleaning-effect"]] - 0.5 * pre_outbreak_priors_means[["toilet-cleaning-effect"]]),
                                                              max=pre_outbreak_variables[["toilet-cleaning-effect"]]$max,
                                                              qfun='qunif'),
                            "o-toilet-cleaning-rate" = list(min=0.5,
                                                            max=pre_outbreak_variables[["toilet-cleaning-rate"]]$max,
                                                            qfun='qunif'),
                            "o-community-colonisation-rate" = list(min=pre_outbreak_variables[["community-colonisation-rate"]]$min,
-                                                                  max=pre_outbreak_variables[["community-colonisation-rate"]]$max + 0.5 * pre_outbreak_variables[["community-colonisation-rate"]]$max,
+                                                                  max=pre_outbreak_priors_means[["community-colonisation-rate"]] + 0.5 * pre_outbreak_priors_means[["community-colonisation-rate"]],
                                                                   qfun='qunif'),
                            "o-antibiotic-prescription-rate" = list(min=antibiotic_prescription_rate,
[…]

### 2025-03-04 scripts/tune/simulate_using_consts.R
added mean to sim consts rates

@@ -110,8 +110,16 @@ salgado %>%
 # simulated data
 pre_outbreak_sims_rates <- consts_sim_rates %>% 
   dplyr::filter(date_sim < outbreak_start_date) %>% 
-  summarise(across(rate, list(median = median, q_low = \(x) quantile(x, 0.25), q_high = \(x) quantile(x, 0.75))))
+  summarise(across(rate, list(median = median,
+                              q_low = \(x) quantile(x, 0.25),
+                              q_high = \(x) quantile(x, 0.75),
+                              mean = mean)))
 
 outbreak_sims_rates <- consts_sim_rates %>% 
   dplyr::filter(date_sim >= outbreak_start_date & date_sim <= outbreak_end_date) %>% 
-  summarise(across(rate, list(median = median, q_low = \(x) quantile(x, 0.25), q_high = \(x) quantile(x, 0.75), min = min, max = max)))
+  summarise(across(rate, list(median = median,
+                              q_low = \(x) quantile(x, 0.25),
+                              q_high = \(x) quantile(x, 0.75),
+                              min = min,
+                              max = max,
+                              mean = mean)))

### 2025-03-04 paper/paper.qmd
mean instead of median

@@
 Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-simulated-rates).
-In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
-During the outbreak period, simulations had a median rate of `r round(outbreak_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_sims_rates$rate_mean, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+During the outbreak period, simulations had a mean rate of `r round(outbreak_sims_rates$rate_mean, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real mean rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
 

### 2025-03-11 scripts/calibrate.R
variable scaling

@@ -60,6 +60,13 @@ source(here::here("scripts", "tune", "pre_outbreak.R"))
 calibration_samples <- availableCores() * 10
 calibration_seeds <- 3
 abc_tol <- 0.025
+
+upper_bound <- "mean" # mean, min, or max (of pre-outbreak priors)
+upper_bound_scale <- 1 # scale the upper bound by this factor
+
+lower_bound <- "mean" # mean, min, or max (of pre-outbreak priors)
+lower_bound_scale <- 1 # scale the lower bound by this factor
+
 message("Running pre-outbreak calibration with variables")
 message(glue::glue("Calibration samples: {calibration_samples},
                    Calibration seeds: {calibration_seeds},

### 2025-03-11 scripts/tune/variable_pre_outbreak.R

@@ -41,6 +41,8 @@ pre_outbreak_priors_upper_sd <- pre_outbreak_priors_means + pre_outbreak_priors_
 pre_outbreak_variables <- apply(pre_outbreak_priors, 2, \(x) {
   list(min=unname(quantile(x, 0.25)),
        max=unname(quantile(x, 0.75)),
+       mean=mean(x),
+       sd=sd(x),
        qfun='qunif')
 })
 
@@ -54,22 +56,22 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                                                 max=control_end + 30,
                                                 qfun='qunif'),
                            "o-toilet-frequenting-rate" = list(min=pre_outbreak_variables[["toilet-frequenting-rate"]]$min,
-                                                              max=pre_outbreak_priors_means[["toilet-frequenting-rate"]] + 0.5 * pre_outbreak_priors_means[["toilet-frequenting-rate"]],
+                                                              max=pre_outbreak_variables[["toilet-frequenting-rate"]][[upper_bound]] + upper_bound_scale * pre_outbreak_variables[["toilet-frequenting-rate"]][[upper_bound]],
                                                               qfun='qunif'),
                            "o-toilet-contamination-effect" = list(min=pre_outbreak_variables[["toilet-contamination-effect"]]$min,
-                                                                  max=min(0.99, pre_outbreak_priors_means[["toilet-contamination-effect"]] + 0.5 * pre_outbreak_priors_means[["toilet-contamination-effect"]]),
+                                                                  max=min(0.99, pre_outbreak_variables[["toilet-contamination-effect"]][[upper_bound]] + upper_bound_scale * pre_outbreak_variables[["toilet-contamination-effect"]][[upper_bound]]),
                                                                   qfun='qunif'),
-                           "o-toilet-cleaning-effect" = list(min=max(0.01, pre_outbreak_priors_means[["toilet-cleaning-effect"]] - 0.5 * pre_outbreak_priors_means[["toilet-cleaning-effect"]]),
+                           "o-toilet-cleaning-effect" = list(min=max(0.01, pre_outbreak_variables[["toilet-cleaning-effect"]][[lower_bound]] - lower_bound_scale * pre_outbreak_variables[["toilet-cleaning-effect"]][[upper_bound]]),
                                                              max=pre_outbreak_variables[["toilet-cleaning-effect"]]$max,
                                                              qfun='qunif'),
                            "o-toilet-cleaning-rate" = list(min=0.5,
[…]

### 2025-03-12 paper/paper.qmd
discussion

@@
 
-The model was designed to simulate the spread of infection/colonisation at a baseline (or background rate), followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate.
+The model was designed to simulate the spread of infection/colonisation at a baseline (or background) rate, followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate.
 The model was calibrated to a report of a _C.
 difficile_ outbreak reported by @salgadoAnalysisOutbreakClostridium2009.
-Approximate Bayesian computation was used to calibrate the model to the reported rates of infection.
+Approximate Bayesian computation was used to calibrate the model to the rates of infection reported in the paper.
 When model inputs were drawn from the posterior distributions, the model was able to simulate plausible outbreak scenarios that were similar to the observed rates.
@@
 
-Repeated simulations using the model generated nosocomial CDI rates that showed a clear outbreak peak, followed by control of colonisation/infection through a change in infection control practice, resembling a real hospital CDI outbreak.
-The interactive nature of the model can be used for outbreak preparedness by testing out different infection control strategies.
-In summary, we report a pathogen-agnostic model that can be used to simulate any infection with features such as faecal-oral transmission, colonisation, and antimicrobial selective pressure effects.
+To simulate outbreaks and periods of enhanced infection control, the model was designed to change parameters at discrete time points.
+For example, the rate of antimicrobial prescriptions was higher during the outbreak period, promoting the spread of infection and colonisation.
+Later, the prescription rate was reduced to help control the outbreak.
+The NetLogo model implementation was designed so that a fixed parameter set was used as an input, and the outbreak and enhanced IPC parameters were only used during their respective periods.
+In effect, all input parameters were provided as a fixed input -- some parameters were only used by the simulation at defined discrete intervals.
+This approach may be unsuitable for modelling of transmission without a clear outbreak pattern -- more complex implementations of time-varying parameters could be explored.
+
+Although agent-based models have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations.
+The model described here is a simplification of what is likely to be a highly intricate and complicated system.
+For example, transmission was assumed to be 'random' or acquired through the use of a contaminated toilet.
+In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding.
+Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected or colonised.
+Antimicrobial class and duration of exposure are factors that are not accounted for in this model.
+An agent-based model that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated.
+A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016].
+Another limitation of agent-based models is the requirement for detailed data to calibrate and validate the model.
+Here, the data used was primarily from a retrospective descriptive dataset.
+As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature.
+An important future direction would be to prospectively inform and validate the model with data collected specifically for this purpose.
+
+Agent-based models that simulate the hospital transmission of pathogens transmitted through the faecal-oral route, including _C.
+difficile_ and vancomycin-resistant _Enterococci_, have been reported previously [@codellaAgentbasedSimulationModel2015; @debosckerAgentbasedModelSimulate2021].
+Some models have also explored the impact of one or more infection control interventions on the spread of infection [@barkerInterventionsReduceIncidence2018; @rubinSimulationBasedAssessmentStrategies2013].
+Our model builds on these works by designing and calibrating the model to an outbreak and enhanced IPC period.
+The motivation for this was to design a model that can be used to support infection and prevention teams in the following: 1) preparedness for outbreak situations; and 2) decision-making on the implementation of infection control interventions during an outbreak.
+Furthermore, the model was designed to be pathogen-agnostic to allow for a stepwise approach for future work -- pathogen-specific features (e.g., diagnostic and laboratory testing) can be added as required.
+The choice of NetLogo as a platform allows for the model to be used interactively by non-programmers, such as infection control practitioners; the model can also be modified and extended with the relatively simple NetLogo language.
+
+In conclusion, the agent-based model described here can simulate outbreaks of infection in a hospital setting, and estimate the impact of different infection control interventions on the number of nosocomial infections.
+Although the model was calibrated to a _C.
+difficile_ outbreak, it can be adapted for use in other pathogens that are transmitted through the faecal-oral route under antimicrobial selective pressure, such as vancomycin-resistant _Enterococci_ and extended-spectrum beta-lactamase-producing Enterobacterales.

### 2025-03-12 scripts/calibrate.R
creates out/ if necessary

@@ -23,6 +23,10 @@ bay_proportion <- 0.6
 
 ##============== pre-outbreak =============##
 
+if (!dir.exists(out_path)) {
+  dir.create(out_path)
+}
+
 calibration_samples <- availableCores() * 8
 calibration_seeds <- 3
 abc_tol <- 0.05

### 2025-03-12 paper/paper.qmd
intro part

@@
 Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection.
-Examples include *Clostridioides difficile* infection (CDI) and Vancomycin-resistant *Enterococcus*.
-In addition to patient factors, colonisation risk depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents.
+Examples of such pathogens include *Clostridioides difficile* infection (CDI), vancomycin-resistant *Enterococcus* (VRE), extended-spectrum beta-lactamase-producing (ESBL) Enterobacterales, and carbapenem-resistant Enterobacterales (CPE).
+Infection with these pathogens can lead to increased morbidity and mortality -- for example, CPE bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
+_C. difficile_ infection is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
+Furthermore, the therapeutic antimicrobial options available for VRE, ESBL, and CPE are limited, leading to increased use of WHO reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].
+
+Infection prevention and control (IPC) strategies are essential to prevent the spread of these pathogens within a healthcare setting, and include measures such as hand hygiene, environmental cleaning, isolation of infected/colonised patients, and more [@worldhealthorganizationGlobalReportInfection2022].
+In many healthcare settings, IPC strategies and operational practices are led by a multi-disciplinary IPC team.
+A key role of IPC teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
+In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
+These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
+The impact of these interventions is often difficult to quantify.
+Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- the relative importance of each intervention is rarely quantified.
+
+In addition to patient factors, the risk of infection and/or colonisation depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents.
+
 To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by @salgadoAnalysisOutbreakClostridium2009.

### 2025-03-12 scripts/tune/fine_tune_pre_outbreak.R
sumstat now ordered to match target

@@ -66,7 +66,7 @@ results_baseline_fine_tune <- results_baseline_fine_tune %>%
                         length.out = n()))
 
 results_baseline_rates <- results_baseline_fine_tune %>% 
-  mutate(year = year(date_sim), month = month(date_sim)) %>% 
+  mutate(year = year(date_sim), month = str_pad(month(date_sim), 2, pad = "0")) %>% 
   group_by(`random-seed`, siminputrow, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
@@ -100,7 +100,8 @@ pre_outbreak_sims <- results_baseline_summary %>%
 
 sumstats <- pre_outbreak_sims %>% 
   ungroup %>% 
-  dplyr::select(starts_with("rate_"))
+  dplyr::select(starts_with("rate_")) %>% 
+  dplyr::select(sort(names(.)))
 
 abc_params <- abc(target = pre_outbreak_data$rates,
                   param = pre_outbreak_sims[params_names_pre_outbreak],

### 2025-03-12 scripts/tune/outbreak_and_control.R

@@ -109,7 +109,7 @@ results_outbreak_control <- results_outbreak_control %>%
                         length.out = n()))
 
 results_outbreak_control_rates <- results_outbreak_control %>% 
-  mutate(year = year(date_sim), month = month(date_sim)) %>% 
+  mutate(year = year(date_sim), month = str_pad(month(date_sim), 2, pad = "0")) %>% 
   group_by(siminputrow, `random-seed`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
@@ -143,7 +143,8 @@ outbreak_control_sims <- results_outbreak_control_summary %>%
 
 sumstats <- outbreak_control_sims %>% 
   ungroup %>% 
-  dplyr::select(starts_with("rate_"))
+  dplyr::select(starts_with("rate_")) %>% 
+  dplyr::select(sort(names(.)))
 
 abc_params_outbreak_control <-abc(target = salgado$rates,
                                   param = outbreak_control_sims[params_names_outbreak_control],

### 2025-03-12 scripts/tune/pre_outbreak.R

@@ -63,7 +63,7 @@ results_baseline <- results_baseline %>%
                         length.out = n()))
 
 results_baseline_rates <- results_baseline %>% 
-  mutate(year = year(date_sim), month = month(date_sim)) %>% 
+  mutate(year = year(date_sim), month = str_pad(month(date_sim), 2, pad = "0")) %>% 
   group_by(`random-seed`, siminputrow, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
@@ -97,7 +97,8 @@ pre_outbreak_sims <- results_baseline_summary %>%
 
 sumstats <- pre_outbreak_sims %>% 
   ungroup %>% 
-  dplyr::select(starts_with("rate_"))
+  dplyr::select(starts_with("rate_")) %>% 
+  dplyr::select(sort(names(.)))
 
 abc_params <- abc(target = pre_outbreak_data$rates,
                   param = pre_outbreak_sims[params_names_pre_outbreak],

### 2025-03-12 scripts/tune/variable_pre_outbreak.R

@@ -135,7 +135,7 @@ results_outbreak_control_pre_included <- results_outbreak_control_pre_included %
                         length.out = n()))
 
 results_outbreak_control_pre_included_rates <- results_outbreak_control_pre_included %>% 
-  mutate(year = year(date_sim), month = month(date_sim)) %>% 
+  mutate(year = year(date_sim), month = str_pad(month(date_sim), 2, pad = "0")) %>% 
   group_by(siminputrow, `random-seed`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
@@ -169,8 +169,9 @@ outbreak_control_sims <- results_outbreak_control_pre_included_summary %>%
 
 sumstats <- outbreak_control_sims %>% 
   ungroup %>% 
-  dplyr::select(starts_with("rate_"))
-
+  dplyr::select(starts_with("rate_")) %>% 
+  dplyr::select(sort(names(.)))
+  
 abc_params_outbreak_control <-abc(target = salgado$rates,
                                   param = outbreak_control_sims[params_names_outbreak_control],
                                   sumstat = sumstats,

### 2025-03-12 scripts/tune/simulate_using_consts.R
blank line

@@ -80,7 +80,6 @@ consts_sim_rates <- consts_sim %>%
   ungroup() %>% 
   dplyr::select(!c(year, month))
 
-
 consts_sim_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 
   #slice_sample(n=10) %>% 

### 2025-03-12 scripts/calibrate.R
progressr

@@ -6,6 +6,8 @@ library(abc)
 library(progressr)
 library(here)
 handlers("cli")
+options(progressr.enable=TRUE)
+
 wards_total <- 9
 bedspaces_per_ward <- 9
 set.seed(42)

### 2025-03-12 scripts/tune/variable_pre_outbreak.R
avoid simulating from non-outbreak parameters that will be rejected anyways

@@ -110,9 +110,16 @@ nl@experiment <- experiment(expname = "outbreak_control",
                             constants = consts)
 
 nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples,
+                              samples = calibration_samples * 4,
                               nseeds = calibration_seeds,
                               precision = 3)
+
+# keep simulated parameters that fit with outbreak generation
+nl@simdesign@siminput <- nl@simdesign@siminput %>%
+  dplyr::filter(`o-toilet-contamination-effect` >= `toilet-contamination-effect` &
+                  `o-toilet-frequenting-rate` >= `toilet-frequenting-rate` &
+                  `o-community-colonisation-rate` >= `community-colonisation-rate`) %>% 
+  sample_n(calibration_samples)
 
 plan(list(sequential, multisession))
 
@@ -139,17 +146,13 @@ results_outbreak_control_pre_included_rates <- results_outbreak_control_pre_incl
   group_by(siminputrow, `random-seed`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
-results_outbreak_control_pre_included_summary <- results_outbreak_control_pre_included %>% distinct(`random-seed`, siminputrow, .keep_all = TRUE) %>% 
+results_outbreak_control_pre_included_summary <- results_outbreak_control_pre_included %>%
+  dplyr::select(!`random-seed`) %>% 
+  distinct(siminputrow, .keep_all = TRUE) %>% 
   # mutate(siminputrow = siminputrow) %>% 
-  right_join(results_outbreak_control_pre_included_rates, by=c("siminputrow", "random-seed"))
+  right_join(results_outbreak_control_pre_included_rates, by="siminputrow")
 
 params_names_outbreak_control <- names(nl@experiment@variables)
-
-# keep simulated parameters that fit with outbreak generation
-results_outbreak_control_pre_included_summary <- results_outbreak_control_pre_included_summary %>% 
-  dplyr::filter(`o-toilet-contamination-effect` >= `toilet-contamination-effect` &
-                  `o-toilet-frequenting-rate` >= `toilet-frequenting-rate` &
-                  `o-community-colonisation-rate` >= `community-colonisation-rate`)
 
 outbreak_control_sims <- results_outbreak_control_pre_included_summary %>% 
   dplyr::select(all_of(params_names_outbreak_control), `random-seed`, siminputrow, month, year, rate) %>% 

### 2025-03-12 scripts/tune/variable_pre_outbreak.R
patch previous commit

@@ -147,10 +147,10 @@ results_outbreak_control_pre_included_rates <- results_outbreak_control_pre_incl
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000)
 
 results_outbreak_control_pre_included_summary <- results_outbreak_control_pre_included %>%
-  dplyr::select(!`random-seed`) %>% 
-  distinct(siminputrow, .keep_all = TRUE) %>% 
-  # mutate(siminputrow = siminputrow) %>% 
-  right_join(results_outbreak_control_pre_included_rates, by="siminputrow")
+  # dplyr::select(!`random-seed`) %>% 
+  distinct(`random-seed`, siminputrow, .keep_all = TRUE) %>% 
+  # mutate(siminputrow = siminputrow) %>%
+  right_join(results_outbreak_control_pre_included_rates, by=c("siminputrow", "random-seed"))
 
 params_names_outbreak_control <- names(nl@experiment@variables)
 

### 2025-03-12 scripts/calibrate.R
post pred check in calibrate

@@ -96,6 +96,13 @@ source(here::here("scripts", "tune", "single_outbreak.R"))
 
 # save.image(file.path(out_path, "calibrate.RData"))
 
+##================= posterior predictive checks ==================##
+
+calibration_samples <- availableCores() * 1
+calibration_seeds <- 3
+
+source(here::here("scripts", "tune", "posterior_predictive_checks.R"))
+
 ##================= sensitivity analysis ==================##
 
 sens_samples <- availableCores() * 1

### 2025-03-12 scripts/tune/posterior_predictive_checks.R
posterior predictive checks

@@ -1,0 +1,128 @@
+sim_days <- (365 * 7) + 30
+
+# outbreak_control_model <- "salgado.nlogo"
+
+nl <- nl(nlversion = "6.2.2",
+         nlpath = netlogo_path,
+         modelpath = model_path,
+         jvmmem = 1024)
+
+consts <- list("wards-total" = wards_total,
+               "bedspaces-per-ward" = bedspaces_per_ward,
+               "antibiotic-prescription-rate" = antibiotic_prescription_rate, # Vesporten 2018
+               "admission-days" = admission_days,
+               "bay-proportion" = bay_proportion)
+
+# set outbreak and control
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
+
+# estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
+#   apply(., 2, mean)
+# 
+# estimated_consts_sd <- abc_params_outbreak_control$unadj.values %>%
+#   apply(., 2, sd)
+# 
+# estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
+#   list(min = x - y,
+#        max = x + y,
+#        qfun = 'qunif')
+# })
+
+# estimated_consts_fixed <- c(estimated_consts_mean, consts)
+
+posterior_samples <- abc_params_outbreak_control$unadj.values %>% 
+  as.data.frame() %>% 
+  sample_n(calibration_samples, replace = TRUE) %>% 
+  as.list() %>% 
+  map(\(x) list(values = x))
+
+nl@experiment <- experiment(expname = "posterior_predictive_sim",
+                            outpath = out_path,
+                            repetition = 1,
+                            tickmetrics = "true",
+                            idsetup = "setup",
+                            idgo = "go",
+                            runtime = sim_days,
+                            metrics = c("total-colonised",
+                                        "total-patients-admitted",
+                                        "total-hospital-infections",
+                                        "current-community-infections",
+                                        "current-hospital-infections",
+                                        "current-inpatients",
+                                        "current-colonised"),
+                            constants = consts,
+                            variables = posterior_samples)
+
+nl@simdesign <- simdesign_distinct(nl,
[…]

### 2025-03-12 paper/paper.qmd
intro update

@@
 The impact of these interventions is often difficult to quantify.
-Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- the relative importance of each intervention is rarely quantified.
+Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
+Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such rates of antimicrobial usage, and to structural differences such as hospital layout.
+
+In practice, the approach to managing outbreaks is often reactive.
+Once an outbreak is identified and declared, the IPC team implements a set of enhanced infection control measures to attempt to attempt to halt transmission.
+The number of cases of infection or colonisation is monitored over a time frame dependent on the pathogen, setting, and scale of the outbreak.
+If the measures are successful in controlling the spread, then they are gradually relaxed once rates return to baseline.
+If they are not, then further measures are implemented, or the existing measures are intensified, until control is achieved.
 

### 2025-03-12 paper/paper.qmd
post predictive check figures

@@
 
-```{r fig-consts-sims-coloured, include=FALSE}
+```{r fig-consts-sims-coloured, include=TRUE}
 #| echo: false
@@
 
+```{r fig-single-outbreak-coloured, include=TRUE}
+#| echo: false
+#| fig.width: 18
+#| fig.height: 8
+#| message: false
+single_outbreak_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  #slice_sample(n=10) %>% 
+  dplyr::select(!`siminputrow`) %>% 
+  pivot_longer(cols = !c(`random-seed`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim, `random-seed`) %>%
+  summarise(rate = mean(rate)) %>%
+  ggplot(aes(x = date_sim,
+             y = rate,
+             color = factor(`random-seed`))) +
+  geom_line() +
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
+  guides(color="none")
+```
+
 ```{r fig-single-outbreak, include=FALSE}
@@
 #   summarise(m = mean(prop_infected))
+```
+
+```{r fig-posterior-predictive-check, include=TRUE}
+#| echo: false
+posterior_predictive_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim) %>%
+  summarise(m = median(rate),
+            q_high = quantile(rate, 0.75),
+            q_low = quantile(rate, 0.25),
+            .groups = "keep") %>%
+  ggplot() +
+  geom_line(aes(x = date_sim, y = m)) +
+  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
+  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none")
+```
+
+```{r fig-posterior-predictive-check-coloured, include=TRUE}
+#| echo: false
+#| message: false
+posterior_predictive_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  #slice_sample(n=10) %>% 
+  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(siminputrow, date_sim) %>%
+  summarise(rate = mean(rate)) %>%
+  ggplot(aes(x = date_sim,
+             y = rate,
+             color = factor(`siminputrow`))) +
+  geom_line() +
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
+  guides(color="none")
 ```

### 2025-03-12 scripts/tune/posterior_predictive_checks.R
bug in ipc not on

@@ -14,7 +14,7 @@ consts <- list("wards-total" = wards_total,
                "bay-proportion" = bay_proportion)
 
 # set outbreak and control
-consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
 
 # estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
 #   apply(., 2, mean)
@@ -30,9 +30,11 @@ consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
 
 # estimated_consts_fixed <- c(estimated_consts_mean, consts)
 
-posterior_samples <- abc_params_outbreak_control$unadj.values %>% 
+posterior_samples_df <- abc_params_outbreak_control$unadj.values %>% 
   as.data.frame() %>% 
-  sample_n(calibration_samples, replace = TRUE) %>% 
+  sample_n(calibration_samples, replace = TRUE)
+
+posterior_samples <- posterior_samples_df %>%
   as.list() %>% 
   map(\(x) list(values = x))
 
@@ -126,3 +128,9 @@ outbreak_sims_rates <- posterior_predictive_sim_rates %>%
                               min = min,
                               max = max,
                               mean = mean)))
+
+# below saves one of the parameter sets
+# posterior_samples_df %>% 
+#   filter(row_number() == 3) %>% 
+#   t() %>% 
+#   write.table("out/singleparam.csv", col.names = FALSE, sep = ",")

### 2025-03-12 scripts/tune/single_outbreak.R
bug ipc not on

@@ -14,7 +14,7 @@ consts <- list("wards-total" = wards_total,
                "bay-proportion" = bay_proportion)
 
 # set outbreak and control
-consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = FALSE)
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
 
 estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
   apply(., 2, mean)

### 2025-03-14 scripts/calibrate.R
posterior predictive check properly uses seeds

@@ -99,8 +99,6 @@ source(here::here("scripts", "tune", "single_outbreak.R"))
 ##================= posterior predictive checks ==================##
 
 calibration_samples <- availableCores() * 1
-calibration_seeds <- 3
-
 source(here::here("scripts", "tune", "posterior_predictive_checks.R"))
 
 ##================= sensitivity analysis ==================##

### 2025-03-14 scripts/tune/posterior_predictive_checks.R

@@ -14,7 +14,8 @@ consts <- list("wards-total" = wards_total,
                "bay-proportion" = bay_proportion)
 
 # set outbreak and control
-consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
+consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE,
+            "force-seed?" = TRUE)
 
 # estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
 #   apply(., 2, mean)
@@ -38,6 +39,8 @@ posterior_samples <- posterior_samples_df %>%
   as.list() %>% 
   map(\(x) list(values = x))
 
+posterior_samples[["seed-value"]] <- list(values = nlrx:::util_generate_seeds(nrow(posterior_samples_df)))
+
 nl@experiment <- experiment(expname = "posterior_predictive_sim",
                             outpath = out_path,
                             repetition = 1,
@@ -56,7 +59,7 @@ nl@experiment <- experiment(expname = "posterior_predictive_sim",
                             variables = posterior_samples)
 
 nl@simdesign <- simdesign_distinct(nl,
-                              nseeds = calibration_seeds)
+                              nseeds = 1)
 
 plan(list(sequential, multisession))
 

### 2025-03-14 scripts/tune/single_outbreak.R
redundant code

@@ -19,15 +19,6 @@ consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
 estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
   apply(., 2, mean)
 
-estimated_consts_sd <- abc_params_outbreak_control$unadj.values %>%
-  apply(., 2, sd)
-
-estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
-  list(min = x - y,
-       max = x + y,
-       qfun = 'qunif')
-})
-
 estimated_consts_fixed <- c(estimated_consts_mean, consts)
 
 nl@experiment <- experiment(expname = "single_outbreak_sim",

### 2025-03-14 paper/paper.qmd
captions

--- 2025-03-12 paper/paper.qmd
@@
 
-```{r, include=FALSE}
+```{r text-vals, include=FALSE}
 posteriors <- abc_params_outbreak_control$unadj.values
@@
 ```{r tbl-posteriors, include=TRUE}
-#| tbl-cap: "Summary of parameter posteriors after calibration using Approximate bayesian computation."
+#| tbl-cap: "Summary of parameter posteriors after calibration using approximate Bayesian computation."
 #| echo: false
@@
 
-```{r fig-abc-rates-retained-posterios, include=TRUE}
+```{r fig-abc-rates-retained-posteriors, include=TRUE}
 #| echo: false
 #| label: fig-abc-rates-retained-posteriors
-#| fig.cap: "Observed and simulated rates of _C.
-difficile_ infection over the simulated period.
-Approximate Bayesian computation was used to determine the unobserved parameters of the agent-based model.
+#| fig.cap: "Simulated rates of _C.
+difficile_ infection generated from the simulations retained after applying the rejection approximate Bayesian computation algorithm.
 The red line shows the true observed infection rates.
-The black solid line shows the mean simulated rates; the dashed lines show the standard deviation."
+The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
 #| fig.width: 18
@@
 ```{r fig-consts-sims-mean-sd, include=TRUE}
+#| echo: false
+#| fig-cap: "Simulated rates of _C.
+difficile_ infection generated by using the posterior parameter distributions.
+Input parameter samples were generated by sampling the uniform distribution $U_[\\mu - \\sigma, \\mu + \\sigma]$, where $\\mu$ and $\\sigma$ are the mean and standard deviation of the parameter's posterior distribution respectively.
+The red line shows the true observed infection rates.
+The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
 # plot but with mean and sd instead rather than individual
@@
   group_by(date_sim) %>%
-  summarise(m = mean(rate),
+  summarise(m = median(rate),
             q_high = quantile(rate, 0.75),
@@
   geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none")
+```
+
+```{r fig-single-outbreak-coloured, include=FALSE}
+#| echo: false
+#| fig.width: 18
+#| fig.height: 8
+#| message: false
+single_outbreak_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  #slice_sample(n=10) %>% 
+  dplyr::select(!`siminputrow`) %>% 
+  pivot_longer(cols = !c(`random-seed`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim, `random-seed`) %>%
+  summarise(rate = mean(rate)) %>%
+  ggplot(aes(x = date_sim,
+             y = rate,
+             color = factor(`random-seed`))) +
+  geom_line() +
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
   guides(color="none")
@@
 
+```{r fig-posterior-predictive-check, include=TRUE}
+#| echo: false
+#| fig-cap: "Posterior predictive check of the model.
+Simulations were generated by sampling with replacement from the posterior parameter distributions after approximate Bayesian computation.
+Each simulation was run using different random number generators.
+The red line shows the true observed infection rates.
+The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
+posterior_predictive_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim) %>%
+  summarise(m = median(rate),
+            q_high = quantile(rate, 0.75),
+            q_low = quantile(rate, 0.25),
+            .groups = "keep") %>%
+  ggplot() +
+  geom_line(aes(x = date_sim, y = m)) +
+  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
+  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none")
+```
+
+```{r fig-posterior-predictive-check-coloured, include=FALSE}
+#| echo: false
+#| message: false
+posterior_predictive_sim_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  #slice_sample(n=10) %>% 
+  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(siminputrow, date_sim) %>%
+  summarise(rate = mean(rate)) %>%
+  ggplot(aes(x = date_sim,
+             y = rate,
+             color = factor(`siminputrow`))) +
+  geom_line() +
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "black") +
+  guides(color="none")
+```
+
 The sensitivity analysis explored the impact of different infection control practices on the number of cases observed during the outbreak period -- the results are shown in @fig-sensitivity-analysis.
-The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateaeu beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits.
+The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateau beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits.
 The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate during the outbreak period.
@@
 ```{r fig-sensitvity-pre-outbreak-abx, include=TRUE, eval=TRUE}
+#| echo: false
 sens_abx_prescription_pre_outbreak_outbreak_cases %>% 

### 2025-03-14 paper/paper.qmd
better names - uniform sampling script

@@
 If they are not, then further measures are implemented, or the existing measures are intensified, until control is achieved.
+A inherent limitation of this approach is that decisions are made based on observed data, and the impact of interventions cannot be quantified until after a change in the rate of infection has been observed, by which time further preventable nosocomial infections may have occurred.
 
-In addition to patient factors, the risk of infection and/or colonisation depends on environmental factors (e.g., surface cleaning), proximity to other colonised patients, and use of antimicrobial agents.
+Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
+Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community [@kerrCovasimAgentbasedModel2021].
 
@@
 
-The posterior parameter distributions were used to simulate `r length(unique(consts_sim$siminputrow))` outbreak scenarios. Parameter samples were drawn from a uniform distribution:
+The posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios. Parameter samples were drawn from a uniform distribution:
 
@@
 ```{r, include=FALSE}
-sens_n_seeds <- length(unique(consts_sim$siminputrow))
+sens_n_seeds <- length(unique(uniform_sims$siminputrow))
 ```
@@
 
-Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-simulated-rates).
-In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_sims_rates$rate_mean, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_uniform_sims_rates$rate_mean, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_uniform_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
 During the outbreak period, simulations had a mean rate of `r round(outbreak_sims_rates$rate_mean, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real mean rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
@@
 #| message: false
-consts_sim_rates %>% 
+uniform_sims_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 
@@
 # plot but with mean and sd instead rather than individual
-consts_sim_rates %>% 
+uniform_sims_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 

### 2025-03-14 scripts/calibrate.R

@@ -79,13 +79,13 @@ message(glue::glue("Calibration samples: {calibration_samples},
                    ABC tolerance: {abc_tol}"))
 source(here::here("scripts", "tune", "variable_pre_outbreak.R"))
 
-##================= simulate using consts only ==================##
+##=============== simulate by uniform sampling from posterior ===============##
 # This draws parameters from the posterior means +/- SD
 # These parameter sets are used as inputs to outbreak model
 
 calibration_samples <- availableCores() * 1
 calibration_seeds <- 10
-source(here::here("scripts", "tune", "simulate_using_consts.R"))
+source(here::here("scripts", "tune", "simulate_uniform.R"))
 
 
 ###========== single outbreak only simulation =========###

### 2025-03-14 scripts/tune/simulate_uniform.R

@@ -30,7 +30,7 @@ estimated_consts <- map2(estimated_consts_mean, estimated_consts_sd, \(x, y) {
 
 estimated_consts_fixed <- c(estimated_consts_mean, consts)
 
-nl@experiment <- experiment(expname = "consts_sim",
+nl@experiment <- experiment(expname = "uniform_sims",
                             outpath = out_path,
                             repetition = 1,
                             tickmetrics = "true",
@@ -55,24 +55,24 @@ nl@simdesign <- simdesign_lhs(nl,
 plan(list(sequential, multisession))
 
 if (run_sims) {
-  consts_sim <- progressr::with_progress(
+  uniform_sims <- progressr::with_progress(
     run_nl_all(nl))
-  write_rds(consts_sim, file.path(out_path, "consts_sim.rds"))
+  write_rds(uniform_sims, file.path(out_path, "uniform_sims.rds"))
 }
 
-consts_sim <- read_rds(file.path(out_path, "consts_sim.rds"))
+uniform_sims <- read_rds(file.path(out_path, "uniform_sims.rds"))
 
-consts_sim_bak <- consts_sim
-consts_sim <- consts_sim %>% 
+uniform_sims_bak <- uniform_sims
+uniform_sims <- uniform_sims %>% 
   filter(`[step]` > 31)
 
-consts_sim <- consts_sim %>% 
+uniform_sims <- uniform_sims %>% 
   group_by(`random-seed`, `siminputrow`) %>% 
   mutate(date_sim = seq(from = ymd("2002-01-01"),
                         by = "1 day",
                         length.out = n()))
 
-consts_sim_rates <- consts_sim %>% 
+uniform_sims_rates <- uniform_sims %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
   group_by(`random-seed`, `siminputrow`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000) %>% 
@@ -80,7 +80,7 @@ consts_sim_rates <- consts_sim %>%
   ungroup() %>% 
   dplyr::select(!c(year, month))
 
-consts_sim_rates %>% 
+uniform_sims_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 
   #slice_sample(n=10) %>% 
   pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
[…]

### 2025-03-14 paper/paper.qmd
variable renames

@@
 In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_uniform_sims_rates$rate_mean, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_uniform_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
-During the outbreak period, simulations had a mean rate of `r round(outbreak_sims_rates$rate_mean, 2)` \[IQR `r round(outbreak_sims_rates$rate_q_low, 2)`--`r round(outbreak_sims_rates$rate_q_high, 2)`\] cases, compared with the real mean rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
+During the outbreak period, simulations had a mean rate of `r round(outbreak_uniform_sims_rates$rate_mean, 2)` \[IQR `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real mean rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
 

### 2025-03-14 scripts/tune/posterior_predictive_checks.R

@@ -116,14 +116,14 @@ salgado %>%
   summarise(across(rates, list(median = median, q_low = \(x) quantile(x, 0.25), q_high = \(x) quantile(x, 0.75), min = min, max = max)))
 
 # simulated data
-pre_outbreak_sims_rates <- posterior_predictive_sim_rates %>% 
+pre_outbreak_posterior_predictive_sims_rates <- posterior_predictive_sim_rates %>% 
   dplyr::filter(date_sim < outbreak_start_date) %>% 
   summarise(across(rate, list(median = median,
                               q_low = \(x) quantile(x, 0.25),
                               q_high = \(x) quantile(x, 0.75),
                               mean = mean)))
 
-outbreak_sims_rates <- posterior_predictive_sim_rates %>% 
+outbreak_posterior_predictive_sims_rates <- posterior_predictive_sim_rates %>% 
   dplyr::filter(date_sim >= outbreak_start_date & date_sim <= outbreak_end_date) %>% 
   summarise(across(rate, list(median = median,
                               q_low = \(x) quantile(x, 0.25),

### 2025-03-14 scripts/calibrate.R
better names

@@ -92,7 +92,7 @@ source(here::here("scripts", "tune", "simulate_uniform.R"))
 # This uses the posterior means as inputs to a single outbreak
 
 calibration_seeds <- availableCores() * 1
-source(here::here("scripts", "tune", "single_outbreak.R"))
+source(here::here("scripts", "tune", "simulate_mean.R"))
 
 # save.image(file.path(out_path, "calibrate.RData"))
 

### 2025-03-14 scripts/tune/simulate_mean.R

@@ -21,7 +21,7 @@ estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
 
 estimated_consts_fixed <- c(estimated_consts_mean, consts)
 
-nl@experiment <- experiment(expname = "single_outbreak_sim",
+nl@experiment <- experiment(expname = "mean_sims",
                             outpath = out_path,
                             repetition = 1,
                             tickmetrics = "true",
@@ -43,24 +43,24 @@ nl@simdesign <- simdesign_simple(nl,
 plan(multisession)
 
 if (run_sims) {
-  single_outbreak_sim <- progressr::with_progress(
+  mean_sims <- progressr::with_progress(
     run_nl_all(nl))
-  write_rds(single_outbreak_sim, file.path(out_path, "single_outbreak_sim.rds"))
+  write_rds(mean_sims, file.path(out_path, "mean_sims.rds"))
 }
 
-single_outbreak_sim <- read_rds(file.path(out_path, "single_outbreak_sim.rds"))
+mean_sims <- read_rds(file.path(out_path, "mean_sims.rds"))
 
-single_outbreak_sim_bak <- single_outbreak_sim
-single_outbreak_sim <- single_outbreak_sim %>% 
+mean_sims_bak <- mean_sims
+mean_sims <- mean_sims %>% 
   filter(`[step]` > 31)
 
-single_outbreak_sim <- single_outbreak_sim %>% 
+mean_sims <- mean_sims %>% 
   group_by(`random-seed`, `siminputrow`) %>% 
   mutate(date_sim = seq(from = ymd("2002-01-01"),
                         by = "1 day",
                         length.out = n()))
 
-single_outbreak_sim_rates <- single_outbreak_sim %>% 
+mean_sims_rates <- mean_sims %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
   group_by(`random-seed`, `siminputrow`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000) %>% 
@@ -69,7 +69,7 @@ single_outbreak_sim_rates <- single_outbreak_sim %>%
   dplyr::select(!c(year, month))
 
 
-single_outbreak_sim_rates %>% 
+mean_sims_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 
   #slice_sample(n=10) %>% 
[…]

### 2025-03-14 paper/paper.qmd
intro

@@
 library(flextable)
-```
-
-```{r}
-# print(here::here())
-# file.exists(here::here("scripts", "calibrate.R"))
+library(progressr)
 ```
@@
 Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
-Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community [@kerrCovasimAgentbasedModel2021].
+Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community population [@kerrCovasimAgentbasedModel2021].
+An important advantage of ABM is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019].
+This can be particularly useful in the preparedness for, and management of, outbreaks of healthcare-associated infections, where the impact of interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].
+"What if?" testing allows for different scenarios or bundles of interventions can be simulated to determine their impact.
 
-To guide infection control strategies, especially during outbreaks, we developed an agent-based simulator and estimated its parameters using data from a CDI outbreak documented by @salgadoAnalysisOutbreakClostridium2009.
+In this study, we describe an agent-based model that simulates the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
+The aim of the model is to guide infection control strategies and interventions, particularly during outbreak periods.
+The parameters are estimated using data from a CDI outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although the model is designed to be adaptable to other related antimicrobial-resistant pathogens such as VRE, ESBL, and CPE.
 
@@
 #| message: false
-single_outbreak_sim_rates %>% 
+mean_sims_rates %>% 
   pivot_wider(names_from = date_sim, values_from = rate) %>% 
@@
 ```{r fig-single-outbreak, include=FALSE}
-single_outbreak_sim_mod <- single_outbreak_sim %>% 
+mean_sims_mod <- mean_sims %>% 
   mutate(prop_infected = `current-colonised` / `current-inpatients`,
          current_not_infected = `current-inpatients` - `current-colonised`)
-single_outbreak_sim_mod %>% 
+mean_sims_mod %>% 
   pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
@@
 
-# single_outbreak_sum_stat <- single_outbreak_sim_mod %>% 
+# single_outbreak_sum_stat <- mean_sims_mod %>% 
 #   group_by(`[step]`) %>%
@@
 parameters <- list(
-  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
+  # p = 1 - unique(mean_sims$`community-colonisation-rate`),      # Proportion of non-infected admissions
   A = 50        # Total admission/discharge rate
@@
 
-single_seeds <- unique(single_outbreak_sim$`random-seed`)
-single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
+single_seeds <- unique(mean_sims$`random-seed`)
+single_outbreak_one_seed <- mean_sims_mod %>% 
   filter(`random-seed` == single_seeds[3])
@@
 beta_function <- function(t, baseline_rate, outbreak_rate) {
-  if (t > unique(single_outbreak_sim$`outbreak-start`) &
-      t < unique(single_outbreak_sim$`outbreak-end`)) {
+  if (t > unique(mean_sims$`outbreak-start`) &
+      t < unique(mean_sims$`outbreak-end`)) {
     return (outbreak_rate)
@@
 
-run_to <- max(single_outbreak_sim$`[step]`)
+run_to <- max(mean_sims$`[step]`)
 
@@
 plan(multisession)
-ode_sims <- furrr::future_pmap_dfr(rate_grid,
+ode_sims <-
+
+    furrr::future_pmap_dfr(rate_grid,
           \(b_base, b_out, c_base, c_out, r) {
@@
                        outbreak = mean(I[1100:1200] / ( N[1100:1200] + I[1100:1200] )))
-        })
+          }, .progress = T)
 
@@
 
-# baselinecolonisation <- 1 - unique(single_outbreak_sim$`community-colonisation-rate`)
-# outbreakcolonisation <- 1 - unique(single_outbreak_sim$`o-community-colonisation-rate`)
+# baselinecolonisation <- 1 - unique(mean_sims$`community-colonisation-rate`)
+# outbreakcolonisation <- 1 - unique(mean_sims$`o-community-colonisation-rate`)
 baselinecolonisation <- mean(abc_single_outbreak$unadj.values[, "c_base"])
@@
 
-```{r fig-single-outbreak-deterministic, include=TRUE}
+```{r fig-single-outbreak-deterministic-one-plot, include=FALSE}
 #| echo: false
+#| message: false
+#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The r"
 single_outbreak_one_seed %>% 
@@
        color = "Compartment") +
-  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
-  geom_line(data = output_df, aes(x = time, y = I), color = "red") +
+  geom_line(data = output_df, aes(x = time, y = N), color = "blue", linetype = "dashed") + 
+  geom_line(data = output_df, aes(x = time, y = I), color = "red", linetype = "dashed") +
   xlim(500, 2000)
+```
+
+```{r fig-single-outbreak-deterministic-faceted, include=TRUE}
+#| echo: false
+#| message: false
+#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model.
+The parameters for the deterministic model were determined using approximate Bayesian computation, using the output from a single run of the agent-based model as a reference."
+output_df_long <- output_df %>% 
+  pivot_longer(cols = c(N, I), names_to = "compartment", values_to = "value") %>% 
+  mutate(sim_type = "deterministic") %>% 
+  mutate(compartment = case_when(
+    compartment == "N" ~ "not_infected",
+    compartment == "I" ~ "infected"
+  ))
+
+single_outbreak_one_seed %>% 
+  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
+  filter(`[step]` > 31) %>%
+  group_by(compartment, `[step]`) %>% 
+  summarise(m = mean(value)) %>% 
+  mutate(sim_type = "abm") %>% 
+  rename(time = `[step]`, value = m) %>%
+  mutate(compartment = case_when(
+    compartment == "current_not_infected" ~ "not_infected",
+    compartment == "current-colonised" ~ "infected"
+  )) %>%
+  bind_rows(output_df_long) %>% 
+  ggplot(aes(x = time, y = value, color = compartment)) +
+  geom_line() +
+  facet_wrap(~ sim_type, nrow = 2) +
+  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
+       x = "Time",
+       y = "Number of Patients",
+       color = "Compartment")
 ```

### 2025-03-14 paper/paper.qmd
paper

@@
 
+## Outbreak generation {#sec-outbreak-generation}
+
+To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.
+
 ## Model inputs and outputs
@@
 
-The approach to calibration was sequential.
-Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate.
+A two-step approach to calibration was taken.
+Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that simulated the baseline rate.
 Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
-To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset.
+To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
 The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
+To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded.
 At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
 The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.
+Again, only samples from the latin hypercube that satisfied these requirements were used.
 
-The posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios. Parameter samples were drawn from a uniform distribution:
+As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates.
+Three approaches to sampling from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were used:
+
+1. Parameter samples drawn from a uniform distribution:
 
@@
 
-where: $\mu$ and \$\sigma\$ are the mean and standard deviation of the posterior distribution, respectively.
-The simulated rates of nosocomial CDI were compared to the observed rates to assess the model's performance.
-One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.
-Since the agent-based model was stochastic, all simulations were aggregated
+where: $\mu$ and $\sigma$ are the mean and standard deviation of the posterior distribution, respectively;
+
+2. The mean of each parameter's posterior distribution; and
+3. Samples with replacement taken from the posterior distribution.
+
+```{r text-seeds, include=FALSE}
+results_outbreak_control_n_seeds <- length(unique(results_outbreak_control_pre_included$`random-seed`))
+uniform_sims_n_seeds <- length(unique(uniform_sims$`random-seed`))
+mean_sims_n_seeds <- length(unique(mean_sims$`random-seed`))
+posterior_predictive_n_seeds <- length(unique(posterior_predictive_sim$`seed-value`))
+```
+
+Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection/colonisation were calculated.
+For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
+for the posterior predictive checks, `r uniform_sims_n_seeds` seeds were used for the uniform distribution,
+`r mean_sims_n_seeds` seeds were used for the mean of the posterior distribution, and
+`r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.
+
+One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method. Since the agent-based model was stochastic, all simulations were aggregated
 
@@
 | Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
-| Antimicrobial effect | Uniform | \$ a = 1.31, b = 1.87 \$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
+| Antimicrobial effect | Uniform | $ a = 1.31, b = 1.87 $ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
 | CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
@@
 
-![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="30%"}
+![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="7in"}
 
 Users can run the model interactively using the interface shown in @fig-screenshot.
-
-## Outbreak generation {#sec-outbreak-generation}
-
-To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.
-During calibration, the upper or lower bounds for parameter priors during the outbreak period were extended by 50% of baseline.
-For example, the upper bound of prior for the parameter determining the rate of antimicrobial prescriptions was 50% higher than the baseline rate.
-On the other hand, these parameters were reversed to control the outbreak.
 
@@
 Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
-In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_uniform_sims_rates$rate_mean, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_uniform_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
-During the outbreak period, simulations had a mean rate of `r round(outbreak_uniform_sims_rates$rate_mean, 2)` \[IQR `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real mean rate of 3.74 \[IQR 3.32--4.41\] (see @fig-simulated-rates).
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-posterior-predictive-check).
 
@@
 #| echo: false
+#| fig-cap: "Sensitivity analysis of the model to changes in key infection control parameters. The y axis shows the total number of nosocomial infections during the outbreak period."
 list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
@@
             .groups = "keep") %>% 
+  mutate(param = case_when(
+    param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
+    param == "c-proportion-redistributed" ~ "Proportion redistributed",
+    param == "c-toilet-cleaning-rate" ~ "Toilet cleaning rate",
+    param == "c-toilet-cleaning-effect" ~ "Toilet cleaning effect"
+  )) %>% 
   ggplot(aes(x = value, y = mean_cases)) + 
@@
   geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
-  facet_wrap(~ param, scales = "free")
+  facet_wrap(~ param, scales = "free") +
+  xlab("Parameter value") +
+  ylab("Number of infected patients")
 ```
 
-```{r fig-sensitvity-pre-outbreak-abx, include=TRUE, eval=TRUE}
+```{r fig-sensitvity-pre-outbreak-abx, include=TRUE}
 #| echo: false
+#| fig-cap: "Sensitivity analysis of the model to changes in the pre-outbreak antimicrobial prescription rate. The y axis shows the total number of nosocomial infections during the outbreak period."
 sens_abx_prescription_pre_outbreak_outbreak_cases %>% 
@@
                      lower = mean_cases - sd_cases),
-            aes(y = lower), linetype = "dashed")
+            aes(y = lower), linetype = "dashed") +
+  xlab("Antimicrobial prescription rate") +
+  ylab("Number of infected patients")
 ```
@@
   pivot_longer(cols = c(N, I), names_to = "compartment", values_to = "value") %>% 
-  mutate(sim_type = "deterministic") %>% 
+  mutate(sim_type = "Deterministic") %>% 
   mutate(compartment = case_when(
-    compartment == "N" ~ "not_infected",
-    compartment == "I" ~ "infected"
+    compartment == "N" ~ "Not infected",
+    compartment == "I" ~ "Infected"
   ))
@@
   summarise(m = mean(value)) %>% 
-  mutate(sim_type = "abm") %>% 
+  mutate(sim_type = "AMB") %>% 
   rename(time = `[step]`, value = m) %>%
   mutate(compartment = case_when(
-    compartment == "current_not_infected" ~ "not_infected",
-    compartment == "current-colonised" ~ "infected"
+    compartment == "current_not_infected" ~ "Not infected",
+    compartment == "current-colonised" ~ "Infected"
   )) %>%
@@
   facet_wrap(~ sim_type, nrow = 2) +
-  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
-       x = "Time",
+  labs(x = "Time",
        y = "Number of Patients",
-       color = "Compartment")
+       color = "Compartment/Agent state")
+
 ```

### 2025-09-10 paper/paper.qmd
quarto settings

@@
 ---
-title: Paper
+title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
 authors:
@@
 bibliography: references.bib
+csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
 ---

### 2025-09-15 paper/paper.qmd
acronyms markdown and feedback

--- 2025-03-14 paper/paper.qmd
@@
 bibliography: references.bib
+csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
+filters:
+  - acronyms
+acronyms:
+  keys:
+  - shortname: CDI
+    longname: _Clostridium difficile_ infection
+    markdown: longname
+  - shortname: VRE
+    longname: vancomycin-resistant _Enterococcus_
+    markdown: longname
+  - shortname: ESBL
+    longname: extended-spectrum beta-lactamase
+  - shortname: CPE
+    longname: carbapenemase-producing Enterobacterales
+  - shortname: WHO
+    longname: World Health Organization
+  - shortname: IPC
+    longname: infection prevention and control
+  - shortname: ABM
+    longname: agent-based model
+  - shortname: MRSA
+    longname: methicillin-resistant _Staphylococcus aureus_
+    markdown: longname
+  - shortname: SARS-CoV-2
+    longname: severe acute respiratory syndrome coronavirus 2
+  - shortname: ABC
+    longname: approximate Bayesian computation
+  - shortname: ICR
+    longname: interquartile range
 ---
@@
 
-```{r load-data, include=FALSE}
+```{r load-data, include=FALSE, eval=FALSE}
+# This runs the calibrate scripts and loads objects into memory.
 # print(file.exists(here::here("scripts", "calibrate.R")))
@@
 
-Transmission of potentially pathogenic bacteria under antimicrobial selective pressure has important implications for control of hospital-acquired infection.
-Examples of such pathogens include *Clostridioides difficile* infection (CDI), vancomycin-resistant *Enterococcus* (VRE), extended-spectrum beta-lactamase-producing (ESBL) Enterobacterales, and carbapenem-resistant Enterobacterales (CPE).
-Infection with these pathogens can lead to increased morbidity and mortality -- for example, CPE bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
-_C. difficile_ infection is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
-Furthermore, the therapeutic antimicrobial options available for VRE, ESBL, and CPE are limited, leading to increased use of WHO reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].
+```{r load-pre-processed-data, include=FALSE}
+# Alternatively, just load previously dumped .RData from a run of calibrate.R
+load(file.path("..", "out", "calibrate.RData"))
+```
 
-Infection prevention and control (IPC) strategies are essential to prevent the spread of these pathogens within a healthcare setting, and include measures such as hand hygiene, environmental cleaning, isolation of infected/colonised patients, and more [@worldhealthorganizationGlobalReportInfection2022].
-In many healthcare settings, IPC strategies and operational practices are led by a multi-disciplinary IPC team.
-A key role of IPC teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
+The transmission of pathogenic bacteria in hospitals results in excessive morbidity and mortality.
+Relevant examples include *Clostridioides difficile*, \acr{VRE}, \acr{ESBL}-producing Enterobacerales, and \acr{CPE}.
+Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
+\acr{CDI} is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
+Furthermore, the therapeutic antimicrobial options available for \acr{VRE}, \acr{ESBL}, and \acr{CPE} are limited, leading to increased use of \acr{WHO} reserve agents, which further exacerbates the problem of antimicrobial resistance [@organization2019WHOAWaRe2019].
+
+\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings.
+Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
+A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
 In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
@@
 In practice, the approach to managing outbreaks is often reactive.
-Once an outbreak is identified and declared, the IPC team implements a set of enhanced infection control measures to attempt to attempt to halt transmission.
-The number of cases of infection or colonisation is monitored over a time frame dependent on the pathogen, setting, and scale of the outbreak.
+The number of cases of infection is monitored over a time frame dependent on the pathogen, setting, and scale of the outbreak.
 If the measures are successful in controlling the spread, then they are gradually relaxed once rates return to baseline.
@@
 
-Computer simulation approaches such as agent-based modelling (ABM) can be used to simulate the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
-Agent-based modelling has been used to model various infectious diseases, such as the transmission of important nosocomial pathogens such as methicillin-resistant _Staphylococcus aureus_ (MRSA) within a healthcare setting [@jaramilloAgentBasedModel2015], and the spread of SARS-CoV-2 within a community population [@kerrCovasimAgentbasedModel2021].
-An important advantage of ABM is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019].
-This can be particularly useful in the preparedness for, and management of, outbreaks of healthcare-associated infections, where the impact of interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].
-"What if?" testing allows for different scenarios or bundles of interventions can be simulated to determine their impact.
+The use of computer simulation techniques such as \acr{ABM} can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
+Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021].
+An advantage of \acr{ABM} over other simulation approaches is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019].
+This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].
 
-In this study, we describe an agent-based model that simulates the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
-The aim of the model is to guide infection control strategies and interventions, particularly during outbreak periods.
-The parameters are estimated using data from a CDI outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although the model is designed to be adaptable to other related antimicrobial-resistant pathogens such as VRE, ESBL, and CPE.
+Here, we describe an agent-based model that describes the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
+Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods.
+The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.
 
@@
 
-An agent-based model (ABM) was implemented using NetLogo [@wilenskyu.NetLogo1999].
+An \acr{ABM} was implemented using NetLogo [@wilenskyu.NetLogo1999].
 Briefly, the simulation started by building a virtual hospital with a configurable high-level structure, defined by its number of wards, beds per ward, and proportion of bedspaces sharing toilet facilities.
-Next, the hospital was populated with patients, each having an infected/colonised or non-infected/colonised state, defined as:
+Next, the hospital was populated with patients, each having an infected or non-infected state, defined as:
 
@@
 
--   $x_p(t) = 0$ if patient $p$ is not infected/colonised at time $t$,
--   $x_p(t) = 1$ if patient $p$ is infected/colonised at time $t$.
+-   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
+-   $x_p(t) = 1$ if patient $p$ is infected at time $t$.
 
-It was assumed that infection/colonisation was not reversible during the simulation, i.e., that infected/colonised patients remained so until discharge.
+Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained colonised until discharge.
 At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:
@@
 
--   $R_i(t)$ is the risk of patient $i$ acquiring infection/colonisation at time $t$,
+-   $R_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,
 
@@
 
-The agent-based model conducted Bernoulli trials to determine if a patient acquired infection/colonisation at each time step $t$:
+The agent-based model conducted Bernoulli trials to determine if a patient became infected at each time step $t$:
 
@@
 
-where $X_\text{de novo}$ = 1 if patient acquires infection/colonisation *de novo* at time $t$, and 0 otherwise.
+where $X_\text{de novo}$ = 1 if patient becomes infected *de novo* at time $t$, and 0 otherwise.
 
@@
 
-where $X_\text{toilet}$ = 1 if patient acquires infection/colonisation from a contaminated toilet at time $t$, and 0 otherwise.
+where $X_\text{toilet}$ = 1 if patient becomes infected from a contaminated toilet at time $t$, and 0 otherwise.
 
@@
 
-where $X_i(t)$ = 1 if patient $i$ acquires infection/colonisation at time $t$, and 0 otherwise.
+where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.
 
@@
 
-To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures.
+To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
 
@@
 
-The outputs of the model included: 1) number of patients admitted; 2) number of patients with community-acquired colonisation/infection; 3) number of patients with hospital-acquired (nosocomial) colonisation/infection.
+The outputs of the model included: 1) number of patients admitted to the hospital; 2) number of patients with community-acquired infection; 3) number of patients with hospital-acquired (nosocomial) infection.
 The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.
@@
 
-Data from a hospital *Clostridioides difficile* infection (CDI) outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009].
+Data from a hospital \acr{CDI} outbreak reported by Salgado et al was used to parameterise and calibrate the model [@salgadoAnalysisOutbreakClostridium2009].
 Raw data from this study were not available, therefore, the Engauge digitizer software was used to extract observed CDI rates from the manuscript's figures [@mitchellEngaugeDigitizer2020].
@@
 
-Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnakerApproximateBayesianComputation2013].
-Priors were determined from domain knowledge or extracted from existing literature.
+Rejection sampling \acr{ABC} was employed to fit the model to the outbreak data [@sunnakerApproximateBayesianComputation2013].
+Priors were extracted from existing literature; flat priors when data were unavailable.
 A summary of the priors used is available in @tbl-priors.
+The proportion of toilets that were shared between multiple patients was set to 60%.
 
@@
 Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
-The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
-Results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.
+The number of daily cases of nosocomial \acr{CDI} were extracted from the model outputs, and converted to a time series of nosocomial \acr{CDI} per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
+Since each simulation starts with an empty hospital that is gradually populated with patient admissions, results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.
 
@@
 The parameter samples that generated the accepted simulations were used to produce a posterior distribution.
-$\epsilon$ was set so that the acceptance rate was approximately 2.5%.
+$\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.
 
 A two-step approach to calibration was taken.
-Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that simulated the baseline rate.
+Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate.
 Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
@@
 
-Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection/colonisation were calculated.
+Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated.
 For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
@@
 
-One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method. Since the agent-based model was stochastic, all simulations were aggregated
+One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.
 
@@
 To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
-Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after colonisation/infection; 3) toilet cleaning rate; and 4) toilet cleaning effect.
-To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection ABC.
-Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value.
+Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after infection; 3) toilet cleaning rate; and 4) toilet cleaning effect.
+To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value.
 The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period.
@@
 
-An agent-based model that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced.
-After calibrating the agent-based model using approximate Bayesian computation (ABC), three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control.
-A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection ABC) is shown in @tbl-posteriors.
+An \acr{ABM} that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced.
+After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control.
+A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors.
 The densities of the posterior parameter distributions are shown in @fig-posterior-densities.
@@
 #| warning: false
-#| label: fig-posterior-densities
 #| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods.
@@
 Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
-In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[interquartile (IQR) range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
-During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[IQR `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[IQR 3.32--4.41\] (see @fig-posterior-predictive-check).
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{ICR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[\acr{IQR} `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[\acr{IQR} 3.32--4.41\] (see @fig-posterior-predictive-check).
 
@@
 #| echo: false
-#| label: fig-abc-rates-retained-posteriors
 #| fig.cap: "Simulated rates of _C.
@@
   geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
-  guides(color="none")
+  guides(color="none") +
+  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
 ```
@@
 #| echo: false
-#| fig-cap: "Sensitivity analysis of the model to changes in key infection control parameters. The y axis shows the total number of nosocomial infections during the outbreak period."
-list(abx_prescription_rate = sens_abx_prescription_outbreak_cases,
+#| fig-cap: "Sensitivity analysis of the model to changes in key infection control parameters.
+The y axis shows the total number of nosocomial infections during the outbreak period.
+Antimicrobial prescription rate represents the percentage of patients that were administered an antimicrobial during their admission.
+Proportion redistributed represents the proportion of infected patients that were moved to a single-occupancy room (lower values represent fewer patients being moved).
+Toilet cleaning effect represents the effectiveness of cleaning in reducing toilet contamination (100% = complete removal of contamination; 0% = no cleaning effect).
+Toilet cleaning rate represents the number of times toilets were cleaned per day, on average."
+list(
+  abx_prescription_rate    = sens_abx_prescription_outbreak_cases,
   proportion_redistributed = sens_prop_redist_outbreak_cases,
   toilet_cleaning_rate     = sens_toilet_cleaning_rate_outbreak_cases,
-     toilet_cleaning_effect = sens_toilet_cleaning_effect_outbreak_cases) %>%
+  toilet_cleaning_effect   = sens_toilet_cleaning_effect_outbreak_cases
+) %>%
   bind_rows() %>% 
-  pivot_longer(cols = starts_with("c-"),
+  pivot_longer(
+    cols = starts_with("c-"),
     names_to = "param",
-               values_drop_na = TRUE) %>% 
+    values_drop_na = TRUE
+  ) %>% 
   group_by(param, value) %>% 
-  summarise(mean_cases = mean(`outbreak-cases`),
+  summarise(
+    mean_cases = mean(`outbreak-cases`),
     sd_cases   = sd(`outbreak-cases`),
-            .groups = "keep") %>% 
-  mutate(param = case_when(
+    .groups    = "keep"
+  ) %>% 
+  mutate(
+    # convert antibiotic prescription rate to 0-100 scale, leave others unchanged
+    # value = if_else(param == "c-antibiotic-prescription-rate", value * 100, value),
+    param = case_when(
       param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
@@
       param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate",
-    param == "c-toilet-cleaning-effect" ~ "Toilet cleaning effect"
-  )) %>% 
+      param == "c-toilet-cleaning-effect"       ~ "Toilet cleaning effect",
+      TRUE                                      ~ param
+    )
+  ) %>% 
   ggplot(aes(x = value, y = mean_cases)) + 
   geom_line() + 
-  # shade sd
   geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
-  facet_wrap(~ param, scales = "free") +
-  xlab("Parameter value") +
-  ylab("Number of infected patients")
+  facet_wrap(~ param, scales = "free") +    # free_x required to use facetted_pos_scales
+  # apply percentage labels and a per-facet x-title only for the antibiotic facet
+  ggh4x::facetted_pos_scales(
+    x = list(
+      # LHS is evaluated against the facet layout; RHS is the scale to use for matching facets
+      param %in% c("Antimicrobial prescription rate", "Proportion redistributed", "Toilet cleaning effect") ~
+        scale_x_continuous(
+          labels = scales::label_percent(scale = 100)   # data already on 0-100 so use scale=1
+        )
+    )
+  ) +
+  # fallback labels (these apply to facets that did not get a custom scale)
+  labs(x = "Parameter value",
+       y = "Number of infected patients")
 ```
@@
   summarise(m = mean(value)) %>% 
-  mutate(sim_type = "AMB") %>% 
+  mutate(sim_type = "ABM") %>% 
   rename(time = `[step]`, value = m) %>%
@@
 
-Infection prevention and control (IPC) is the practice of preventing the spread of infectious diseases in healthcare settings.
-A central responsibility is the operational management of an outbreak, defined as an increase in the number of infected cases above the expected rate.
+\acr[case=sentence, first_use=true]{IPC} is the practice of preventing the spread of infectious diseases in healthcare settings [@worldhealthorganizationGuidelinesCoreComponents2016].
+Central to this is the operational management of an outbreak, defined as an increase in the number of infected cases above the expected rate.
 This research describes the design and calibration of an agent-based model that simulates the spread of infection in a hospital setting.
 The primary motivation of this work is to assess the impact of different infection prevention interventions (such as change in practice) on rates of infection.
-The model was designed to simulate a paradigm of infections that are faecal-oral-transmitted, with the risk increasing in the presence of the selective pressure introduced by antimicrobial therapy.
-Since some of the pathogens that fit this definition may be associated with long-term colonisation of the gastro-intestinal tract (without the presence of clinical illness), infection and colonisation are used inter-changeably within this research.
-The model can be used to simulate any infection with these characteristics, such as: _Clostridioides difficile_, vancomycin-resistant _Enterococcus_, extended-spectrum beta-lactamase-producing Enterobacterales, and carbapenem-resistant _Enterobacterales_.
+The model was designed to simulate a paradigm of infections that are characterised by faecal-oral transmission, with the risk increasing in the presence of the selective pressure introduced by antimicrobial therapy.
+Since some of the pathogens that fit this definition are associated with long-term colonisation and risk of onward transmission of the gastro-intestinal tract (without the presence of clinical illness), it was assumed that infected patients remained colonised during their admission.
+The model can be used to simulate any infection with these characteristics, such as: \acr{CDI}, \acr{VRE}, \acr{ESBL}, and \acr{CPE}.
 
@@
 Some models have also explored the impact of one or more infection control interventions on the spread of infection [@barkerInterventionsReduceIncidence2018; @rubinSimulationBasedAssessmentStrategies2013].
-Our model builds on these works by designing and calibrating the model to an outbreak and enhanced IPC period.
+Our model builds on these works by designing and calibrating the model to an outbreak and enhanced \acr{IPC} period.
 The motivation for this was to design a model that can be used to support infection and prevention teams in the following: 1) preparedness for outbreak situations; and 2) decision-making on the implementation of infection control interventions during an outbreak.

### 2025-09-15 scripts/calibrate.R

@@ -141,3 +141,6 @@ source(here::here("scripts", "sensitivity", "toilet_cleaning_effect.R"))
 ####================= sensitivity analysis abx prescription rates preoutbreak ==============####
 
 source(here::here("scripts", "sensitivity", "abx_prescription_rates_pre_outbreak.R"))
+
+### save all objects
+save.image(file.path(out_path, "calibrate.RData"))

### 2025-10-02 paper/paper.qmd
SD model

@@
     longname: approximate Bayesian computation
-  - shortname: ICR
+  - shortname: IQR
     longname: interquartile range
+  insert_loa: false
 ---
@@
 run_sims <- FALSE
+set.seed(42)
 
@@
 Relevant examples include *Clostridioides difficile*, \acr{VRE}, \acr{ESBL}-producing Enterobacerales, and \acr{CPE}.
-Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with an increased risk of in-hospital mortality (risk difference of 0.25) compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
+Drug resistance further compounds the problem --- for example, \acr{CPE} bloodstream infection is associated with a considerably increased risk of in-hospital mortality compared to carbapenem-susceptible infection [@budhramHealthOutcomesAttributable2020].
 \acr{CDI} is associated with prolonged hospital stay and increased healthcare costs, causing a significant operational burden [@forsterEffectHospitalacquiredInfection2012].
@@
 Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
-Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such rates of antimicrobial usage, and to structural differences such as hospital layout.
+Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.
 
@@
 
-Here, we describe an agent-based model that describes the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
+Here, we describe an agent-based model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
 Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods.
@@
 $$
-x_p(t) \epsilon \{0, 1\}
+x_p(t) \in \{0, 1\}
 $$
@@
 
+-   $t \in \mathbb{N}_0 = \{0, 1, 2, \ldots\}$ is the time step (in days),
 -   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
@@
 $$ 
-P_\text{de novo}(t) = 1 - \text{Exp} {( - r_0 T(t) )},
+P_\text{de novo}(t) = 1 - \text{exp}{ {( - r_0 T(t) )} },
 $$
@@
 $$
-P_\text{toilet}(t) = 1 - \text{Exp} ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
+P_\text{toilet}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} \right) }
 $$
@@
 
--   $T(t)$ is a binary indicator for antimicrobial therapy:
+-   $m$ is the increased (or decreased) relative risk of infection/colonisation due to antimicrobial therapy,
+
+-   $U_i(t)$ is the number of times patient $p$ frequents the toilet within the time interval $t$,
+
+-   $C(t_j)$ is the contamination level of the toilet at time $t_j$, and
 
@@
 
--   $m$ is the increased (or decreased) relative risk of infection/colonisation due to antimicrobial therapy,
-
--   $U_i(t)$ is the number of times patient $i$ frequents the toilet within the time interval $t$,
-
--   $C(t_j)$ is the contamination level of the toilet at time $t_j$.
-
 The agent-based model conducted Bernoulli trials to determine if a patient became infected at each time step $t$:
@@
 
-The results of the trials were combined to determine the outcome for patient $i$ at time $t$:
-
-$$
-X_i(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
-$$
-
-where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.
+The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was sucessful:
+
+$$
+X_p(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
+$$
+
+where $X_i(t)$ = 1 if patient $p$ becomes infected at time $t$, and 0 otherwise.
+
+## Deterministic model
 
@@
 $$
-\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S - rS
-$$
-
-$$
-\frac{dI}{dt} = (1 - p)A - D \cdot \frac{I}{S + I} + \beta S + rS
+\frac{dS}{dt} = (1-p)A - D \cdot \frac{S}{S + I} - \beta S - rS
+$$
+
+$$
+\frac{dI}{dt} = pA - D \cdot \frac{I}{S + I} + \beta S + rS
 $$
@@
 -   $A$ is the rate of new admissions,
--   $p$ is the proportion of new admissions that are susceptible/non-infected, while $(1 - p)$ is community colonisation/infection rate,
+-   $p$ is the community colonisatio rate,
 -   $\beta$ is the transmission coefficient,
@@
 The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $D = A$.
+
+![System dynamics model](images/sd_model.svg)
 
@@
 | Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
-| Antimicrobial effect | Uniform | $ a = 1.31, b = 1.87 $ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
+| Antimicrobial effect | Uniform | $a = 1.31, b = 1.87$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
 | CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
@@
 #| fig.width: 18
-#| fig.height: 8
+#| fig.height: 18
 #| warning: false
@@
 Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
-#| fig-subcap:
-#|   - "General parameters"
-#|   - "Toilet-related parameters"
-#| layout-ncol: 1
-abc_params_outbreak_control$unadj.values %>% 
+general_params <- abc_params_outbreak_control$unadj.values %>% 
   as_tibble %>% 
@@
   mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
-  ggplot(aes(x = value, ..scaled.., fill = type)) +
-  geom_density(alpha = 0.1) +
+  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
+  geom_density(alpha = .3) +
   facet_wrap(~name, scales = "free_x") +
+  theme_minimal() +
   # remove y axis
@@
   # manually set colors
-  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))
-
-abc_params_outbreak_control$unadj.values %>% 
+  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
+  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
+  # remove gridlines
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+toilet_params <- abc_params_outbreak_control$unadj.values %>% 
   as_tibble %>% 
@@
   mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
-  ggplot(aes(x = value, ..scaled.., fill = type)) +
-  geom_density(alpha = 0.1) +
+  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
+  geom_density(alpha = 0.3) +
   facet_wrap(~name, scales = "free_x") +
+  theme_minimal() + 
   # remove y axis
@@
   # manually set colors
-  scale_fill_manual(values = c("Baseline" = "green", "Outbreak" = "red", "Control" = "blue"))
+  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
+  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
+  # remove gridlines
+  theme(panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank())
+
+# combine plots
+library(patchwork)
+general_params / toilet_params + plot_layout(heights = c(0.5, 1))
 ```
@@
 Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
-In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{ICR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[IQR 0.94--1.56\].
+In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{IQR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[\acr{IQR} 0.94--1.56\].
 During the outbreak period, simulations had a median rate of `r round(outbreak_uniform_sims_rates$rate_median, 2)` \[\acr{IQR} `r round(outbreak_uniform_sims_rates$rate_q_low, 2)`--`r round(outbreak_uniform_sims_rates$rate_q_high, 2)`\] cases, compared with the real median rate of 3.74 \[\acr{IQR} 3.32--4.41\] (see @fig-posterior-predictive-check).
@@
 
+```{r}
+stochastic_sd <- read_csv("../out/sd_stochastic experiment-table.csv", skip = 6)
+```
+
+```{r}
+Rcpp::sourceCpp("../models/sd_stochastic.cpp")
+```
+
+```{r}
+set.seed(42)
+randomLHS_sd <- randomLHS(1000000, 8)
+sd_param_grid <- data.frame(init_population = rep(610, 1000),
+                            A = 68,
+                            r = qunif(randomLHS_sd[, 3], 0.001, 0.01),
+                            baseline_beta = qunif(randomLHS_sd[, 4], 0.0001, 0.01),
+                            baseline_p = qunif(randomLHS_sd[, 5], 0.001, 0.1),
+                            outbreak_beta = qunif(randomLHS_sd[, 6], 0.0001, 0.01),
+                            outbreak_p = qunif(randomLHS_sd[, 7], 0.001, 0.1),
+                            outbreak_start = rep(1042, 1000),
+                            outbreak_end = rep(1234, 1000),
+                            n_days = rep(2585, 1000))
+
+sd_param_grid <- sd_param_grid %>% 
+  filter(baseline_beta < outbreak_beta & baseline_p < outbreak_p) %>% 
+  slice_sample(n = 10000) %>% 
+  mutate(param_row = row_number())
+n_seeds <- 2
+sd_sims <- 
+  pmap_dfr(sd_param_grid,
+        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row) {
+          map(1:n_seeds,
+              \(seed) {
+                set.seed(seed)
+                sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
+                  mutate(`random-seed` = seed,
+                         param_row = param_row)
+              }) %>% 
+            bind_rows()
+        }, .progress = T)
+
+sd_sims <- sd_sims %>% 
+  filter(day > 31) %>% 
+  group_by(param_row, `random-seed`) %>% 
+  # add dates
+  mutate(date_sim = seq(from = ymd("2002-01-01"),
+                        by = "1 day",
+                        length.out = n()))
+
+# rates by month
+sd_sims_rates <- sd_sims %>% 
+  group_by(param_row, `random-seed`) %>% 
+  mutate(year = year(date_sim),
+         month = month(date_sim)) %>%
+  group_by(param_row, `random-seed`, year, month) %>%
+  # calculate rate of hosp_cases per 1000 patient days
+  summarise(rate = (sum(hosp_cases) / sum(S + I)) * 1000,
+            date_sim = first(date_sim))
+
+# collapse to mean rates by seed
+sd_sims_rates_mean <- sd_sims_rates %>% 
+  group_by(param_row, year, month) %>%
+  summarise(rate = mean(rate),
+            date_sim = first(date_sim))
+
+# pivot wider to get ready for abc
+
+sd_sumstats <- sd_sims_rates_mean %>% 
+  ungroup() %>% 
+  dplyr::select(-year, -month) %>%
+  filter(date_sim <= max(salgado$x)) %>% 
+  pivot_wider(names_from = date_sim, values_from = rate)
+
+# abc
+abc_sd <- abc(salgado$rates,
+              sd_param_grid, dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')
+
+# posterior predictive checks
+posterior_check_params <- abc_sd$unadj.values %>% 
+  as_tibble() %>% 
+  slice_sample(n = 100, replace = TRUE) %>% 
+  mutate(seed = sample(1:1000, n(), replace = FALSE))
+
+posterior_predictive_sim_sd <- 
+  pmap_dfr(posterior_check_params,
+        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row, seed) {
+            set.seed(seed)
+            sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
+              mutate(`random-seed` = seed,
+                     param_row = param_row) %>% 
+              bind_rows()
+        }, .progress = T)
+
+posterior_predictive_sim_sd <- posterior_predictive_sim_sd %>%
+  filter(day > 31) %>% 
+  group_by(`random-seed`) %>% 
+  # add dates
+  mutate(date_sim = seq(from = ymd("2002-01-01"),
+                        by = "1 day",
+                        length.out = n()))
+
+posterior_predictive_sim_sd_rates <- posterior_predictive_sim_sd %>%
+  group_by(`random-seed`) %>% 
+  mutate(year = year(date_sim),
+         month = month(date_sim)) %>%
+  group_by(`random-seed`, year, month) %>%
+  # calculate rate of hosp_cases per 1000 patient days
+  summarise(rate = (sum(hosp_cases) / sum(S + I)) * 1000,
+            date_sim = first(date_sim))
+
+```
+
 ```{r fig-single-outbreak-deterministic-one-plot, include=FALSE}
@@
 ```
+```{r fig-abc-sd-posteriors, include=TRUE}
+#| echo: false
+#| fig.cap: "Posterior distributions of parameters obtained from approximate Bayesian computation using summary statistics generated from the stochastic deterministic model.
+The red line shows the prior distribution; the blue line shows the posterior distribution."
+```
+
+```{r fig-abc-sd-simulated-rates, include=TRUE}
+#| echo: false
+#| fig.cap: "Simulated rates of _C.
+difficile_ infection generated from the simulations retained after applying the rejection approximate Bayesian computation algorithm to the stochastic deterministic model.
+The red line shows the true observed infection rates.
+The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
+
+abc_sd$ss %>% as.matrix %>% 
+  as.data.frame() %>% 
+  pivot_longer(cols = everything(), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim) %>%
+  summarise(m = median(rate),
+            q_high = quantile(rate, 0.75),
+            q_low = quantile(rate, 0.25),
+            .groups = "keep") %>%
+  ggplot() +
+  geom_line(aes(x = date_sim, y = m)) +
+  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
+  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none") +
+  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
+```
+
+```{r fig-abc-sd-posterior-predictive-check, include=TRUE}
+#| echo: false
+#| fig.cap: "Posterior predictive check of the stochastic deterministic model.
+Simulations were generated by sampling with replacement from the posterior parameter distributions after approximate Bayesian computation.
+Each simulation was run using different random number generators.
+The red line shows the true observed infection rates.
+The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
+
+posterior_predictive_sim_sd_rates %>% 
+  pivot_wider(names_from = date_sim, values_from = rate) %>% 
+  pivot_longer(cols = !c(`random-seed`, `param_row`), names_to = "date_sim", values_to = "rate") %>% 
+  mutate(date_sim = ymd(date_sim)) %>% 
+  group_by(date_sim) %>%
+  summarise(m = median(rate),
+            q_high = quantile(rate, 0.75),
+            q_low = quantile(rate, 0.25),
+            .groups = "keep") %>%
+  ggplot() +
+  geom_line(aes(x = date_sim, y = m)) +
+  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
+  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
+  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
+  guides(color="none") +
+  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
+
+```
 

### 2025-10-02 models/sd_stochastic.cpp

@@ -1,0 +1,94 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// helper: integer binomial draw
+inline int rbinom_int(int n, double p) {
+  if (n <= 0 || p <= 0.0) return 0;
+  if (p >= 1.0) return n;
+  return R::rbinom(n, p);
+}
+
+// [[Rcpp::export]]
+DataFrame sd_stochastic(
+    int init_population,
+    int A,                  // mean discharges per day
+    double r,               // spontaneous colonisation rate
+    double baseline_beta,
+    double baseline_p,
+    double outbreak_beta,
+    double outbreak_p,
+    int outbreak_start,
+    int outbreak_end,
+    int n_days
+) {
+  // initial state
+  int S = std::round(init_population * (1.0 - baseline_p));
+  int I = std::round(init_population * baseline_p);
+  
+  // output storage
+  IntegerVector days(n_days + 1);
+  IntegerVector S_out(n_days + 1);
+  IntegerVector I_out(n_days + 1);
+  IntegerVector comm_cases(n_days + 1);
+  IntegerVector hosp_cases(n_days + 1);
+  IntegerVector dis_S_out(n_days + 1);
+  IntegerVector dis_I_out(n_days + 1);
+  
+  days[0]        = 0;
+  S_out[0]       = S;
+  I_out[0]       = I;
+  comm_cases[0]  = 0;
+  hosp_cases[0]  = 0;
+  
+  // simulation loop
+  for (int t = 1; t <= n_days; ++t) {
+    double beta, p;
+    if (t > outbreak_start && t < outbreak_end) {
+      beta = outbreak_beta;
+      p    = outbreak_p;
+    } else {
+      beta = baseline_beta;
+      p    = baseline_p;
+    }
+    
+    // --- Discharges ---
+    int total_leave = R::rpois(A);
+    int total_pop   = S + I;
+    if (total_pop == 0) total_pop = 1;
+    
+    int dis_S = std::round(total_leave * ((double) S / total_pop));
+    int dis_I = total_leave - dis_S;
+    
+    dis_S_out[t] = dis_S;
+    dis_I_out[t] = dis_I;
+    
+    // --- Admissions (equal to discharges) ---
+    int adm_S = rbinom_int(total_leave, 1.0 - p);
+    int adm_I = total_leave - adm_S;
+    
+    // --- In-hospital infections ---
[…]

### 2025-10-07 scripts/calibrate.R
fix nl version 6.2.2

@@ -13,6 +13,7 @@ bedspaces_per_ward <- 9
 set.seed(42)
 
 netlogo_path <- "/Applications/NetLogo 6.2.2/"
+nl_version <- "6.2.2"
 model_path <- here::here("models", "hospital_transmission.nlogo")
 out_path <- here::here("out")
 data_path <- here::here("data")

### 2025-10-07 scripts/sensitivity/proportion_redistributed_perfect_clean.R

@@ -10,7 +10,7 @@ sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado.nlogo"
 
-nl <- nl(nlversion = "6.2.2",
+nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
          jvmmem = 1024)

### 2025-10-07 scripts/sensitivity/toilet_cleaning_effect.R

@@ -2,7 +2,7 @@ sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado.nlogo"
 
-nl <- nl(nlversion = "6.2.2",
+nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
          jvmmem = 1024)

### 2025-10-07 scripts/tune/fine_tune_pre_outbreak.R

@@ -1,6 +1,6 @@
 sim_days <- (365 * 3) + 30
 
-nl <- nl(nlversion = "6.2.2",
+nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
          jvmmem = 1024)

### 2025-10-07 scripts/tune/outbreak_and_control.R

@@ -2,7 +2,7 @@ sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado_v0.3.nlogo"
 
-nl <- nl(nlversion = "6.2.2",
+nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
          jvmmem = 1024)

### 2025-10-07 scripts/tune/pre_outbreak.R

@@ -1,7 +1,7 @@
 sim_days <- (365 * 3) + 30
 # sim_hours <- sim_days * 24
 
-nl <- nl(nlversion = "6.2.2",
+nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
          jvmmem = 1024)

### 2025-10-07 paper/paper.qmd
tune and sens now use side-room-delay-mean

@@
                  "random-colonisation" = "cases per 10,000 bed days",
-                 "proportion-redistributed" = "proportion of patients")
+                 "side-room-delay-mean" = "days")
 
@@
                  "random-colonisation" = "Random colonisation",
-                 "proportion-redistributed" = "Proportion redistributed")
+                 "side-room-delay-mean" = "Side room delay mean")
 
@@
       param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
-      param == "c-proportion-redistributed"     ~ "Proportion redistributed",
+      param == "c-side-room-delay-mean"     ~ "Side room delay mean",
       param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate",

### 2025-10-07 scripts/sensitivity/proportion_redistributed.R

@@ -19,7 +19,7 @@ consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
 estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
   apply(., 2, mean)
 
-estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-proportion-redistributed"]
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-side-room-delay-mean"]
 
 nl@experiment <- experiment(expname = "sens_analysis",
                             outpath = out_path,
@@ -36,8 +36,8 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-proportion-redistributed" = list(min=0.01,
-                                                                                 max=0.99,
+                            variables = list("c-side-room-delay-mean" = list(min=0.01,
+                                                                                 max=7,
                                                                                  qfun='qunif')
                             )
 )
@@ -70,7 +70,7 @@ sens_prop_redist <- sens_prop_redist %>%
 
 sens_prop_redist_rates <- sens_prop_redist %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
-  group_by(`random-seed`, `c-proportion-redistributed`, year, month) %>% 
+  group_by(`random-seed`, `c-side-room-delay-mean`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000) %>% 
   mutate(date_sim = ymd(paste0(year, "-", month, "-01"))) %>% 
   ungroup() %>% 
@@ -86,19 +86,19 @@ filter_to <- round(max(time_triggers))
 sens_prop_redist_outbreak_rates <- sens_prop_redist %>% 
   filter(`[step]` >= filter_from & `[step]` <= filter_to) %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
[…]

### 2025-10-07 scripts/sensitivity/proportion_redistributed_perfect_clean.R

@@ -27,7 +27,7 @@ consts <- c(consts, "outbreak?" = TRUE, "infection-control?" = TRUE)
 estimated_consts_mean <- abc_params_outbreak_control$unadj.values %>%
   apply(., 2, mean)
 
-estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-proportion-redistributed"]
+estimated_consts_mean <- estimated_consts_mean[names(estimated_consts_mean) != "c-side-room-delay-mean"]
 
 # fix to perfect clean
 estimated_consts_mean["c-toilet-cleaning-effect"] <- 1
@@ -49,8 +49,8 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-proportion-redistributed" = list(min=0.01,
-                                                                                 max=0.99,
+                            variables = list("c-side-room-delay-mean" = list(min=0.01,
+                                                                                 max=7,
                                                                                  qfun='qunif')
                             )
 )
@@ -83,7 +83,7 @@ sens_prop_redist_perfect_clean <- sens_prop_redist_perfect_clean %>%
 
 sens_prop_redist_perfect_clean_rates <- sens_prop_redist_perfect_clean %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
-  group_by(`random-seed`, `c-proportion-redistributed`, year, month) %>% 
+  group_by(`random-seed`, `c-side-room-delay-mean`, year, month) %>% 
   summarise(rate = (max(`total-hospital-infections`) - min(`total-hospital-infections`)) / sum(`current-inpatients`) * 1000) %>% 
   mutate(date_sim = ymd(paste0(year, "-", month, "-01"))) %>% 
   ungroup() %>% 
@@ -99,19 +99,19 @@ filter_to <- round(max(time_triggers))
 sens_prop_redist_perfect_clean_outbreak_rates <- sens_prop_redist_perfect_clean %>% 
   filter(`[step]` >= filter_from & `[step]` <= filter_to) %>% 
   mutate(year = year(date_sim), month = month(date_sim)) %>% 
[…]

### 2025-10-07 scripts/tune/outbreak_and_control.R

@@ -48,8 +48,8 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                            "o-antibiotic-prescription-rate" = list(min=antibiotic_prescription_rate,
                                                                    max=antibiotic_prescription_rate + 0.5 * antibiotic_prescription_rate,
                                                                    qfun='qunif'),
-                           "o-proportion-redistributed" = list(min=0.01,
-                                                               max=consts[["proportion-redistributed"]],
+                           "o-side-room-delay-mean" = list(min=0.01,
+                                                               max=7,
                                                                qfun='qunif'),
                            "c-toilet-cleaning-effect" = list(min=consts[["toilet-cleaning-effect"]],
                                                              max=0.99,
@@ -60,8 +60,8 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                            "c-antibiotic-prescription-rate" = list(min=antibiotic_prescription_rate - 0.5 * antibiotic_prescription_rate,
                                                                    max=antibiotic_prescription_rate,
                                                                    qfun='qunif'),
-                           "c-proportion-redistributed" = list(min=consts[["proportion-redistributed"]],
-                                                               max=0.99,
+                           "c-side-room-delay-mean" = list(min=0.01,
+                                                               max=7,
                                                                qfun='qunif')
 )
 

### 2025-10-07 scripts/tune/pre_outbreak.R

@@ -27,7 +27,7 @@ nl@experiment <- experiment(expname = "baseline_rate",
                                              "antibiotic-effect" = list(min=1.31, max=1.87, qfun='qunif'),
                                              "toilet-cleaning-rate" = list(min=0.5, max=3, qfun='qunif'),
                                              "random-colonisation" = list(min=2.8, max=12.1, qfun='qunif'),
-                                             "proportion-redistributed" = list(min=0.01, max=0.99, qfun='qunif')
+                                             "side-room-delay-mean" = list(min=0.01, max=7, qfun='qunif')
                             ),
                             constants = list("wards-total" = wards_total,
                                              "outbreak?" = FALSE,

### 2025-10-07 scripts/tune/variable_pre_outbreak.R

@@ -73,8 +73,8 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                            "o-antibiotic-prescription-rate" = list(min=antibiotic_prescription_rate,
                                                                    max=min(0.99, antibiotic_prescription_rate + upper_bound_scale * antibiotic_prescription_rate), # also from Vesporten 2018
                                                                    qfun='qunif'),
-                           "o-proportion-redistributed" = list(min=0.01,
-                                                               max=pre_outbreak_variables[["proportion-redistributed"]]$max,
+                           "o-side-room-delay-mean" = list(min=0.01,
+                                                               max=7,
                                                                qfun='qunif'),
                            "c-toilet-cleaning-effect" = list(min=pre_outbreak_variables[["toilet-cleaning-effect"]]$min,
                                                              max=0.99,
@@ -85,8 +85,8 @@ outbreak_variables <- list("outbreak-start" = list(min=outbreak_start - 90,
                            "c-antibiotic-prescription-rate" = list(min=max(0.01, antibiotic_prescription_rate - lower_bound_scale * antibiotic_prescription_rate),
                                                                    max=antibiotic_prescription_rate,
                                                                    qfun='qunif'),
-                           "c-proportion-redistributed" = list(min=pre_outbreak_variables[["proportion-redistributed"]]$min,
-                                                               max=0.99,
+                           "c-side-room-delay-mean" = list(min=0.01,
+                                                               max=7,
                                                                qfun='qunif')
 )
 

### 2025-10-07 paper/paper.qmd
paper updates

@@
 
-```{r}
-stochastic_sd <- read_csv("../out/sd_stochastic experiment-table.csv", skip = 6)
-```
-
-```{r}
+```{r load-stochastic-sd-model, include=FALSE}
 Rcpp::sourceCpp("../models/sd_stochastic.cpp")
@@
 
-```{r}
+```{r run-stochastic-sd-model, include=FALSE, cache=TRUE}
 set.seed(42)
@@
 ```
+
 ```{r fig-abc-sd-posteriors, include=TRUE}
@@
 
-```{r fig-abc-sd-posterior-predictive-check, include=TRUE}
+```{r fig-abc-sd-posterior-predictive-check, include=FALSE, eval=FALSE}
 #| echo: false
@@
 The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
-
-posterior_predictive_sim_sd_rates %>% 
-  pivot_wider(names_from = date_sim, values_from = rate) %>% 
-  pivot_longer(cols = !c(`random-seed`, `param_row`), names_to = "date_sim", values_to = "rate") %>% 
-  mutate(date_sim = ymd(date_sim)) %>% 
-  group_by(date_sim) %>%
-  summarise(m = median(rate),
-            q_high = quantile(rate, 0.75),
-            q_low = quantile(rate, 0.25),
-            .groups = "keep") %>%
-  ggplot() +
-  geom_line(aes(x = date_sim, y = m)) +
-  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
-  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
-  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
-  guides(color="none") +
-  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
 

### 2025-10-09 paper/paper.qmd
hid non stochastic figures

@@
 Antimicrobial prescription rate represents the percentage of patients that were administered an antimicrobial during their admission.
-Proportion redistributed represents the proportion of infected patients that were moved to a single-occupancy room (lower values represent fewer patients being moved).
+Side room delay is the mean number of days an infected patient remained in a bay before being moved to a side room.
 Toilet cleaning effect represents the effectiveness of cleaning in reducing toilet contamination (100% = complete removal of contamination; 0% = no cleaning effect).
@@
       param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
-      param == "c-side-room-delay-mean"     ~ "Side room delay mean",
-      param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate",
+      param == "c-side-room-delay-mean"     ~ "Side room delay mean (days)",
+      param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate (per day)",
       param == "c-toilet-cleaning-effect"       ~ "Toilet cleaning effect",
@@
 
-```{r run-deterministic-model, include=FALSE, cache=TRUE}
+```{r run-deterministic-model, eval=FALSE, include=FALSE, cache=TRUE}
 # Parameters and time-dependent functions
@@
 
-```{r run-stochastic-sd-model, include=FALSE, cache=TRUE}
+```{r run-stochastic-sd-model, include=FALSE, cache=TRUE, cache.lazy=FALSE}
 set.seed(42)
@@
 abc_sd <- abc(salgado$rates,
-              sd_param_grid, dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')
+              dplyr::select(sd_param_grid, -param_row, n_days),
+              dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')
 
@@
   pmap_dfr(posterior_check_params,
-        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row, seed) {
+        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, seed) {
             set.seed(seed)
             sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
-              mutate(`random-seed` = seed,
-                     param_row = param_row) %>% 
+              mutate(`random-seed` = seed) %>% 
               bind_rows()
@@
 
-```{r fig-single-outbreak-deterministic-one-plot, include=FALSE}
+```{r fig-single-outbreak-deterministic-one-plot, include=FALSE, eval=FALSE}
 #| echo: false
@@
 
-```{r fig-single-outbreak-deterministic-faceted, include=TRUE}
+```{r fig-single-outbreak-deterministic-faceted, include=FALSE, eval=FALSE}
 #| echo: false
@@
 The red line shows the prior distribution; the blue line shows the posterior distribution."
+
+sd_posterior_params_abc <- abc_sd$unadj.values
+colnames(sd_posterior_params_abc) <- names(dplyr::select(sd_param_grid, -param_row))
+sd_posterior_params_abc %>% 
+  as.data.frame() %>% 
+  dplyr::select(r, baseline_beta, baseline_p, outbreak_beta, outbreak_p) %>% 
+  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>% 
+  mutate(parameter = factor(parameter, 
+                            levels = c("r", "baseline_beta", "outbreak_beta", "baseline_p", "outbreak_p"))) %>% 
+  ggplot(aes(x = value)) +
+  geom_density(fill = "blue", alpha = 0.5) +
+  facet_wrap(~ parameter, scales = "free")
+
 ```

### 2025-10-09 paper/paper.qmd
sd section in paper

@@
     longname: interquartile range
+  - shortname: FIFO
+    longname: first-in, first-out
   insert_loa: false
@@
 
-Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained colonised until discharge.
-At each time increment, the probability of a non-colonised/infected patient acquiring infection (i.e., transitioning from a non-infected/colonised to infected/colonised state) depended on: 1) a baseline risk of *de novo* infection/colonisation:
+Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge.
+At each time increment, the probability of a non-infected patient acquiring infection (i.e., transitioning from a non-infected to infected state) depended on: 1) a baseline risk of *de novo* infection:
 
@@
 
-## Deterministic model
-
-A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
+During the simulation level of contamination of each toilet increased when an infected patient used it, and decreased
+when the toilet was cleaned.
+
+Infected patients were moved to single-occupancy rooms (side rooms) after infection.
+To capture delays in moving patients into side rooms (which can be due to delays in diagnosis, and/or lack of available side rooms), the simulation included an optional parameter that samples the delay from a Poisson distribution with a configurable mean (in days).
+Once the infected patient has completed their delay, they became eligible to move to the next available side room.
+To implement this, the simulation used a \acr{FIFO} queue to track infected patients waiting for a side room.
+While direct patient swaps were possible (i.e., swapping a non-infected patient in a side room with an infected patient waiting for a side room), more complex swaps involving multiple patients were not implemented.
+
+## Stochastic system dynamics model
+
+To build a high-level system dynamics model, a deterministic model was firstly described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
 
@@
 -   $S$ is the number of susceptible inpatients,
--   $I$ is the number of infected/colonised inpatients,
+-   $I$ is the number of infected inpatients,
 -   $A$ is the rate of new admissions,
--   $p$ is the community colonisatio rate,
+-   $p$ is the community colonisation rate,
 -   $\beta$ is the transmission coefficient,
--   $r$ is the rate of random (*de novo*) colonisation/infection,
+-   $r$ is the rate of random (*de novo*) infection,
 -   $D$ is the rate of patient discharges.
@@
 
-![System dynamics model](images/sd_model.svg)
+![System dynamics model](images/sd_model.svg){#fig-sd-model}
+
+To add stochasticity to the model, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:
+
+$$
+D_{\text{total}, t} \sim \text{Poisson}(D),
+$$
+
+where $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.
+Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:
+
+$$
+D_{S,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \Bigr \rceil,
+$$
+
+and
+
+$$
+D_{I,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \Bigr \rceil = D_{\text{total}, t} - D_{S,t},,
+$$
+
+where $\lfloor x \rceil$ is the rounding of $x$ to the nearest integer.
+
+Next, new admissions were calculated. Since $A_{\text{total}, t} = D_{\text{total}, t}$,
+
+$$
+A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, p),
+$$
+
+and
+
+$$
+A_{S,t} = A_{\text{total}, t} - A_{I,t},
+$$
+
+To calculate the number of hospital-acquired infections (i.e., $\beta S_t + r S_t$), the binomial distribution was used:
+
+$$
+\text{HAI}_{\text{transmission}, t} \sim \text{Binomial}(S_t - D_{S, t}, \beta),
+$$
+
+and
+
+$$
+\text{HAI}_{\text{de novo}, t} \sim \text{Binomial}(S_t - D_{S, t} - \text{HAI}_{\text{transmission}, t}, r),
+$$
+
+where $\text{HAI}_{\text{transmission}, t}$ is the number of hospital-acquired infections due to transmission at time $t$, and $\text{HAI}_{\text{de novo}, t}$ is the number of hospital-acquired infections due to *de novo* infection at time $t$.
+
+Finally, the number of susceptible and infected patients at time $t + 1$ was updated as:
+
+$$
+S_{t+1} = S_t + A_{S,t} - D_{S,t} - \text{HAI}_{\text{transmission}, t} - \text{HAI}_{\text{de novo}, t},
+$$
+
+and
+
+$$
+I_{t+1} = I_t + A_{I,t} - D_{I,t} + \text{HAI}_{\text{transmission}, t} + \text{HAI}_{\text{de novo}, t}.
+$$
 
@@
 To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
+
+A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the exception that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak.
+The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate).
+The parameter $r$ (rate of random infection) was kept constant throughout the simulation.
 
@@
 To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
-Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) proportion of patients redistributed to a single-occupancy bed after infection; 3) toilet cleaning rate; and 4) toilet cleaning effect.
+Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect.
 To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value.
@@
       # LHS is evaluated against the facet layout; RHS is the scale to use for matching facets
-      param %in% c("Antimicrobial prescription rate", "Proportion redistributed", "Toilet cleaning effect") ~
+      param %in% c("Antimicrobial prescription rate", "Toilet cleaning effect") ~
         scale_x_continuous(

### 2025-10-09 paper/paper.qmd
describe models in results

@@
     longname: first-in, first-out
+  - shortname: SD
+    longname: system dynamics
   insert_loa: false
@@
 
-![System dynamics model](images/sd_model.svg){#fig-sd-model}
+```{r fig-sd-model, include=TRUE}
+#| echo: false
+#| fig.cap: "System dynamics model.
+$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community colonisation rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
+knitr::include_graphics("./images/sd_model.svg")
+```
 
@@
 
-A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the exception that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak.
+A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak.
 The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate).
@@
 
+```{r fig-screenshot, include=FALSE}
+#| echo: false
+#| out.height: "7in"
+#| fig.cap: "Screenshot of model graphical user interface during a simulation run"
+knitr::include_graphics("./images/screenshot.png")
+```
+
 ## Data sources
@@
 
-Rejection sampling \acr{ABC} was employed to fit the model to the outbreak data [@sunnakerApproximateBayesianComputation2013].
+### Calibration of \acr[first_use=TRUE]{ABM}
+
+Rejection sampling \acr{ABC} was employed to fit the \acr{ABM} to the observed outbreak data [@sunnakerApproximateBayesianComputation2013].
 Priors were extracted from existing literature; flat priors when data were unavailable.
@@
 The proportion of toilets that were shared between multiple patients was set to 60%.
-
-Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$.
-Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
-The number of daily cases of nosocomial \acr{CDI} were extracted from the model outputs, and converted to a time series of nosocomial \acr{CDI} per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
-Since each simulation starts with an empty hospital that is gradually populated with patient admissions, results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.
-
-Each simulated summary statistic $S_i$ was compared to the summary statistic generated from the observed data, $S_\text{obs}$, and $S_i$ was accepted if:
-
-$$
-d(S_i, S_\text{obs}) \leq \epsilon,
-$$
-
-where:
-
--   $d$ is the Euclidean distance between $S_i$ and $S_\text{obs}$, and
-
--   $\epsilon$ is the tolerance.
-
-Otherwise, $S_i$ was rejected.
-The parameter samples that generated the accepted simulations were used to produce a posterior distribution.
-$\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.
-
-A two-step approach to calibration was taken.
-Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate.
-Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
-To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
-The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
-To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded.
-At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
-The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.
-Again, only samples from the latin hypercube that satisfied these requirements were used.
-
-As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates.
-Three approaches to sampling from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were used:
-
-1. Parameter samples drawn from a uniform distribution:
-
-$$
-U_[a,b] = U(\mu - \sigma, \mu + \sigma),
-$$
-
-where: $\mu$ and $\sigma$ are the mean and standard deviation of the posterior distribution, respectively;
-
-2. The mean of each parameter's posterior distribution; and
-3. Samples with replacement taken from the posterior distribution.
-
-```{r text-seeds, include=FALSE}
-results_outbreak_control_n_seeds <- length(unique(results_outbreak_control_pre_included$`random-seed`))
-uniform_sims_n_seeds <- length(unique(uniform_sims$`random-seed`))
-mean_sims_n_seeds <- length(unique(mean_sims$`random-seed`))
-posterior_predictive_n_seeds <- length(unique(posterior_predictive_sim$`seed-value`))
-```
-
-Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated.
-For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
-for the posterior predictive checks, `r uniform_sims_n_seeds` seeds were used for the uniform distribution,
-`r mean_sims_n_seeds` seeds were used for the mean of the posterior distribution, and
-`r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.
-
-One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.
 
@@
 
-![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="7in"}
-
-Users can run the model interactively using the interface shown in @fig-screenshot.
+Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$.
+Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
+The number of daily cases of nosocomial \acr{CDI} were extracted from the model outputs, and converted to a time series of nosocomial \acr{CDI} per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
+Since each simulation starts with an empty hospital that is gradually populated with patient admissions, results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.
+
+Each simulated summary statistic $S_i$ was compared to the summary statistic generated from the observed data, $S_\text{obs}$, and $S_i$ was accepted if:
+
+$$
+d(S_i, S_\text{obs}) \leq \epsilon,
+$$
+
+where:
+
+-   $d$ is the Euclidean distance between $S_i$ and $S_\text{obs}$, and
+
+-   $\epsilon$ is the tolerance.
+
+Otherwise, $S_i$ was rejected.
+The parameter samples that generated the accepted simulations were used to produce a posterior distribution.
+$\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.
+
+A two-step approach to calibration was taken.
+Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate.
+Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
+To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
+The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
+To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded.
+At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
+The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.
+Again, only samples from the latin hypercube that satisfied these requirements were used.
+
+As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates.
+To sample from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were samples with replacement were taken and used as an input to the stochastic \acr{SD} model.
+
+```{r text-seeds, include=FALSE}
+results_outbreak_control_n_seeds <- length(unique(results_outbreak_control_pre_included$`random-seed`))
+uniform_sims_n_seeds <- length(unique(uniform_sims$`random-seed`))
+mean_sims_n_seeds <- length(unique(mean_sims$`random-seed`))
+posterior_predictive_n_seeds <- length(unique(posterior_predictive_sim$`seed-value`))
+```
+
+Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated.
+For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
+for the posterior predictive checks, `r uniform_sims_n_seeds` seeds were used for the uniform distribution,
+`r mean_sims_n_seeds` seeds were used for the mean of the posterior distribution, and
+`r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.
+
+### Calibration of the \acr[first_use=TRUE]{SD} model
+
+\acr[first_use=TRUE, case=sentence]{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).
 
@@
 
-To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
+To assess the sensitivity of the calibrated \acr{ABM} to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
 Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect.
@@
 
-An \acr{ABM} that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced.
+Both the \acr{ABM} and the \acr{SD} model were able to simulate the faecal-oral transmission of infection in a hospital setting.
+Although the \acr{SD} was computationally simpler, and produced simulated infection rates that were very similar to the observed rates (see @fig-posterior-predictive-check), the \acr{ABM} was based on a simulation that more closely resembled the workings of a real hospital.
+Importantly, the parameters of the \acr{ABM} were directly based on modifiable infection control practices, such as toilet cleaning frequency and antimicrobial prescription rates.
+Therefore, the main analyses and senstivity analyses were performed using the \acr{ABM}.
+
 After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control.
@@
 
-```{r fig-consts-sims-mean-sd, include=TRUE}
+```{r fig-consts-sims-mean-sd, include=FALSE}
 #| echo: false
@@
 
-```{r fig-sensitvity-pre-outbreak-abx, include=TRUE}
+```{r fig-sensitvity-pre-outbreak-abx, include=FALSE}
 #| echo: false
@@
 
+```{r}
+# compare ABM community colonisation rate to SD model p
+mean_sims %>% 
+  summarise(mean = mean(`community-colonisation-rate`),
+            sd = sd(`community-colonisation-rate`))
+```
+
 # Discussion

### 2025-10-26 scripts/helpers/simdesign_random_helpers.R
simdesign_random

@@ -1,0 +1,66 @@
+#' Generate a random sample of input factor sets
+#'
+#' @description Generate a random sample of input factor sets
+#'
+#' @param input list with variables and value ranges
+#' @param samples number of random samples
+#' @param precision number of digits for the decimal fraction of parameter
+#' values
+#' @aliases util_create_random
+#' @rdname util_create_random
+#' @keywords internal
+util_create_random <- function(input, samples, precision) {
+
+  # create a random sample of input factor sets (independent draws from priors)
+  rnd.design <- lapply(seq(1, length(input)), function(i) {
+    spec <- input[[i]]
+    qfun <- spec[["qfun"]]
+    if (is.null(qfun)) qfun <- "qunif"
+    args <- spec
+    args[["qfun"]] <- NULL
+    probs <- stats::runif(samples)
+    do.call(match.fun(qfun), c(list(p = probs), args))
+  })
+  names(rnd.design) <- names(input)
+  rnd.final <- tibble::as_tibble(rnd.design)
+  ## Precision:
+  rnd.final <- round(rnd.final, digits = precision)
+
+  return(rnd.final)
+}
+
+## simdesign_random: like simdesign_lhs but draws iid random samples from priors
+## - nl: nlrx nl object with a defined experiment
+## - samples: number of random parameter sets
+## - nseeds: number of seeds for this simulation design
+## - precision: number of digits for the decimal fraction of parameter values
+simdesign_random <- function(nl, samples, nseeds, precision) {
+  # Evaluate nl object and variables (mirror simdesign_lhs behavior when available)
+  nlrx:::util_eval_experiment(nl)
+  nlrx:::util_eval_variables(nl)
+  nlrx:::util_eval_variables_sa(nl)
+  message("Creating random simulation design")
+
+  rnd <- util_create_random(input = nlrx::getexp(nl, "variables"),
+                             samples = samples,
+                             precision = precision)
+
+  ## Bind constants if any:
+  if(length(nlrx::getexp(nl, "constants")) > 0)
+  {
+    rnd <- cbind(rnd, nlrx::getexp(nl, "constants"), stringsAsFactors=FALSE)
+  }
+
+  ## Convert to tibble:
+  rnd <- tibble::as_tibble(rnd)
+
+  ## Generate seeds
+  seeds <- nlrx:::util_generate_seeds(nseeds = nseeds)
+
+  # Add simdesign to nl
+  new_simdesign <- nlrx::simdesign(simmethod="random",
[…]

### 2025-10-26 scripts/helpers/test_simdesign_random.R

@@ -1,0 +1,62 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
+# Test simdesign_random with a simple example similar to simdesign_lhs
+
+# Setup nl object (using dummy paths for test)
+nl <- nlrx::nl(nlversion = "6.2.2",
+               nlpath = ".",
+               modelpath = "dummy.nlogo")
+
+# Define experiment (adapted from nlrx example)
+nl@experiment <- nlrx::experiment(expname = "wolf-sheep-random",
+                                  outpath = "/tmp/out/",
+                                  repetition = 1,
+                                  tickmetrics = "true",
+                                  idsetup = "setup",
+                                  idgo = "go",
+                                  idfinal = NA_character_,
+                                  idrunnum = NA_character_,
+                                  runtime = 50,
+                                  evalticks = seq(40, 50),
+                                  metrics = c("count sheep", "count wolves", "count patches with [pcolor = green]"),
+                                  variables = list('initial-number-sheep' = list(min = 50, max = 150, qfun = "qunif"),
+                                                   'initial-number-wolves' = list(min = 50, max = 150, qfun = "qunif")),
+                                  constants = list("model-version" = "\"sheep-wolves-grass\"",
+                                                   "grass-regrowth-time" = 30,
+                                                   "sheep-gain-from-food" = 4,
+                                                   "wolf-gain-from-food" = 20,
+                                                   "sheep-reproduce" = 4,
+                                                   "wolf-reproduce" = 5,
+                                                   "show-energy?" = "false"))
+
+# Attach simdesign_random
+nl@simdesign <- simdesign_random(nl = nl,
+                                 samples = 10,  # Smaller for test
+                                 nseeds = 3,
+                                 precision = 3)
+
+# Print summary
+print("Simdesign random created successfully")
+print(paste("Simmethod:", nl@simdesign@simmethod))
+print(paste("Siminput dimensions:", paste(dim(nl@simdesign@siminput), collapse = " x ")))
+print(paste("Number of seeds:", length(nl@simdesign@simseeds)))
+print("First few rows of siminput:")
+print(head(nl@simdesign@siminput))
+print("Seeds:")
+print(nl@simdesign@simseeds)
+
+# Now, run using lhs
+nl@simdesign <- nlrx::simdesign_lhs(nl = nl,
+                                    samples = 10,
[…]

### 2025-10-26 scripts/tune/pre_outbreak.R

@@ -1,10 +1,12 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
 sim_days <- (365 * 3) + 30
 # sim_hours <- sim_days * 24
 
 nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
-         jvmmem = 1024)
+         jvvmem = 1024)
 
 nl@experiment <- experiment(expname = "baseline_rate",
                             outpath = out_path,
@@ -37,10 +39,14 @@ nl@experiment <- experiment(expname = "baseline_rate",
                                              "admission-days" = admission_days,
                                              "bay-proportion" = bay_proportion))
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples,
-                              nseeds = calibration_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_random(nl,
+                                 samples = calibration_samples,
+                                 nseeds = calibration_seeds,
+                                 precision = 3)
+# Note: simdesign_random draws independent random samples from the prior distributions
+# defined in nl@experiment@variables, unlike simdesign_lhs which uses Latin Hypercube Sampling.
+# Seeds are generated using nlrx:::util_generate_seeds for reproducibility.
+# To control randomness, set a seed before calling this function.
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/sensitivity/abx_prescription_rates.R
random or discrete

@@ -36,16 +36,15 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-antibiotic-prescription-rate" = list(min=0.01,
-                                                                                     max=0.99,
-                                                                                     qfun='qunif')
+                            variables = list(
+                              "c-antibiotic-prescription-rate" = list(
+                                values = seq(from = 0.01, to = 0.99, length.out = sens_samples)
+                              )
                             )
 )
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = sens_samples,
-                              nseeds = sens_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_distinct(nl,
+                                   nseeds = sens_seeds)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/sensitivity/proportion_redistributed.R

@@ -36,16 +36,15 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-side-room-delay-mean" = list(min=0.01,
-                                                                                 max=7,
-                                                                                 qfun='qunif')
+                            variables = list(
+                              "c-side-room-delay-mean" = list(
+                                values = seq(from = 0.01, to = 7, length.out = sens_samples)
+                              )
                             )
 )
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = sens_samples,
-                              nseeds = sens_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_distinct(nl,
+                                   nseeds = sens_seeds)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/sensitivity/proportion_redistributed_perfect_clean.R

@@ -49,16 +49,15 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-side-room-delay-mean" = list(min=0.01,
-                                                                                 max=7,
-                                                                                 qfun='qunif')
+                            variables = list(
+                              "c-side-room-delay-mean" = list(
+                                values = seq(from = 0.01, to = 7, length.out = sens_samples)
+                              )
                             )
 )
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = sens_samples,
-                              nseeds = sens_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_distinct(nl,
+                                   nseeds = sens_seeds)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/sensitivity/toilet_cleaning_effect.R

@@ -36,16 +36,15 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-toilet-cleaning-effect" = list(min=0.01,
-                                                                               max=0.99,
-                                                                               qfun='qunif')
+                            variables = list(
+                              "c-toilet-cleaning-effect" = list(
+                                values = seq(from = 0.01, to = 0.99, length.out = sens_samples)
+                              )
                             )
 )
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = sens_samples,
-                              nseeds = sens_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_distinct(nl,
+                                   nseeds = sens_seeds)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/sensitivity/toilet_cleaning_rate.R

@@ -36,16 +36,15 @@ nl@experiment <- experiment(expname = "sens_analysis",
                                         "current-inpatients",
                                         "current-colonised"),
                             constants = c(consts, estimated_consts_mean), 
-                            variables = list("c-toilet-cleaning-rate" = list(min=0.25,
-                                                                             max=4,
-                                                                             qfun='qunif')
+                            variables = list(
+                              "c-toilet-cleaning-rate" = list(
+                                values = seq(from = 0.25, to = 4, length.out = sens_samples)
+                              )
                             )
 )
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = sens_samples,
-                              nseeds = sens_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_distinct(nl,
+                                   nseeds = sens_seeds)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/tune/fine_tune_pre_outbreak.R

@@ -1,3 +1,5 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
 sim_days <- (365 * 3) + 30
 
 nl <- nl(nlversion = nl_version,
@@ -40,10 +42,10 @@ nl@experiment <- experiment(expname = "fine_tune",
                                              "admission-days" = admission_days,
                                              "bay-proportion" = bay_proportion))
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples,
-                              nseeds = calibration_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_random(nl,
+                                 samples = calibration_samples,
+                                 nseeds = calibration_seeds,
+                                 precision = 3)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/tune/outbreak_and_control.R

@@ -1,3 +1,5 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
 sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado_v0.3.nlogo"
@@ -83,10 +85,10 @@ nl@experiment <- experiment(expname = "outbreak_control",
                             variables = outbreak_variables,
                             constants = consts)
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples,
-                              nseeds = calibration_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_random(nl,
+                                 samples = calibration_samples,
+                                 nseeds = calibration_seeds,
+                                 precision = 3)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/tune/simulate_uniform.R

@@ -1,3 +1,5 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
 sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado.nlogo"
@@ -47,10 +49,10 @@ nl@experiment <- experiment(expname = "uniform_sims",
                             constants = consts, 
                             variables = estimated_consts)
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples,
-                              nseeds = calibration_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_random(nl,
+                                 samples = calibration_samples,
+                                 nseeds = calibration_seeds,
+                                 precision = 3)
 
 plan(list(sequential, multisession))
 

### 2025-10-26 scripts/tune/variable_pre_outbreak.R

@@ -1,3 +1,5 @@
+source("scripts/helpers/simdesign_random_helpers.R")
+
 sim_days <- (365 * 7) + 30
 
 # outbreak_control_model <- "salgado.nlogo"
@@ -109,10 +111,10 @@ nl@experiment <- experiment(expname = "outbreak_control",
                             variables = outbreak_variables,
                             constants = consts)
 
-nl@simdesign <- simdesign_lhs(nl,
-                              samples = calibration_samples * 4,
-                              nseeds = calibration_seeds,
-                              precision = 3)
+nl@simdesign <- simdesign_random(nl,
+                                 samples = calibration_samples * 4,
+                                 nseeds = calibration_seeds,
+                                 precision = 3)
 
 # keep simulated parameters that fit with outbreak generation
 nl@simdesign@siminput <- nl@simdesign@siminput %>%

### 2025-10-27 scripts/calibrate.R
globally set multi and simdesign

@@ -7,6 +7,17 @@ library(progressr)
 library(here)
 handlers("cli")
 options(progressr.enable=TRUE)
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
+
+# optionally, set jvm here:
+Sys.setenv(JAVA_HOME = "/usr/lib/jvm/java-1.17.0-openjdk-amd64/")
+
+# set parameter sampling functions
+simdesign_fn <- simdesign_random
+
+# set future plan (multicore or multisession) and workers
+future_plan <- "multisession"
+num_workers <- 4
 
 wards_total <- 9
 bedspaces_per_ward <- 9

### 2025-10-27 scripts/sensitivity/abx_prescription_rates.R

@@ -46,7 +46,7 @@ nl@experiment <- experiment(expname = "sens_analysis",
 nl@simdesign <- simdesign_distinct(nl,
                                    nseeds = sens_seeds)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   sens_abx_prescription <- progressr::with_progress(

### 2025-10-27 scripts/sensitivity/proportion_redistributed.R

@@ -46,7 +46,7 @@ nl@experiment <- experiment(expname = "sens_analysis",
 nl@simdesign <- simdesign_distinct(nl,
                                    nseeds = sens_seeds)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   sens_prop_redist <- progressr::with_progress(

### 2025-10-27 scripts/sensitivity/proportion_redistributed_perfect_clean.R

@@ -59,7 +59,7 @@ nl@experiment <- experiment(expname = "sens_analysis",
 nl@simdesign <- simdesign_distinct(nl,
                                    nseeds = sens_seeds)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   sens_prop_redist_perfect_clean <- progressr::with_progress(

### 2025-10-27 scripts/sensitivity/toilet_cleaning_effect.R

@@ -46,7 +46,7 @@ nl@experiment <- experiment(expname = "sens_analysis",
 nl@simdesign <- simdesign_distinct(nl,
                                    nseeds = sens_seeds)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   sens_toilet_cleaning_effect <- progressr::with_progress(

### 2025-10-27 scripts/sensitivity/toilet_cleaning_rate.R

@@ -46,7 +46,7 @@ nl@experiment <- experiment(expname = "sens_analysis",
 nl@simdesign <- simdesign_distinct(nl,
                                    nseeds = sens_seeds)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   sens_toilet_cleaning_rate <- progressr::with_progress(

### 2025-10-27 scripts/tune/fine_tune_pre_outbreak.R

@@ -1,4 +1,4 @@
-source("scripts/helpers/simdesign_random_helpers.R")
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
 
 sim_days <- (365 * 3) + 30
 
@@ -42,12 +42,12 @@ nl@experiment <- experiment(expname = "fine_tune",
                                              "admission-days" = admission_days,
                                              "bay-proportion" = bay_proportion))
 
-nl@simdesign <- simdesign_random(nl,
+nl@simdesign <- simdesign_fn(nl,
                                  samples = calibration_samples,
                                  nseeds = calibration_seeds,
                                  precision = 3)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   results_baseline_fine_tune <- progressr::with_progress(

### 2025-10-27 scripts/tune/outbreak_and_control.R

@@ -1,4 +1,4 @@
-source("scripts/helpers/simdesign_random_helpers.R")
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
 
 sim_days <- (365 * 7) + 30
 
@@ -85,12 +85,12 @@ nl@experiment <- experiment(expname = "outbreak_control",
                             variables = outbreak_variables,
                             constants = consts)
 
-nl@simdesign <- simdesign_random(nl,
+nl@simdesign <- simdesign_fn(nl,
                                  samples = calibration_samples,
                                  nseeds = calibration_seeds,
                                  precision = 3)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   results_outbreak_control <- progressr::with_progress(

### 2025-10-27 scripts/tune/posterior_predictive_checks.R

@@ -61,7 +61,7 @@ nl@experiment <- experiment(expname = "posterior_predictive_sim",
 nl@simdesign <- simdesign_distinct(nl,
                               nseeds = 1)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   posterior_predictive_sim <- progressr::with_progress(

### 2025-10-27 scripts/tune/pre_outbreak.R

@@ -1,4 +1,4 @@
-source("scripts/helpers/simdesign_random_helpers.R")
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
 
 sim_days <- (365 * 3) + 30
 # sim_hours <- sim_days * 24
@@ -6,7 +6,7 @@ sim_days <- (365 * 3) + 30
 nl <- nl(nlversion = nl_version,
          nlpath = netlogo_path,
          modelpath = model_path,
-         jvvmem = 1024)
+         jvmmem = 1024)
 
 nl@experiment <- experiment(expname = "baseline_rate",
                             outpath = out_path,
@@ -39,16 +39,16 @@ nl@experiment <- experiment(expname = "baseline_rate",
                                              "admission-days" = admission_days,
                                              "bay-proportion" = bay_proportion))
 
-nl@simdesign <- simdesign_random(nl,
+nl@simdesign <- simdesign_fn(nl,
                                  samples = calibration_samples,
                                  nseeds = calibration_seeds,
                                  precision = 3)
-# Note: simdesign_random draws independent random samples from the prior distributions
-# defined in nl@experiment@variables, unlike simdesign_lhs which uses Latin Hypercube Sampling.
+# Note: simdesign_fn draws independent random samples from the prior distributions
+# defined in nl@experiment@variables, unlike simdesign_fn which uses Latin Hypercube Sampling.
 # Seeds are generated using nlrx:::util_generate_seeds for reproducibility.
 # To control randomness, set a seed before calling this function.
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   results_baseline <- progressr::with_progress(
@@ -111,7 +111,7 @@ abc_params <- abc(target = pre_outbreak_data$rates,
                   sumstat = sumstats,
                   tol = abc_tol,
                   method="rejection")
-
+print(abc_params)
 # plot
 long_retained_ss <- abc_params$ss %>% 
   as_tibble %>% 

### 2025-10-27 scripts/tune/simulate_uniform.R

@@ -1,4 +1,4 @@
-source("scripts/helpers/simdesign_random_helpers.R")
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
 
 sim_days <- (365 * 7) + 30
 
@@ -49,12 +49,12 @@ nl@experiment <- experiment(expname = "uniform_sims",
                             constants = consts, 
                             variables = estimated_consts)
 
-nl@simdesign <- simdesign_random(nl,
+nl@simdesign <- simdesign_fn(nl,
                                  samples = calibration_samples,
                                  nseeds = calibration_seeds,
                                  precision = 3)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   uniform_sims <- progressr::with_progress(

### 2025-10-27 scripts/tune/variable_pre_outbreak.R

@@ -1,4 +1,4 @@
-source("scripts/helpers/simdesign_random_helpers.R")
+source(here::here("scripts", "helpers", "simdesign_random_helpers.R"))
 
 sim_days <- (365 * 7) + 30
 
@@ -110,8 +110,8 @@ nl@experiment <- experiment(expname = "outbreak_control",
                                         "current-colonised"),
                             variables = outbreak_variables,
                             constants = consts)
-
-nl@simdesign <- simdesign_random(nl,
+print(outbreak_variables)
+nl@simdesign <- simdesign_fn(nl,
                                  samples = calibration_samples * 4,
                                  nseeds = calibration_seeds,
                                  precision = 3)
@@ -123,7 +123,7 @@ nl@simdesign@siminput <- nl@simdesign@siminput %>%
                   `o-community-colonisation-rate` >= `community-colonisation-rate`) %>% 
   sample_n(calibration_samples)
 
-plan(list(sequential, multisession))
+plan(list(sequential, future_plan), workers = num_workers)
 
 if (run_sims) {
   results_outbreak_control_pre_included <- progressr::with_progress(

### 2025-10-27 scripts/helpers/simdesign_random_helpers.R
patch unused args simdesign_random

@@ -16,10 +16,17 @@ util_create_random <- function(input, samples, precision) {
     spec <- input[[i]]
     qfun <- spec[["qfun"]]
     if (is.null(qfun)) qfun <- "qunif"
+    fun <- match.fun(qfun)
     args <- spec
     args[["qfun"]] <- NULL
+    # Keep only arguments accepted by the chosen quantile function.
+    # If the function supports '...', keep all; otherwise, filter.
+    formal_names <- tryCatch(names(formals(fun)), error = function(e) NULL)
+    if (!is.null(formal_names) && !("..." %in% formal_names)) {
+      args <- args[names(args) %in% formal_names]
+    }
     probs <- stats::runif(samples)
-    do.call(match.fun(qfun), c(list(p = probs), args))
+    do.call(fun, c(list(p = probs), args))
   })
   names(rnd.design) <- names(input)
   rnd.final <- tibble::as_tibble(rnd.design)

### 2025-10-27 paper/paper.qmd

@@
 ---
-title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
+title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
 authors:
@@
   insert_loa: false
+  insert_links: false
 ---
@@
 \acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings.
-Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
+Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected patients [@worldhealthorganizationGlobalReportInfection2022].
 A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
 In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
-These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
+These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
 The impact of these interventions is often difficult to quantify.
@@
 
-The use of computer simulation techniques such as \acr{ABM} can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
+The use of computer simulation techniques such as \acr{ABM} and can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
 Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021].
@@
 This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].
-
-Here, we describe an agent-based model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
+Another simulation approach is stochastic \acr{SD} modelling, which is a high-level approach that models the flow of patients between different states, such as susceptible and infected states, and is particularly useful for understanding feedback loops within systems.
+
+Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
 Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods.
 The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.
+We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach.
 
@@
 
--   $r_0$ is the baseline/*de novo* risk of infection/colonisation,
-
--   $m$ is the increased (or decreased) relative risk of infection/colonisation due to antimicrobial therapy,
+-   $r_0$ is the baseline/*de novo* risk of infection,
+
+-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy,
 
@@
 -   $A$ is the rate of new admissions,
--   $p$ is the community colonisation rate,
+-   $p$ is the community infection rate,
 -   $\beta$ is the transmission coefficient,
@@
 #| fig.cap: "System dynamics model.
-$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community colonisation rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
+$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
 knitr::include_graphics("./images/sd_model.svg")
@@
 $$
-D_{S,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \Bigr \rceil,
+D_{S,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \right \rceil,
 $$
@@
 $$
-D_{I,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \Bigr \rceil = D_{\text{total}, t} - D_{S,t},,
+D_{I,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \right \rceil = D_{\text{total}, t} - D_{S,t},,
 $$
@@
 A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak.
-The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate).
+The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community infection rate).
 The parameter $r$ (rate of random infection) was kept constant throughout the simulation.
@@
 
-Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community colonisation rate, antimicrobial prescription rate); and 3) colonisation parameters (e.g., relative risk increase of antimicrobial therapy on colonisation/infection).
+Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community infection rate, antimicrobial prescription rate); and 3) infection parameters (e.g., relative risk increase of antimicrobial therapy on infection).
 
@@
 | Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
-| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
+| Community infection rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
 | Antimicrobial effect | Uniform | $a = 1.31, b = 1.87$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
@@
 To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
-The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
+The priors used for ABC were selected to be compatible with increased transmission or infection rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
 To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded.
 At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
-The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.
+The prior distributions for these parameters were selected to decrease transmission and infection -- generally the reverse of the 'outbreak' parameters.
 Again, only samples from the latin hypercube that satisfied these requirements were used.
@@
 
-### Calibration of the \acr[first_use=TRUE]{SD} model
+### Calibration of \acr[first_use=TRUE]{SD} model
 
@@
 To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value.
-The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period.
+The output metric was the number of cases of hospital-acquired infection within the outbreak period.
 For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.
@@
                  "toilet-cleaning-effect" = "Toilet cleaning effect",
-                 "community-colonisation-rate" = "Community colonisation rate",
+                 "community-colonisation-rate" = "Community infection rate",
                  "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                  "antibiotic-effect" = "Antibiotic effect",
-                 "random-colonisation" = "Random colonisation",
+                 "random-colonisation" = "Random infection",
                  "side-room-delay-mean" = "Side room delay mean")
@@
   mutate(name = str_replace_all(name, "-", " ")) %>%
+  mutate(name = str_replace_all(name, "colonisation", "infection")) %>%
   mutate(name = str_to_title(name)) %>% 
@@
 The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate during the outbreak period.
-The cohorting of patients into single-occupancy rooms after colonisation/infection had an almost linear relationship with the number of cases.
+The cohorting of patients into single-occupancy rooms after infection had an almost linear relationship with the number of cases.
 
@@
 
-```{r}
-# compare ABM community colonisation rate to SD model p
-mean_sims %>% 
-  summarise(mean = mean(`community-colonisation-rate`),
-            sd = sd(`community-colonisation-rate`))
-```
-
 # Discussion
+
+_The discussion needs to explore the different modelling approaches used (SD vs ABM) and why we eventually settled on the ABM as the main model.
+This is sort of captured in the results, but maybe it needs to be here as well?_
 
@@
 
-The model was designed to simulate the spread of infection/colonisation at a baseline (or background) rate, followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate.
+The model was designed to simulate the spread of infection at a baseline (or background) rate, followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate.
 The model was calibrated to a report of a _C.
@@
 To simulate outbreaks and periods of enhanced infection control, the model was designed to change parameters at discrete time points.
-For example, the rate of antimicrobial prescriptions was higher during the outbreak period, promoting the spread of infection and colonisation.
+For example, the rate of antimicrobial prescriptions was higher during the outbreak period, promoting the spread of infection.
 Later, the prescription rate was reduced to help control the outbreak.
@@
 In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding.
-Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected or colonised.
+Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected.
 Antimicrobial class and duration of exposure are factors that are not accounted for in this model.

### 2025-10-28 scripts/calibrate.R
Update calibrate.R

@@ -18,7 +18,12 @@ model_path <- here::here("models", "hospital_transmission.nlogo")
 out_path <- here::here("out")
 data_path <- here::here("data")
 
-run_sims <- FALSE
+if (!exists("run_sims")) {
+  run_sims <- FALSE
+}
+
+# to force running sims, set run_sims <- TRUE here
+run_sims <- TRUE
 
 antibiotic_prescription_rate = 0.499 # Vesporten 2018
 admission_days <- 4.6

### 2025-10-28 paper/paper.qmd
out_random path in paper

@@
 # Alternatively, just load previously dumped .RData from a run of calibrate.R
-load(file.path("..", "out", "calibrate.RData"))
+load(file.path("..", "out_random", "calibrate.RData"))
 ```

### 2025-10-28 scripts/tune/pre_outbreak.R
removed stray print statement

@@ -111,7 +111,7 @@ abc_params <- abc(target = pre_outbreak_data$rates,
                   sumstat = sumstats,
                   tol = abc_tol,
                   method="rejection")
-print(abc_params)
+
 # plot
 long_retained_ss <- abc_params$ss %>% 
   as_tibble %>% 

### 2025-11-03 paper/paper.qmd
Pete comments

@@
 The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.
-We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach.
+We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach for this application.
 
@@
 $$
-x_p(t) \in \{0, 1\}
+x_i(t) \in \{0, 1\}
 $$
@@
 -   $t \in \mathbb{N}_0 = \{0, 1, 2, \ldots\}$ is the time step (in days),
--   $x_p(t) = 0$ if patient $p$ is not infected at time $t$,
--   $x_p(t) = 1$ if patient $p$ is infected at time $t$.
-
-Since the simulation's target diseases were associated with long term colonisation, and risk of onward transmission, after the initial infection event, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge.
-At each time increment, the probability of a non-infected patient acquiring infection (i.e., transitioning from a non-infected to infected state) depended on: 1) a baseline risk of *de novo* infection:
-
-$$ 
-P_\text{de novo}(t) = 1 - \text{exp}{ {( - r_0 T(t) )} },
-$$
-
-and 2) the usage of contaminated toilets:
-
-$$
-P_\text{toilet}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} \right) }
+-   $x_i(t) = 0$ if patient $i$ is not infected at time $t$,
+-   $x_i(t) = 1$ if patient $i$ is infected at time $t$.
+
+Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge.
+At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets:
+
+$$
+P_i(t) = 1 - \left[ (1 - P_{\text{de novo}, i}(t)) \cdot (1 - P_{\text{toilet}, i}(t)) \right],
 $$
@@
 
--   $R_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,
+$$ 
+P_{\text{de novo}, i}(t) = 1 - \text{exp}{ {( - r_0 \cdot T_i(t) )} },
+$$
+
+$$
+P_{\text{toilet}, i}(t) = 1 - \text{exp}{ \left( {-\sum_{j=0}^{U_i(t)}[C(t) \cdot T_i(t)]} \right) }
+$$
+
+where:
+
+-   $P_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,
 
@@
 
--   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy,
-
--   $U_i(t)$ is the number of times patient $p$ frequents the toilet within the time interval $t$,
-
--   $C(t_j)$ is the contamination level of the toilet at time $t_j$, and
+-   $U_i(t)$ is the number of times patient $i$ frequents the toilet within the time interval $t$,
+
+-   $C(t)$ is the contamination level of the toilet at time $t$,
+
+-   $T(t)$ is the treatment multiplier that depends on antimicrobial therapy and is defined as:
 
@@
     T(t) = \begin{cases}
-    m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
-    1 & \text{otherwise.}
+      m & \text{if on antimicrobial therapy} \\
+      1 & \text{otherwise, and}
     \end{cases}
@@
 
-The agent-based model conducted Bernoulli trials to determine if a patient became infected at each time step $t$:
-
-$$
-X_\text{de novo} \sim \text{Bernoulli}(P_\text{de novo}(t)),
-$$
-
-where $X_\text{de novo}$ = 1 if patient becomes infected *de novo* at time $t$, and 0 otherwise.
-
-$$
-X_\text{toilet} \sim \text{Bernoulli}(P_\text{toilet}(t)),
-$$
-
-where $X_\text{toilet}$ = 1 if patient becomes infected from a contaminated toilet at time $t$, and 0 otherwise.
-
-The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was sucessful:
-
-$$
-X_p(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
-$$
-
-where $X_i(t)$ = 1 if patient $p$ becomes infected at time $t$, and 0 otherwise.
-
-During the simulation level of contamination of each toilet increased when an infected patient used it, and decreased
-when the toilet was cleaned.
+-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy.
+
+In the agent-based model implementation, Bernoulli trials were conducted to determine if a patient became infected at each time step $t$:
+
+$$
+X_\text{de novo, i}(t) \sim \text{Bernoulli}(P_{\text{de novo}, i}(t)),
+$$
+
+where $X_{\text{de novo}, i}$ = 1 if patient $i$ becomes infected *de novo* at time $t$, and 0 otherwise.
+
+Further Bernoulli trials were conducted to determine if a patient became infected from a contaminated toilet at each time step $t$:
+
+$$
+X_{\text{toilet}, i}(t) \sim \text{Bernoulli}(P_{\text{toilet}, i}(t)),
+$$
+
+where $X_{\text{toilet}, i}$ = 1 if patient $i$ becomes infected from a contaminated toilet at time $t$, and 0 otherwise.
+
+The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was successful:
+
+$$
+X_i(t) = \text{Max}(X_{\text{de novo}, i}, X_{\text{toilet}, i}),
+$$
+
+where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.
+
+During the simulation, the contamination level $C$ of each toilet increased when an infected patient used it, and decreased when the toilet was cleaned.
 
@@
 
+Each patient's admission duration was sampled from a Poisson distribution with a configurable mean (in days).
+Upon discharge, patients were removed from the simulation, and replaced with new admissions to maintain 100% occupancy.
+New admissions had a probability of being infected based on the community infection rate parameter.
+
 ## Stochastic system dynamics model
 
-To build a high-level system dynamics model, a deterministic model was firstly described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
-
-$$
-\frac{dS}{dt} = (1-p)A - D \cdot \frac{S}{S + I} - \beta S - rS
-$$
-
-$$
-\frac{dI}{dt} = pA - D \cdot \frac{I}{S + I} + \beta S + rS
+To build a high-level stochastic system dynamics model, the model was first described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
+
+$$
+\frac{dS}{dt} = (1-\lambda)A - D \cdot \frac{S}{S + I} - S (\beta + r)
+$$
+
+$$
+\frac{dI}{dt} = \lambda A - D \cdot \frac{I}{S + I} + S (\beta + r)
 $$
@@
 
--   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
 -   $S$ is the number of susceptible inpatients,
 -   $I$ is the number of infected inpatients,
--   $A$ is the rate of new admissions,
--   $p$ is the community infection rate,
+-   $dS/dt$ and $dI/dt$ are the rates of change of susceptible/non-infected and infected inpatients, respectively,
+-   $A$ is the number of new admissions,
+-   $D$ is the number of patient discharges,
+-   $\lambda$ is the community infection rate,
 -   $\beta$ is the transmission coefficient,
--   $r$ is the rate of random (*de novo*) infection,
--   $D$ is the rate of patient discharges.
-
-The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $D = A$.
+-   $r$ is the rate of random (*de novo*) infection.
+
+The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $d = a$.
 
@@
 
-To add stochasticity to the model, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:
-
-$$
-D_{\text{total}, t} \sim \text{Poisson}(D),
-$$
-
-where $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.
+Next, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:
+
+$$
+D_{\text{total}, t} \sim \text{Poisson}(d),
+$$
+
+where $d$ is the discharge rate and $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.
+
 Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:
@@
 $$
-A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, p),
+A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, \lambda),
 $$

### 2025-11-04 paper/paper.qmd
Renamed SD params

@@
 #| fig.cap: "System dynamics model.
-$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
+$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $\\lambda$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
 knitr::include_graphics("./images/sd_model.svg")
@@
 
-A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak.
-The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community infection rate).
+A similar approach was used to generate outbreaks in the stochastic system dynamics model.
+To align with the agent-based model, parameters were changed at four discrete time points: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
+Three parameter regimes were used: baseline (pre- and post-outbreak), outbreak (pre-control), and control (during enhanced IPC).
+During the control period only the transmission coefficient $\\beta$ changed; the community infection rate $\\lambda$ was unaffected by control measures (it only differed between baseline and outbreak).
 The parameter $r$ (rate of random infection) was kept constant throughout the simulation.
-
-## Model inputs and outputs
-
-Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community infection rate, antimicrobial prescription rate); and 3) infection parameters (e.g., relative risk increase of antimicrobial therapy on infection).
-
-The outputs of the model included: 1) number of patients admitted to the hospital; 2) number of patients with community-acquired infection; 3) number of patients with hospital-acquired (nosocomial) infection.
-The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.
 
@@
 Rejection sampling \acr{ABC} was employed to fit the \acr{ABM} to the observed outbreak data [@sunnakerApproximateBayesianComputation2013].
+Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community infection rate, antimicrobial prescription rate); and 3) infection parameters (e.g., relative risk increase of antimicrobial therapy on infection).
+
+The outputs of the model included: 1) number of patients admitted to the hospital; 2) number of patients with community-acquired infection; 3) number of patients with hospital-acquired (nosocomial) infection.
+The simulation was run with each unit of time corresponding to one day; therefore, the model's outputs were for each simulated day.
+
 Priors were extracted from existing literature; flat priors when data were unavailable.
 A summary of the priors used is available in @tbl-priors.
-The proportion of toilets that were shared between multiple patients was set to 60%.
+The hospital design was configured based on UK standards for hospital inpatient facilities [@departmentofhealthHealthBuildingNote2013].
+This consisted of 600 beds in 24 wards, of which 376 (63%) were single-occupancy rooms.
+Each ward had a median capacity of 24 patients, and a median of two multi-occupancy bay (each of which had four patients).
 
@@
 | Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
-| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
 | Community infection rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
@@
 
-Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$.
+Firstly, $n$ independent random samples were drawn from the parameter prior distributions giving $\theta_1,\theta_2,\ldots,\theta_n$.
 Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
@@
 
-Otherwise, $S_i$ was rejected.
-The parameter samples that generated the accepted simulations were used to produce a posterior distribution.
+Otherwise, $\theta_i$ was rejected.
+The parameter samples that generated the accepted simulations were used to approximate a posterior distribution.
 $\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.
@@
 To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
-The priors used for ABC were selected to be compatible with increased transmission or infection rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency.
-To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded.
+The priors used for ABC to calibrate the parameters during the outbreak period were selected to be compatible with increased transmission or infection rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency, when compared to the pre-outbreak period.
+To achieve this, the posterior distribution of the pre-outbreak calibration was used to inform the outbreak priors.
+For example, the lower bound of the antimicrobial prescription rate parameter outbreak prior was set to the mean of the pre-outbreak posterior.
 At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
 The prior distributions for these parameters were selected to decrease transmission and infection -- generally the reverse of the 'outbreak' parameters.
-Again, only samples from the latin hypercube that satisfied these requirements were used.
 
@@
 
-\acr[first_use=TRUE, case=sentence]{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).
+A similar calibration approach using \acr{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).
 
@@
 
-To assess the sensitivity of the calibrated \acr{ABM} to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
+To assess the sensitivity of the calibrated \acr{ABM} to changes in infection control practice, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019].
 Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect.
@@
   # remove gridlines
+  scale_x_continuous(labels = scales::label_number(accuracy = 0.001)) +
   theme(panel.grid.major = element_blank(),
@@
 set.seed(42)
-randomLHS_sd <- randomLHS(1000000, 8)
+# Sampling configuration: 'lhs' for Latin Hypercube (default) or 'iid' for independent random draws
+sd_sampling_method <- "iid"  # change to "iid" for independent sampling
+n_pool <- 10000
+sd_dims <- 11
+if (sd_sampling_method == "lhs") {
+  randomLHS_sd <- randomLHS(n_pool, sd_dims)
+} else {
+  # independent uniform draws in [0,1]
+  randomLHS_sd <- matrix(runif(n_pool * sd_dims), ncol = sd_dims)
+}
+# Set SD timepoint priors similar to variable_pre_outbreak.R (no ABM dependency)
+eicm_start <- as.numeric(difftime(ymd('2004-11-15'), ymd('2002-01-01'), units = 'days'))
+outbreak_start_center <- as.numeric(difftime(ymd('2004-10-01'), ymd('2002-01-01'), units = 'days'))
+outbreak_end_center   <- as.numeric(difftime(ymd('2005-05-31'), ymd('2002-01-01'), units = 'days'))
+control_end_center    <- outbreak_end_center + 90
 sd_param_grid <- data.frame(init_population = rep(610, 1000),
@@
                             baseline_beta = qunif(randomLHS_sd[, 4], 0.0001, 0.01),
-                            baseline_p = qunif(randomLHS_sd[, 5], 0.001, 0.1),
+                            baseline_lambda = qunif(randomLHS_sd[, 5], 0.001, 0.1),
                             outbreak_beta = qunif(randomLHS_sd[, 6], 0.0001, 0.01),
-                            outbreak_p = qunif(randomLHS_sd[, 7], 0.001, 0.1),
-                            outbreak_start = rep(1042, 1000),
-                            outbreak_end = rep(1234, 1000),
+                            outbreak_lambda = qunif(randomLHS_sd[, 7], 0.001, 0.1),
+                            control_beta = qunif(randomLHS_sd[, 8], 0.0001, 0.01),
+                            outbreak_start = qunif(randomLHS_sd[, 9], outbreak_start_center - 90, outbreak_start_center + 90),
+                            outbreak_end   = qunif(randomLHS_sd[, 10], outbreak_end_center - 90, outbreak_end_center + 90),
+                            control_start  = rep(eicm_start, 1000),
+                            control_end    = qunif(randomLHS_sd[, 11], control_end_center - 30, control_end_center + 30),
                             n_days = rep(2585, 1000))
@@
 sd_param_grid <- sd_param_grid %>% 
-  filter(baseline_beta < outbreak_beta & baseline_p < outbreak_p) %>% 
-  slice_sample(n = 10000) %>% 
+  # filter(baseline_beta < outbreak_beta & baseline_lambda < outbreak_lambda) %>% 
+  # slice_sample(n = 10000) %>% 
   mutate(param_row = row_number())
-n_seeds <- 2
+n_seeds <- 5
 sd_sims <- 
   pmap_dfr(sd_param_grid,
-        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row) {
+  \(init_population, A, r, baseline_beta, baseline_lambda, outbreak_beta, outbreak_lambda, control_beta, outbreak_start, outbreak_end, control_start, control_end, n_days, param_row) {
           map(1:n_seeds,
@@
                 set.seed(seed)
-                sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
+      sd_stochastic(init_population=init_population,
+          A=A,
+          r=r,
+          baseline_beta = baseline_beta,
+          baseline_lambda = baseline_lambda,
+          outbreak_beta = outbreak_beta,
+          outbreak_lambda = outbreak_lambda,
+          control_beta = control_beta,
+          outbreak_start = outbreak_start,
+          control_start = control_start,
+          outbreak_end = outbreak_end,
+          control_end = control_end,
+          n_days = n_days) %>% 
                   mutate(`random-seed` = seed,
@@
   pmap_dfr(posterior_check_params,
-        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, seed) {
+    \(init_population, A, r, baseline_beta, baseline_lambda, outbreak_beta, outbreak_lambda, control_beta, outbreak_start, outbreak_end, control_start, control_end, n_days, seed) {
       set.seed(seed)
-            sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
+    sd_stochastic(init_population = init_population,
+      A = A,
+      r = r,
+      baseline_beta = baseline_beta,
+      baseline_lambda = baseline_lambda,
+      outbreak_beta = outbreak_beta,
+      outbreak_lambda = outbreak_lambda,
+      control_beta = control_beta,
+      outbreak_start = outbreak_start,
+      control_start = control_start,
+      outbreak_end = outbreak_end,
+      control_end = control_end,
+      n_days = n_days) %>% 
         mutate(`random-seed` = seed) %>% 
@@
 
-```{r fig-abc-sd-posteriors, include=TRUE}
-#| echo: false
-#| fig.cap: "Posterior distributions of parameters obtained from approximate Bayesian computation using summary statistics generated from the stochastic deterministic model.
-The red line shows the prior distribution; the blue line shows the posterior distribution."
-
+```{r fig-abc-sd-posteriors, include=TRUE, dev='png'}
+#| echo: false
+#| out.height: "400px"
+#| fig.cap: "Changes in SD model parameters across baseline, outbreak, and control periods.
+Density plots show the posterior distributions of key parameters after calibration using approximate Bayesian computation."
+param_names_sd <- c(
+  "R" = "Community infection rate (_r_)",
+  "Beta" = "Transmission rate (\U03B2)",
+  "Lambda" = "Toilet contamination rate (\U03BB)"
+)
 sd_posterior_params_abc <- abc_sd$unadj.values
@@
   as.data.frame() %>% 
-  dplyr::select(r, baseline_beta, baseline_p, outbreak_beta, outbreak_p) %>% 
-  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>% 
-  mutate(parameter = factor(parameter, 
-                            levels = c("r", "baseline_beta", "outbreak_beta", "baseline_p", "outbreak_p"))) %>% 
-  ggplot(aes(x = value)) +
-  geom_density(fill = "blue", alpha = 0.5) +
-  facet_wrap(~ parameter, scales = "free")
+  # select only parameter columns (exclude timepoints)
+  dplyr::select(r, baseline_beta, outbreak_beta, control_beta, baseline_lambda, outbreak_lambda) %>% 
+  pivot_longer(cols = everything(), names_to = "name", values_to = "value") %>% 
+  mutate(type = case_when(
+    str_starts(name, "outbreak_") ~ "Outbreak",
+    str_starts(name, "control_")  ~ "Control",
+    TRUE                           ~ "Baseline"
+  )) %>% 
+  mutate(name = str_remove(name, "^baseline_|^outbreak_|^control_")) %>% 
+  mutate(name = str_replace_all(name, "_", " ")) %>% 
+  mutate(name = str_to_title(name)) %>% 
+  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
+  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
+  geom_density(alpha = .3) +
+  facet_wrap(~ name, scales = "free_x", labeller = as_labeller(param_names_sd),
+             ncol = 1) +
+  theme_minimal() +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank(),
+        axis.ticks.y = element_blank(),
+        strip.text = ggtext::element_markdown()) +
+  xlab("Parameter value") + 
+  theme(legend.title = element_blank(),
+        legend.position = "top"
+        # ,text=element_text(size=24)
+        ) + 
+  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
+  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
+  scale_x_continuous(labels = scales::label_number(accuracy = 0.001)) +
+  theme(panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank())
 
@@
 
-_The discussion needs to explore the different modelling approaches used (SD vs ABM) and why we eventually settled on the ABM as the main model.
-This is sort of captured in the results, but maybe it needs to be here as well?_
-
 \acr[case=sentence, first_use=true]{IPC} is the practice of preventing the spread of infectious diseases in healthcare settings [@worldhealthorganizationGuidelinesCoreComponents2016].
@@
 difficile_ outbreak, it can be adapted for use in other pathogens that are transmitted through the faecal-oral route under antimicrobial selective pressure, such as vancomycin-resistant _Enterococci_ and extended-spectrum beta-lactamase-producing Enterobacterales.
+
+# References

### 2025-11-04 models/sd_stochastic.cpp

@@ -10,20 +10,23 @@ inline int rbinom_int(int n, double p) {
 
 // [[Rcpp::export]]
 DataFrame sd_stochastic(
-    int init_population,
-    int A,                  // mean discharges per day
-    double r,               // spontaneous colonisation rate
-    double baseline_beta,
-    double baseline_p,
-    double outbreak_beta,
-    double outbreak_p,
-    int outbreak_start,
-    int outbreak_end,
-    int n_days
+  int init_population,
+  int A,                  // mean discharges per day
+  double r,               // spontaneous colonisation rate (constant across phases)
+  double baseline_beta,
+  double baseline_lambda,
+  double outbreak_beta,
+  double outbreak_lambda,
+  double control_beta,
+  int outbreak_start,
+  int control_start,
+  int outbreak_end,
+  int control_end,
+  int n_days
 ) {
   // initial state
-  int S = std::round(init_population * (1.0 - baseline_p));
-  int I = std::round(init_population * baseline_p);
+  int S = std::round(init_population * (1.0 - baseline_lambda));
+  int I = std::round(init_population * baseline_lambda);
   
   // output storage
   IntegerVector days(n_days + 1);
@@ -45,14 +48,22 @@ DataFrame sd_stochastic(
   
   // simulation loop
   for (int t = 1; t <= n_days; ++t) {
-    // choose parameter regime
-    double beta, p;
+    // choose parameter regime using four time points
+    // beta varies across baseline/outbreak/control; p only varies across baseline/outbreak (not affected by control)
+    double beta, lambda;
+    // community colonisation rate (lambda): only changes during outbreak window
     if (t > outbreak_start && t < outbreak_end) {
+      lambda = outbreak_lambda;
+    } else {
+      lambda = baseline_lambda;
+    }
+    // transmission coefficient (beta): changes during outbreak and control windows
+    if (t > control_start && t < control_end) {
+      beta = control_beta;
+    } else if (t > outbreak_start && t < outbreak_end) {
       beta = outbreak_beta;
-      p    = outbreak_p;
     } else {
       beta = baseline_beta;
-      p    = baseline_p;
     }
     
     // --- 1. Discharges ---
@@ -74,8 +85,8 @@ DataFrame sd_stochastic(
     int new_infections = inf_S + spont_S;
     
     // --- 3. Admissions (equal to discharges) ---
[…]

### 2025-12-03 paper/paper.qmd
ABM math updates

@@
 ---
-title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
+title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model
 authors:
@@
 
+On admission, each patient had a probability $\lambda$ of being infected, representing the community infection rate.
+The simulation determined the infection status of each patient on admission using Bernoulli trials:
+
+$$
+X_{\text{community}, i} \sim \text{Bernoulli}(\lambda),
+$$
+
+where $X_{\text{community}, i} = 1$ if patient $i$ is infected on admission, and 0 otherwise.
+
 Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge.
-At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets:
+At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets.
+The probability of infection through these two routes---$P_{\text{de novo}, i}(t)$ and $P_{\text{toilet}, i}(t)$, respectively---were independent; therefore, the probability of patient $i$ acquiring infection at time $t$ was given by:
 
@@
 
+* $(1- P_{\text{de novo}, i}(t))$ is the probability of patient $i$ not acquiring infection *de novo* at time $t$,
+
+* $(1 - P_{\text{toilet}, i}(t))$ is the probability of patient $i$ not acquiring infection from a contaminated toilet at time $t$, and
+
+* $[(1 - P_{\text{de novo}, i}(t)) \cdot (1 - P_{\text{toilet}, i}(t))]$ is the probability of patient $i$ not acquiring infection at time $t$.
+
+The individual probabilities were defined as:
+
 $$ 
@@
 where:
-
--   $P_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,
 
@@
 
--   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy.
+-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy, where $m>0$.
 
@@
 $$
-X_\text{de novo, i}(t) \sim \text{Bernoulli}(P_{\text{de novo}, i}(t)),
-$$
-
-where $X_{\text{de novo}, i}$ = 1 if patient $i$ becomes infected *de novo* at time $t$, and 0 otherwise.
-
-Further Bernoulli trials were conducted to determine if a patient became infected from a contaminated toilet at each time step $t$:
-
-$$
-X_{\text{toilet}, i}(t) \sim \text{Bernoulli}(P_{\text{toilet}, i}(t)),
-$$
-
-where $X_{\text{toilet}, i}$ = 1 if patient $i$ becomes infected from a contaminated toilet at time $t$, and 0 otherwise.
-
-The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was successful:
-
-$$
-X_i(t) = \text{Max}(X_{\text{de novo}, i}, X_{\text{toilet}, i}),
-$$
-
-where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.
+X_{\text{hospital}, i}(t) \sim \text{Bernoulli}(P_i(t))
+$$
+
+where $X_{\text{hospital}, i}(t) = 1$ if patient $i$ acquires infection during their hospital stay at time $t$, and 0 otherwise.
 
@@
 To capture delays in moving patients into side rooms (which can be due to delays in diagnosis, and/or lack of available side rooms), the simulation included an optional parameter that samples the delay from a Poisson distribution with a configurable mean (in days).
-Once the infected patient has completed their delay, they became eligible to move to the next available side room.
+Once the infected patient completed their delay, they became eligible to move to the next available side room.
 To implement this, the simulation used a \acr{FIFO} queue to track infected patients waiting for a side room.
@@
 Each patient's admission duration was sampled from a Poisson distribution with a configurable mean (in days).
-Upon discharge, patients were removed from the simulation, and replaced with new admissions to maintain 100% occupancy.
-New admissions had a probability of being infected based on the community infection rate parameter.
+Upon discharge, patients were removed from the simulation; each patient was then replaced with new patient admission.
+Although real hospitals do not operate at 100% occupancy at all times, this assumption was made to simplify the model.
 

### 2025-12-03 paper/paper.qmd
SD removed

@@
 The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.
-We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach for this application.
 
@@
 
-## Stochastic system dynamics model
-
-To build a high-level stochastic system dynamics model, the model was first described in terms of the following differential equations (see @fig-sd-model), an adaptation of *SI* (susceptible--infection) models [@andersonDiscussionKermackMcKendrickEpidemic1991; @keelingModelingInfectiousDiseases2011]:
-
-$$
-\frac{dS}{dt} = (1-\lambda)A - D \cdot \frac{S}{S + I} - S (\beta + r)
-$$
-
-$$
-\frac{dI}{dt} = \lambda A - D \cdot \frac{I}{S + I} + S (\beta + r)
-$$
-
-where:
-
--   $S$ is the number of susceptible inpatients,
--   $I$ is the number of infected inpatients,
--   $dS/dt$ and $dI/dt$ are the rates of change of susceptible/non-infected and infected inpatients, respectively,
--   $A$ is the number of new admissions,
--   $D$ is the number of patient discharges,
--   $\lambda$ is the community infection rate,
--   $\beta$ is the transmission coefficient,
--   $r$ is the rate of random (*de novo*) infection.
-
-The hospital capacity was constantly kept at 100% occupancy (i.e., patients were immediately replaced on discharge), therefore, $d = a$.
-
-```{r fig-sd-model, include=TRUE}
-#| echo: false
-#| fig.cap: "System dynamics model.
-$S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $\\lambda$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
-knitr::include_graphics("./images/sd_model.svg")
-```
-
-Next, the number of susceptible and infected patients at each discrete time $t + 1$ was calculated by first determining the number of patients discharged:
-
-$$
-D_{\text{total}, t} \sim \text{Poisson}(d),
-$$
-
-where $d$ is the discharge rate and $D_{\text{total}, t}$ is the total number of patients (both susceptible, $D_S$, and infected $D_I$), discharged at time $t$.
-
-Therefore, $D_{\text{total}, t} = D_{S,t} + D_{I,t}$, where:
-
-$$
-D_{S,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \right \rceil,
-$$
-
-and
-
-$$
-D_{I,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \right \rceil = D_{\text{total}, t} - D_{S,t},,
-$$
-
-where $\lfloor x \rceil$ is the rounding of $x$ to the nearest integer.
-
-Next, new admissions were calculated. Since $A_{\text{total}, t} = D_{\text{total}, t}$,
-
-$$
-A_{I,t} \sim \text{Binomial}(A_{\text{total}, t}, \lambda),
-$$
-
-and
-
-$$
-A_{S,t} = A_{\text{total}, t} - A_{I,t},
-$$
-
-To calculate the number of hospital-acquired infections (i.e., $\beta S_t + r S_t$), the binomial distribution was used:
-
-$$
-\text{HAI}_{\text{transmission}, t} \sim \text{Binomial}(S_t - D_{S, t}, \beta),
-$$
-
-and
-
-$$
-\text{HAI}_{\text{de novo}, t} \sim \text{Binomial}(S_t - D_{S, t} - \text{HAI}_{\text{transmission}, t}, r),
-$$
-
-where $\text{HAI}_{\text{transmission}, t}$ is the number of hospital-acquired infections due to transmission at time $t$, and $\text{HAI}_{\text{de novo}, t}$ is the number of hospital-acquired infections due to *de novo* infection at time $t$.
-
-Finally, the number of susceptible and infected patients at time $t + 1$ was updated as:
-
-$$
-S_{t+1} = S_t + A_{S,t} - D_{S,t} - \text{HAI}_{\text{transmission}, t} - \text{HAI}_{\text{de novo}, t},
-$$
-
-and
-
-$$
-I_{t+1} = I_t + A_{I,t} - D_{I,t} + \text{HAI}_{\text{transmission}, t} + \text{HAI}_{\text{de novo}, t}.
-$$
-
 ## Outbreak generation {#sec-outbreak-generation}
@@
 To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
-
-A similar approach was used to generate outbreaks in the stochastic system dynamics model.
-To align with the agent-based model, parameters were changed at four discrete time points: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
-Three parameter regimes were used: baseline (pre- and post-outbreak), outbreak (pre-control), and control (during enhanced IPC).
-During the control period only the transmission coefficient $\\beta$ changed; the community infection rate $\\lambda$ was unaffected by control measures (it only differed between baseline and outbreak).
-The parameter $r$ (rate of random infection) was kept constant throughout the simulation.
+By changing parameters, the model simulated the conditions that led to increased transmission of infection, followed by \acr{IPC} measures that reduced transmission.
 
@@
 ## Model calibration {#sec-model-calibration}
-
-### Calibration of \acr[first_use=TRUE]{ABM}
 
@@
 
-Otherwise, $\theta_i$ was rejected.
-The parameter samples that generated the accepted simulations were used to approximate a posterior distribution.
+Otherwise, $\theta_i$ was rejected, giving a set of accepted parameters $\hat{\theta}_1, \ldots, \hat{\theta}_m$, where $m \leq n$.
+The parameter samples---i.e., $\hat{\theta}$ that generated the accepted simulations were used to approximate a posterior distribution.
 $\epsilon$ was set such that the acceptance rate was approximately 2.5%, which was deemed to be a good compromise between the accuracy of the posterior distribution and the computational cost of running the simulations.
@@
 
-As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates.
-To sample from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were samples with replacement were taken and used as an input to the stochastic \acr{SD} model.
+As a posterior predictive check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates.
+A set of parameters was generated by sampling with replacement from $\hat{\theta}$, and used as an input to the simulation model.
+The results of these simulations were compared to the observed data to assess how well the calibrated model could reproduce the observed outbreak dynamics.
 
@@
 `r posterior_predictive_n_seeds` seeds were used for the samples with replacement from the posterior distribution.
-
-### Calibration of \acr[first_use=TRUE]{SD} model
-
-A similar calibration approach using \acr{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).
 
@@
 ```
-
-Both the \acr{ABM} and the \acr{SD} model were able to simulate the faecal-oral transmission of infection in a hospital setting.
-Although the \acr{SD} was computationally simpler, and produced simulated infection rates that were very similar to the observed rates (see @fig-posterior-predictive-check), the \acr{ABM} was based on a simulation that more closely resembled the workings of a real hospital.
-Importantly, the parameters of the \acr{ABM} were directly based on modifiable infection control practices, such as toilet cleaning frequency and antimicrobial prescription rates.
-Therefore, the main analyses and senstivity analyses were performed using the \acr{ABM}.
 

### 2025-12-04 paper/paper.qmd
conor comments

@@
 
-\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings.
+\acr[case=sentence, first_use]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings.
 Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected patients [@worldhealthorganizationGlobalReportInfection2022].
@@
 
-The use of computer simulation techniques such as \acr{ABM} and can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact.
-Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021].
-An advantage of \acr{ABM} over other simulation approaches is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019].
-This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].
-Another simulation approach is stochastic \acr{SD} modelling, which is a high-level approach that models the flow of patients between different states, such as susceptible and infected states, and is particularly useful for understanding feedback loops within systems.
-
-Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
+The use of computer simulation techniques can be used to model the spread of infectious diseases, and has been used to describe the transmission of pathogens that are important in healthcare settings, such as
+\acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021].
+An advantage of simulation models is that they can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019].
+One such technique is \acr{ABM}, which models complex systems by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact based on a set of user-defined rules.
+Compared to other simulation approaches, such as \acr{SD}, which is a high-level approach that models the flow of patients between different compartments (e.g., susceptible, infected, recovered), \acr{ABM} allows for a detailed representation of individual agents and their interactions.
+This makes \acr{ABM} particularly well-suited for modelling the spread of infections in healthcare settings, where individual patient factors and interactions can influence transmission dynamics.
+
+Here, we describe an \acr{ABM} that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure.
 Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods.
@@
 
-## Agent-based model
+## \acr[case=sentence, first_use]{ABM} description
 
@@
 $$
-x_i(t) \in \{0, 1\}
+x_i(t) \in \{0, 1\},
 $$
@@
 
-In the agent-based model implementation, Bernoulli trials were conducted to determine if a patient became infected at each time step $t$:
+In the \acr{ABM} implementation, Bernoulli trials were conducted to determine if a patient became infected at each time step $t$:
 
@@
 
-To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
+To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the \acr{ABM} optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced \acr{IPC} measures.
 By changing parameters, the model simulated the conditions that led to increased transmission of infection, followed by \acr{IPC} measures that reduced transmission.
@@
 Firstly, $n$ independent random samples were drawn from the parameter prior distributions giving $\theta_1,\theta_2,\ldots,\theta_n$.
-Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets.
+Next, each sample parameter point $\theta_i$ was used as an input to the \acr{ABM}, generating a total of $n$ simulated datasets.
 The number of daily cases of nosocomial \acr{CDI} were extracted from the model outputs, and converted to a time series of nosocomial \acr{CDI} per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$.
@@
 
-A two-step approach to calibration was taken.
-Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate.
-Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
-To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation.
-The priors used for ABC to calibrate the parameters during the outbreak period were selected to be compatible with increased transmission or infection rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency, when compared to the pre-outbreak period.
-To achieve this, the posterior distribution of the pre-outbreak calibration was used to inform the outbreak priors.
-For example, the lower bound of the antimicrobial prescription rate parameter outbreak prior was set to the mean of the pre-outbreak posterior.
-At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures.
-The prior distributions for these parameters were selected to decrease transmission and infection -- generally the reverse of the 'outbreak' parameters.
+The model was calibrated in two stages.
+First, \acr{ABC} was applied to the pre-outbreak data to obtain posterior parameter distributions that reproduced baseline infection levels.
+These posteriors were then used as priors for a second calibration phase covering the outbreak period and the introduction of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009].
+To represent the onset of the outbreak, key parameters were shifted at a defined time point, with priors chosen to allow increased transmission or infection pressure (informed by the earlier posteriors).
+Later, at another defined time point, parameters were shifted again to reflect enhanced \acr{IPC} measures, with priors chosen to reduce transmission, reversing the outbreak-related changes.
+This calibration approach reduced the total number of required simulations, and therefore the computational cost, by identifying baseline parameters before simulating the more complex outbreak dynamics.
 
@@
 
-Due to the inherent stochasticity within the agent-based model, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated.
+Due to the inherent stochasticity within the \acr{ABM}, all calibration and simulation runs were aggregated over multiple runs with different random number generators, and the mean rates of infection were calculated.
 For the samples used for ABC, `r results_outbreak_control_n_seeds` different random number generator seeds were used;
@@
 
-The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.NetLogo1999].
+The \acr{ABM} was implemented in NetLogo v6.2.2 [@wilenskyu.NetLogo1999].
 The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@rcoreteamLanguageEnvironmentStatistical2022; @saleckerNlrxPackageNextgeneration2019].
@@
 The densities of the posterior parameter distributions are shown in @fig-posterior-densities.
-In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).
+The posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).
 
@@
 
-Using the posterior distributions as inputs to the simulation model, as described in @sec-model-calibration, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
+Using the posterior distributions as inputs to the simulation model, the model was able to generate scenarios of outbreaks that closely matched the observed data (see @fig-posterior-predictive-check).
 In these simulated scenarios, the pre-outbreak baseline rate was a median of `r round(pre_outbreak_posterior_predictive_sims_rates$rate_median, 2)` cases per 1000 patient days \[\acr{IQR} range `r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_low, 2)`--`r round(pre_outbreak_posterior_predictive_sims_rates$rate_q_high, 2)`\], compared with the real rate of 1.25 cases per 1000 \[\acr{IQR} 0.94--1.56\].
@@
 
-```{r fig-abc-rates-retained-posteriors, include=TRUE}
+```{r fig-abc-rates-retained-posteriors, include=FALSE}
 #| echo: false
@@
 Simulations were generated by sampling with replacement from the posterior parameter distributions after approximate Bayesian computation.
-Each simulation was run using different random number generators.
-The red line shows the true observed infection rates.
-The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
+Each simulation was run using different random number generators."
 posterior_predictive_sim_rates %>% 
@@
   ggplot() +
-  geom_line(aes(x = date_sim, y = m)) +
-  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
-  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
-  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
-  guides(color="none") +
-  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
+  geom_ribbon(aes(x = date_sim, ymin = q_low, ymax = q_high, fill = "Simulated 25th-75th quantiles"), alpha = 0.3) +
+  # geom_line(aes(x = date_sim, y = q_high, color = "Simulated 25th-75th quantiles"), alpha=0.3) +
+  # geom_line(aes(x = date_sim, y = q_low, color = "Simulated 25th-75th quantiles"), alpha=0.3) +
+  geom_line(aes(x = date_sim, y = m, color = "Median simulated rates")) +
+  geom_line(data = long_observed_ss, aes(x = x, y = rates, color = "Observed rates")) +
+  scale_fill_manual(values = c("Simulated 25th-75th quantiles" = "blue"), name = "") +
+  scale_color_manual(values = c("Median simulated rates" = "black", "Observed rates" = "red", "Simulated 25th-75th quantiles" = "blue"), name = "") +
+  xlab("Date") + ylab("Rate (cases per 1000 patient days)") +
+  theme_minimal() + 
+  theme(legend.position = "top")
 ```
@@
   summarise(
-    mean_cases = mean(`outbreak-cases`),
-    sd_cases   = sd(`outbreak-cases`),
+    med_cases = median(`outbreak-cases`),
+    q_high    = quantile(`outbreak-cases`, 0.75),
+    q_low     = quantile(`outbreak-cases`, 0.25),
     .groups    = "keep"
@@
   ) %>% 
-  ggplot(aes(x = value, y = mean_cases)) + 
-  geom_line() + 
-  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
+  ggplot(aes(x = value)) + 
+  geom_ribbon(aes(ymin = q_low, ymax = q_high, fill = "Simulated 25th-75th quantiles"), alpha = 0.3) +
+  geom_line(aes(y = med_cases, color = "Median simulated rates"), linewidth = 0.7) +
   facet_wrap(~ param, scales = "free") +    # free_x required to use facetted_pos_scales
@@
   labs(x = "Parameter value",
-       y = "Number of infected patients")
+       y = "Number of infected patients") +
+  scale_fill_manual(values = c("Simulated 25th-75th quantiles" = "blue"), name = "") +
+  scale_color_manual(values = c("Median simulated rates" = "black"), name = "") +
+  theme_minimal() +
+  theme(legend.position = "top")
 ```
@@
 
-```{r fig-abc-sd-posteriors, include=TRUE, dev='png'}
+```{r fig-abc-sd-posteriors, include=FALSE, dev='png'}
 #| echo: false
@@
 
-```{r fig-abc-sd-simulated-rates, include=TRUE}
+```{r fig-abc-sd-simulated-rates, include=FALSE}
 #| echo: false
@@
 Central to this is the operational management of an outbreak, defined as an increase in the number of infected cases above the expected rate.
-This research describes the design and calibration of an agent-based model that simulates the spread of infection in a hospital setting.
+This research describes the design and calibration of an \acr{ABM} that simulates the spread of infection in a hospital setting.
 The primary motivation of this work is to assess the impact of different infection prevention interventions (such as change in practice) on rates of infection.
@@
 
-Although agent-based models have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations.
+Although \acrs{ABM} have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations.
 The model described here is a simplification of what is likely to be a highly intricate and complicated system.
@@
 Antimicrobial class and duration of exposure are factors that are not accounted for in this model.
-An agent-based model that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated.
+An \acr{ABM} that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated.
 A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016].
-Another limitation of agent-based models is the requirement for detailed data to calibrate and validate the model.
+Another limitation of \acrs{ABM} is the requirement for detailed data to calibrate and validate the model.
 Here, the data used was primarily from a retrospective descriptive dataset.
@@
 
-Agent-based models that simulate the hospital transmission of pathogens transmitted through the faecal-oral route, including _C.
+\acrs[case=sentence, first_use]{ABM} that simulate the hospital transmission of pathogens transmitted through the faecal-oral route, including _C.
 difficile_ and vancomycin-resistant _Enterococci_, have been reported previously [@codellaAgentbasedSimulationModel2015; @debosckerAgentbasedModelSimulate2021].
@@
 
-In conclusion, the agent-based model described here can simulate outbreaks of infection in a hospital setting, and estimate the impact of different infection control interventions on the number of nosocomial infections.
+In conclusion, the \acr{ABM} described here can simulate outbreaks of infection in a hospital setting, and estimate the impact of different infection control interventions on the number of nosocomial infections.
 Although the model was calibrated to a _C.

### 2025-12-30 paper/paper.qmd
round of edits

@@
 ---
-title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model
-authors:
-  - name: Alessandro Gerada
-    affiliation: University of Liverpool
-    roles: writing
-    corresponding: true
+title: "Simulating interventions to prevent gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings: a model calibration study"
 bibliography: references.bib
@@
   - acronyms
+  - authors-block
 acronyms:
@@
   insert_links: false
+authors:
+  - name: Alessandro Gerada
+    affiliations:
+      - ref: APT
+      - ref: LCL
+    orcid: 0000-0002-6743-4271
+    email: alessandro.gerada@liverpool.ac.uk
+    corresponding: true
+  - name: Charlotte Brookefield
+    affiliations:
+      - ref: APT
+      - ref: LCL
+  - name: Alex Howard
+    affiliations:
+      - ref: APT
+      - ref: LCL
+  - name: Nada Reza
+    affiliations:
+      - ref: APT
+  - name: Conor Rosato
+    affiliations:
+      - ref: APT
+  - name: Peter L. Green
+    affiliations:
+      - ref: APT
+      - ref: SOE
+  - name: William Hope
+    affiliations:
+      - ref: APT
+      - ref: LCL
+affiliations:
+  - id: APT
+    name: Antimicrobial Pharmacodynamics and Therapeutics, Department of Clinical Pharmacology and Therapeutics, Institute of Systems, Molecular & Integrative Biology, University of Liverpool
+  - id: LCL
+    name: Department of Infection and Immunity, Liverpool Clinical Laboratories, Clinical Support Services Building (CSSB), Liverpool University Hospitals NHS Foundation Trust
+  - id: SOE
+    name: School of Engineering, Foundation Building, Brownlow Hill, University of Liverpool, Liverpool, L69 7ZX
 ---
@@
 The impact of these interventions is often difficult to quantify.
-Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
+Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable [@baierMultimodalInfectionControl2018; @davisImpactPreventionBundle2016].
 Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.
@@
 At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets.
+In this context, *de novo* infection was assumed to capture routes of transmission outwith toilet use, and infection which was not directly related to a traceable transmission event.
 The probability of infection through these two routes---$P_{\text{de novo}, i}(t)$ and $P_{\text{toilet}, i}(t)$, respectively---were independent; therefore, the probability of patient $i$ acquiring infection at time $t$ was given by:
@@
 After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control.
-A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors.
+A summary of the posterior estimates of the parameters (reported as the median and interquartile range of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors.
 The densities of the posterior parameter distributions are shown in @fig-posterior-densities.
@@
   dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
-  summarise(across(everything(), ~ paste0(
-    round(mean(.x), 2),
-    " (",
-    round(sd(.x), 2),
-    ")"
-    )))
+  summarise(across(everything(), ~ {
+    med <- median(.x)
+    q25 <- quantile(.x, 0.25)
+    q75 <- quantile(.x, 0.75)
+    paste0(round(med, 2), " (", round(q25, 2), "–", round(q75, 2), ")")
+  }))
 
@@
 
-param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
+param_names <- c("toilet-frequenting-rate" = "Patient toilet frequenting rate",
                  "toilet-cleaning-rate" = "Toilet cleaning rate",
@@
   flextable() %>% 
+  add_header_row(values = c("Parameter name", "Units", "Value"), colwidths = c(1, 1, 3)) %>% 
+  set_header_labels(param = "", units = "", baseline = "Baseline", outbreak = "Outbreak", control = "Control") %>% 
+  align(i = 1, j = 3, align = "center", part = "header") %>% 
   labelizor(j = "units", part = "body",
@@
 Another limitation of \acrs{ABM} is the requirement for detailed data to calibrate and validate the model.
-Here, the data used was primarily from a retrospective descriptive dataset.
+Here, the data used were primarily from a retrospective descriptive dataset.
 As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature.
